---
title: "Pre-registration for Compensatory aspects of the effect of voicing on vowel duration in English"
author: "Stefano Coretta"
date: "20/12/2018"
output:
  pdf_document:
    citation_package: natbib
    highlight: tango
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
bibliography: linguistics.bib
biblio-style: unified.bst
mainfont: DejaVu Sans
fontsize: 11pt
header-includes:
- \usepackage{cleveref}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Study information

## Title

Compensatory aspects of the effect of voicing on vowel duration in English.

## Authorship

Stefano Coretta (The University of Manchester).

## Research questions

Vowel in English (and many more languages) are longer when followed by voiced stops than when followed by voiceless stops \citep{heffner1937}.
The origin of this 'voicing effect' is still object to debate.
According to the compensatory temporal adjustment account, the difference in vowel duration is the consequence of the difference in stop closure durations \citep{lindblom1967, slis1969a, slis1969, lehiste1970, lehiste1970a}.
Voiceless stops have longer closures than voiced stops \citep{lisker1957, van-summers1987, davis1989, de-jong1991}, so that vowels are longer when followed by the shorter closure durations of voiced stops.

An exploratory study by Coretta (submitted) indicates that the duration of the interval between two consecutive stops releases is not affected by consonant voicing in CV́CV words in Italian and Polish.
Coretta argues that the temporal stability of the release to release interval plus the difference in consonant closure durations drives differences in vowel durations by a mechanism of compensation between the duration of the vowel and that of the closure.

This study sets out to test whether English exhibits the same temporal pattern found in Italian and Polish.
The aim of the study is to answer the following questions:

* Q1: Is the duration of the interval between two consecutive stop releases (the release to release interval) in monosyllabic and disyllabic words affected by the voicing of C2 in English?
* Q2: Is the duration of the release to release interval affected by (a) the number of syllables of the word, (b) the quality of the V1, and (c) the place of C2?

## Hypotheses

In relation to Q1, the following hypotheses will be tested:

* H1a: The duration of the release to release interval is not affected by C2 voicing in disyllabic words.
* H1b: The release to release interval is longer in monosyllabic words with a voiced C2 than in monosyllabic words with a voiceless C2.

# Sampling plan

## Existing data

**Registration prior to creation of data**: As of the date of submission of this research plan for preregistration, the data have not yet been collected, created, or realised.

## Explanation of existing data

NA.

## Data collection procedures

### Participants

Inclusion rule: Native speakers of English from the Manchester area, 18+ yo, with no reported hearing or speaking disorders, and with normal or corrected to normal vision.

### Procedure

The participants will be recorded while reading sentences with CV́CV target words (see \Cref{s:other} for the list of target words and \Cref{s:manipulated} for the list of frame sentences) presented on a computer screen with PsychoPy [@peirce2009], in a sound attenuated room in the Phonetics Laboratory at the University of Manchester.
Each target word is combined with each frame sentence, and each word/frame combination will be read once.
Acoustic recording will be obtained with a Zoom H4n Pro recorder and a RØDE Lavalier microphone, at a sample rate of 44100 Hz (16-bit; downsampled to 22050 Hz for analysis).
The recordings will be subject to force alignment with SPPAS [@bigi2015] for analysis in Praat [@boersma2018].

## Sample size

24 words, 5 repetitions (= 120 tokens per participant).
For the number of participants, see the following section.

## Sample size rationale

The approach of the region of practical equivalence (ROPE) will be used for determining the participant sample size \citep{kruschke2015, vasishth2018a}.
This approach is based on establishing a region around 0 which could theoretically be interpreted as a 'no effect' region.
Based on estimates from the literature on the just noticeable difference \citep{huggins1972, nooteboom1980}, the chosen ROPE width is 20 milliseconds (from -10 to +10 milliseconds).

## Stopping rule

Data collection will stop when the width of the 95% credible interval of the effect of C2 voicing on the release to release duration is below 20 milliseconds.
Unfortunately, due to time constraints, data collection will be terminated earlier if the ROPE target hasn't been reached by...

# Variables

## Manipulated variables
\label{s:manipulated}

* **Vowel height**: high (/i, u/), mid-high (/e/), mid-low (/ɔ/), low (/a/).
* **Place of articulation of C1**: labial, coronal, velar.
* **Place of articulation of C2**: labial, coronal, velar.
* **Frame sentence**: *Scrivete X sul foglio*, *Ha detto X sei volte*, *Sentivo X di nuovo*, *Ripete X da sempre*.

See \Cref{s:other} for a list of word stimuli.

## Measured variables

* From the acoustic signal:
    * **Vowel duration**: from onset to offset of higher formant structure.
    * **Release to Vowel Offset**: from release of C1 to V1 offset.
    * **Speech rate**: calculated as syllables per second [`n of syllables / sentence duration`, @plug2018].

* From the EGG signal:
    * **Duration of voiced interval**: from voice onset to voice offset.

* From both the acoustic and EGG signal
    * **Voice Onset Time**: from the release of C1 to voice onset.

## Indices

NA.

# Design plan

## Study type

Experiment---A researcher randomly assigns treatments to study subjects, this includes field or lab experiments.
This is also known as an intervention experiment and includes randomized controlled trials.

## Blinding

No blinding is involved in this study.

## Study design

Repeated measures, mixed design.

## Randomisation

The sentence stimuli will be randomised within participant by means of the built-in randomisation procedure in PsychoPy [@peirce2009].

# Analysis plan

## Statistical models
\label{s:stats}

Separate models will be fitted in `R` [@r-core-team2018] to independently assess effects on vowel duration, voicing interval duration, VOT, and RVoffT.
*P*-values ($\alpha$ = 0.05) will be obtained with the `lmerTest` package [@kuznetsova2017], which uses the Satterthwaite's method of approximation to degrees of freedom.

```{r lmer, eval=FALSE}
library(lme4)
library(lmerTest)

vowel_lm <- lmer(
  vowel_duration ~
    height +
    c2_place +
    height:c2_place +
    speech_rate +
    (1+height|speaker) +
    (1|item)
)

vot_lm <- lmer(
  vot ~
    height +
    c2_place +
    height:c2_place +
    speech_rate +
    (1+height|speaker) +
    (1|item)
)

rvofft_lm <- lmer(
  rvofft ~
    height +
    c2_place +
    height:c2_place +
    speech_rate +
    (1+height|speaker) +
    (1|item)
)
```

Since I wish to directly test both the null and the alternative hypotheses concerning the duration of the voiced interval (H1), the BIC approximation of Bayes factors will be used instead of *p*-values [@raftery1995; @raftery1999; @wagenmakers2007; @jarosz2014].

```{r lmer-bf, eval=FALSE}
voicing_lm_1 <- lme4::lmer(
  voicing_duration ~
    height +
    c2_place +
    speech_rate +
    (1+height|speaker) +
    (1|item),
  REML = FALSE
)

voicing_lm_0 <- lme4::lmer(
  voicing_duration ~
    # height +
    c2_place +
    speech_rate +
    (1+height|speaker) +
    (1|item),
  REML = FALSE
)

bf_01 <- exp((BIC(voicing_lm_1) - BIC(voicing_lm_0)) / 2)
```

## Transformations

NA.

## Follow-up analyses

NA.

## Inference criteria

*P*-values will be used for the effect of vowel height on vowel duration, VOT, and RVoffT, with $\alpha$ = 0.05.
*P*-values below 0.05 will be deemed significant.
Bayes factors will be employed for the effect of vowel height on voicing interval duration, since I am interested in testing both the null and the alternative hypotheses.
The recommendations in \citet[139]{raftery1995} for the interpretation of the Bayes factors will be followed.

## Data exclusion

Tokens in which the consonant closure is entirely voiced will be excluded.

## Missing data

NA.

## Exploratory analysis

An exploratory analysis will assess the effect of vowel identity and the interaction between vowel height and vowel backness on vowel duration, voicing duration, RVoffT.
Also, the effects of vowel and place of articulation on voicing during closure will be examined.

# Script (Optional)

## Analysis scripts (Optional)

See \Cref{s:stats}.

# Other
\label{s:other}

List of target words.

\begin{tabular}{lll}
teep & teepus \\
teeb & teebus \\
terp & terpus \\
terb & terbus \\
tarp & tarpus \\
tarb & tarbus \\
teek & teekus \\
teeg & teegus \\
terk & terkus \\
terg & tergus \\
tark & tarkus \\
targ & targus \\
\end{tabular}

List of frame sentences: I'll say X this Thursday, You'll say X this Monday, She'll say X this Sunday, We'll say X this Friday, They'll say X this Tuesday.
