---
title: "Pre-registration for Compensatory aspects of the effect of voicing on vowel duration in English"
author: "Stefano Coretta"
date: "08/01/2019"
output:
  pdf_document:
    citation_package: natbib
    highlight: tango
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
bibliography: linguistics.bib
biblio-style: unified.bst
mainfont: DejaVu Sans
fontsize: 11pt
header-includes:
- \usepackage{cleveref}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Study information

## Title

Compensatory aspects of the effect of voicing on vowel duration in English.

## Authorship

Stefano Coretta (The University of Manchester).

## Research questions

Vowels in English (and other languages) are longer when followed by voiced stops than when followed by voiceless stops \citep{heffner1937}.
The origin of this 'voicing effect' is still object to debate.
According to the compensatory temporal adjustment account, the difference in vowel duration is the consequence of the difference in stop closure durations \citep{lindblom1967, slis1969a, slis1969, lehiste1970, lehiste1970a}.
Voiceless stops have longer closures than voiced stops \citep{lisker1957, van-summers1987, davis1989, de-jong1991}, so that vowels are longer when followed by the shorter closure durations of voiced stops.

An exploratory study discussed in \citet{coretta2018j} indicates that the duration of the interval between two consecutive stops releases is not affected by consonant voicing in CV́CV words in Italian and Polish.
Coretta argues that the temporal stability of the release to release interval plus the difference in consonant closure durations drives the differences in vowel durations by a mechanism of compensation between the duration of the vowel and that of the closure.

This study sets out to test whether English exhibits the same temporal pattern found in Italian and Polish.
The aim of the study is to answer the following questions:

* Q1: Is the duration of the interval between two consecutive stop releases (the release to release interval) in monosyllabic and disyllabic words affected by the voicing of C2 in English?
* Q2: Is the duration of the release to release interval affected by (a) the number of syllables of the word, (b) the quality of V1, and (c) the place of C2?

A third question pertains to differences between mono- and disyllabic words:

* Q3: What is the estimated difference in the effect of voicing on vowel and stop closure duration between monosyllabic and disyllabic words?

## Hypotheses

In relation to Q1, the following hypotheses will be tested:

* H1a: The duration of the release to release interval is not affected by C2 voicing in disyllabic words.
* H1b: The release to release interval is longer in monosyllabic words with a voiced C2 than in monosyllabic words with a voiceless C2.

H1b is proposed based on \citet{jacewicz2009}, who report that monosyllabic words are longer when C2 is voiced in American English.

The exploratory study of Italian and Polish indicates that the intrinsic duration of the vowel and the consonant contribute to the duration of the release to release interval.
More specifically, the release to release with a high vowel is shorter than that with a low vowel.
It is well known that cross-linguistically high vowels tend to be shorter than low vowels \citep{hertrich1997, esposito2002, mortensen2013, toivonen2015, kawahara2017}.
As for consonant place of articulation, if the consonant is velar the interval is shorter.
The closure of velar stops is shorter than that of labial stops (see for example \citealt{sharf1962}).
It is possible that some of the durational difference of the release to release interval seen in the exploratory study depend on intrinsic vowel and consonant duration.

Considering the relation between intrinsic vowel and consonant duration and release to release, we can formulate the following hypotheses in relation to Q2 above:

* H2a: The release to release interval is longer in monosyllabic than in disyllabic words.
* H2b: The duration of the release to release interval decreases according to the hierarchy /ɑː/, /əː/, /iː/.
* H2c: The release to release interval is shorter when C2 is velar.

As for Q3:

* H3: The effect of voicing on vowel duration is greater in monosyllabic than in disyllabic words.

There is no specific hypothesis concerning stop closures.

# Sampling plan

## Existing data

**Registration prior to creation of data**: As of the date of submission of this research plan for preregistration, the data have not yet been collected, created, or realised.

## Explanation of existing data

NA.

## Data collection procedures

### Participants

Inclusion rule: Native speakers of English from the Manchester area, 18+ yo, with no reported hearing or speaking disorders, and with normal or corrected to normal vision.

### Procedure
\label{s:procedure}

The participants will be recorded while reading sentences with CVC and CV́CVC target words presented on a computer screen with PsychoPy [@peirce2009], in a sound attenuated room in the Phonetics Laboratory at the University of Manchester.
The test words are C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}(VC) words, where C\textsubscript{1} = /t/, V\textsubscript{1} = /iː, əː, ɑː/, C\textsubscript{2} = /p, b, k, g/, and (VC) = /əs/.
This structure leads to 24 possible combinations.

\begin{tabular}{llll}
teep & teepus & teek & teekus \\
teeb & teebus & teeg & teegus \\
terp & terpus & terk & terkus \\
terb & terbus & terg & tergus \\
tarp & tarpus & tark & tarkus \\
tarb & tarbus & targ & targus \\
\end{tabular}

Each target word is combined with the following frame sentences: I'll say X this Thursday, You'll say X this Monday, She'll say X this Sunday, We'll say X this Friday, They'll say X this Tuesday.
Each word/frame combination will be read once.
Acoustic recording will be obtained with a Zoom H4n Pro recorder and a RØDE Lavalier microphone, at a sample rate of 44100 Hz (16-bit; downsampled to 22050 Hz for analysis).
The recordings will be subject to force alignment with SPPAS [@bigi2015] for analysis in Praat [@boersma2018].

## Sample size

24 words, 5 repetitions (= 120 tokens per participant).
We set a minimum of 20 participants with additional participants if needed (see following sections).

## Sample size rationale

The approach of the region of practical equivalence (ROPE) will be used for determining the participant sample size \citep{kruschke2015, vasishth2018a}.
This approach is based on establishing a region around 0 which could theoretically be interpreted as a 'no effect' region.
Based on estimates from the literature on the just noticeable difference \citep{huggins1972, nooteboom1980}, the chosen ROPE width is 20 milliseconds (from -10 to +10 milliseconds).

## Stopping rule

Data collection will stop either:

(a) When the width of the 95\% CI of the effect of C2 voicing will be less than 20 milliseconds, or
(b) If the ROPE target hasn't been reached by April 30th 2019, due to time constraints imposed on the project.

# Variables

## Manipulated variables
\label{s:manipulated}

* **Vowel**: /iː/, /əː/, /ɑː/.
* **Voicing of C2**: voiceless (/p, k/), voiced (/b, g/).
* **Place of articulation of C2**: labial (/p, b/), velar (/k, g/).
* **Number of syllables**: monosyllabic, disyllabic.
* **Frame sentence**: I'll say X this Thursday, You'll say X this Monday, She'll say X this Sunday, We'll say X this Friday, They'll say X this Tuesday.

See \Cref{s:procedure} for a list of word stimuli.

## Measured variables

* From the acoustic signal:
    * **Duration of the release to release interval**: from the release of C1 to the release of C2.
    * **V1 duration**: from appearance to disappearance of higher formant structure in the spectrogram in correspondence of V1 \citep{machac2009}.
    * **C2 closure duration**: from disappearance of higher formant structure in the V1C2 sequence to the release of C2 \citep{machac2009}.
    * **Speech rate**: calculated as number of syllables per second (number of syllables in the sentence divided by the sentence duration in seconds, \citealt{plug2018}).

C1/C2 release detection will be performed with an automatic procedure in Praat based on \citet{ananthapadmanabha2014}.
The output of the automatic force-alignment (see \Cref{s:procedure}) and the release detection algorithm will be checked and corrected manually if needed.

## Indices

NA.

# Design plan

## Study type

Experiment---A researcher randomly assigns treatments to study subjects, this includes field or lab experiments.
This is also known as an intervention experiment and includes randomised controlled trials.

## Blinding

No blinding is involved in this study.

## Study design

Repeated measures, mixed design.

## Randomisation

The sentence stimuli will be randomised within participant by means of the built-in randomisation procedure in PsychoPy \citep{peirce2009}.

# Analysis plan

## Statistical models
\label{s:stats}

Bayesian linear mixed models \citep{vasishth2018, mcelreath2015, kruschke2015} will be fitted with brms \citep{burkner2017, burkner2018} in R \citep{r-core-team2018}.

### Release to release

The following Bayesian regression model will be used to model the duration of the release to release interval.
As fixed effects: C2 voicing (factor, levels = 'voiceless', 'voiced'), number of syllables (factor, levels = 'disyllabic', 'monosyllabic'), centred speech rate, interaction between C2 voicing and number of syllables.
Factors are coded with treatment contrasts.
A by-speaker and by-word random intercept, and a by-speaker random coefficient for C2 voicing.

```{r}
rr_1 <- brm(
  rr ~
    c2_voice +
    n_syl +
    c2_voice:n_syl +
    speech_rate_c +
    (1 + c2_voice | speaker) +
    (1 | word),
  family = gaussian()
)
```

The following priors will be used.
For the intercept of the release to release interval duration, a normal distribution with mean 200 ms and SD = 50, based on the posterior distribution of the intercept in the Italian/Polish exploratory study.
For the effect of C2 voicing, a weakly informative prior as a normal distribution with mean 0 ms and SD = 25, based on results from the exploratory study.
For the effect of number of syllables, a weakly informative prior as a normal distribution with mean 50 ms and SD = 25, based on differences in vowel duration between mono- and disyllabic words, which range between 30 and 100 ms \citep{sharf1962, klatt1973}.
The same prior is chosen for the interaction of C2 voicing and number of syllables, based on a similar range of reported differences in vowel duration in monosyllabic words (30-100 ms).
For the effect of speech rate (centred), a normal distribution with mean -25 and SD = 10, based on results from the exploratory study.
For the random effects, a half Cauchy distribution (location = 0, scale = 25) for the standard deviation and a LKJ(2) distribution for the correlation.
For the residual standard deviation, a half Cauchy distribution with location 0 ms and scale 25.

```{r}
c(
  set_prior("normal(200, 50)", class = "Intercept"),
  set_prior("normal(0, 25)", class = "b", coef = "c2_voicevoiced"),
  set_prior("normal(50, 25)", class = "b", coef = "n_sylmono"),
  set_prior("normal(50, 25)", class = "b", coef = "c2_voicevoiced:n_sylmono"),
  set_prior("normal(-25, 10)", class = "b", coef = "speech_rate_c"),
  set_prior("cauchy(0, 25)", class = "sd"),
  set_prior("lkj(2)", class = "cor"),
  set_prior("cauchy(0, 25)", class = "sigma")
)
```

A separate model will test the effect of vowel and C2 place on the duration of the release to release duration.
As fixed effects: vowel (factor, levels = 'ee', 'er', 'ar'), C2 place (factor, levels = 'labial', 'velar'), centred speech rate, interaction between vowel and C2 place.
Factors are coded with treatment contrasts.
By-speaker and by-word random intercepts.

```{r}
rr_2 <- brm(
  rr ~
    vowel +
    c2_place +
    speech_rate_c +
    (1 | speaker) +
    (1 | word),
  family = gaussian()
)
```

The following priors will be used.
For the intercept of the release to release interval duration, a normal distribution with mean 200 ms and SD = 50, based on the posterior distribution of the intercept in the Italian/Polish exploratory study.
For the effect of vowel, a weakly informative prior as a normal distribution with mean 0 ms and SD = 30, based on differences in vowel duration reported in \citet{heffner1937}, \citet{house1953}, and \citet{hertrich1997}.
For the effect of C2 place of articulation, a weakly informative prior as a normal distribution with mean 0 ms and SD = 30, based on differences in closure durations between labial and velar stops, which range between 10 and 30 ms \citep{sharf1962}.
The remaining priors are the same as the ones of the previous model.

```{r}
c(
  set_prior("normal(200, 50)", class = "Intercept"),
  set_prior("normal(0, 30)", class = "b", coef = "voweler"),
  set_prior("normal(0, 30)", class = "b", coef = "vowelar"),
  set_prior("normal(0, 30)", class = "b", coef = "c2_placevelar"),
  set_prior("normal(-25, 10)", class = "b", coef = "speech_rate_c"),
  set_prior("cauchy(0, 25)", class = "sd"),
  set_prior("cauchy(0, 25)", class = "sigma")
)
```

### Vowel duration

The following Bayesian regression model will be used to model the duration of V1.
As fixed effects: C2 voicing (factor, levels = 'voiceless', 'voiced'), vowel (factor, levels = 'ee', 'er', 'ar'), number of syllables (factor, levels = 'disyllabic', 'monosyllabic'), centred speech rate, all logical interactions between C2 voicing, vowel, and number of syllables.
Factors are coded with treatment contrasts.
A by-speaker and by-word random intercept, and a by-speaker random coefficient for C2 voicing.

```{r}
vow_1 <- brm(
  v1_duration ~
    c2_voice +
    vowel +
    n_syl +
    c2_voice:vowel +
    c2_voice:n_syl +
    vowel:n_syl +
    c2_voice:vowel:n_syl +
    speech_rate_c +
    (1 + c2_voice | speaker) +
    (1 | word),
  family = gaussian()
)
```

The following priors will be used.

```{r}
c(
  set_prior("normal(145, 30)", class = "Intercept"),
  set_prior("normal(50, 20)", class = "b", coef = "c2_voicevoiced"),
  set_prior("normal(0, 30)", class = "b", coef = "voweler"),
  set_prior("normal(0, 30)", class = "b", coef = "vowelar"),
  set_prior("normal(50, 25)", class = "b", coef = "n_sylmono"),
  set_prior("normal(0, 20)", class = "b", coef = "c2_voicevoiced:voweler"),
  set_prior("normal(0, 20)", class = "b", coef = "c2_voicevoiced:vowelar"),
  set_prior("normal(50, 25)", class = "b", coef = "c2_voicevoiced:n_sylmono"),
  set_prior("normal(0, 30)", class = "b", coef = "voweler:n_sylmono"),
  set_prior("normal(0, 30)", class = "b", coef = "vowelar:n_sylmono"),
  set_prior("normal(0, 30)", class = "b", coef = "c2_voicevoiced:voweler:n_sylmono"),
  set_prior("normal(0, 30)", class = "b", coef = "c2_voicevoiced:vowelar:n_sylmono"),
  set_prior("normal(-25, 10)", class = "b", coef = "speech_rate_c"),
  set_prior("cauchy(0, 25)", class = "sd"),
  set_prior("lkj(2)", class = "cor"),
  set_prior("cauchy(0, 25)", class = "sigma")
)
```

### Closure duration
\label{s:closure}

The following Bayesian regression model will be used to model the duration of the closure of C2.
As fixed effects: C2 voicing (factor, levels = 'voiceless', 'voiced'), vowel (factor, levels = 'ee', 'er', 'ar'), number of syllables (factor, levels = 'disyllabic', 'monosyllabic'), centred speech rate, all logical interactions between C2 voicing, vowel, and number of syllables.
Factors are coded with treatment contrasts.
A by-speaker and by-word random intercept, and a by-speaker random coefficient for C2 voicing.

```{r}
c2_1 <- brm(
  c2_duration ~
    c2_voice +
    c2_place +
    n_syl +
    c2_voice:c2_place +
    c2_voice:n_syl +
    vowel:n_syl +
    c2_voice:c2_place:n_syl +
    speech_rate_c +
    (1 + c2_voice | speaker) +
    (1 | word),
  family = gaussian()
)
```

The following priors will be used.
The means reported in \citet{sharf1962} and \citet{luce1985} indicate that stop closures in monosyllabic words are 10-30 ms shorter when the stop is voiced.
A normal distribution with mean -20 ms and SD = 10 seems informative enough as a prior to the effect of C2 voicing on closure duration.
The same studies indicate velar stops have a closure which is 10-20ms shorter in monosyllabic words, hence the prior `normal(mean = -15, SD = 10)`.
Note that the estimates for C2 voicing and place refer to monosyllabic words, while the specified priors refer to disyllabic words.

```{r}
c(
  set_prior("normal(90, 20)", class = "Intercept"),
  set_prior("normal(-20, 10)", class = "b", coef = "c2_voicevoiced"),
  set_prior("normal(-15, 10)", class = "b", coef = "c2_placevelar"),
  set_prior("normal(0, 25)", class = "b", coef = "n_sylmono"),
  set_prior("normal(0, 10)", class = "b", coef = "c2_voicevoiced:c2_placevelar"),
  set_prior("normal(0, 10)", class = "b", coef = "c2_voicevoiced:n_sylmono"),
  set_prior("normal(0, 30)", class = "b", coef = "c2_placevelar:n_sylmono"),
  set_prior("normal(0, 30)", class = "b", coef = "c2_placevelar:n_sylmono"),
  set_prior("normal(0, 30)", class = "b", coef = "c2_voicevoiced:c2_placevelar:n_sylmono"),
  set_prior("normal(0, 30)", class = "b", coef = "c2_voicevoiced:c2_placevelar:n_sylmono"),
  set_prior("normal(-25, 10)", class = "b", coef = "speech_rate_c"),
  set_prior("cauchy(0, 25)", class = "sd"),
  set_prior("lkj(2)", class = "cor"),
  set_prior("cauchy(0, 25)", class = "sigma")
)
```

## Transformations

Speech rate will be centred to make the estimate of the intercept interpretable.
Centring will be obtained with the standard formula (speech rate - mean speech rate).

## Follow-up analyses

NA.

## Inference criteria

Bayesian posterior distributions, rather than point estimates, will be employed for inference.

## Data exclusion

Residual data with mistakes and speech errors will be discarded.

## Missing data

Individual missing observations (due to excluded data or annotation difficulties) will be dropped.

## Exploratory analysis

An exploratory analysis will look into differences in closure durations of voiceless and voiced stops in mono- and disyllabic words (see \Cref{s:closure}).

# Script (Optional)

## Analysis scripts (Optional)

See \Cref{s:stats}.

# Other

NA.
