---
title: |
  Vowel duration and tongue root advancement: Results from an exploratory study of the relation between voicing and vowel duration
author: Stefano Coretta
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
bibliography: linguistics
biblio-style: unified
fontsize: 12pt
header-includes:
- \usepackage{cleveref}
- \usepackage{ctable}
indent: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.lp = "f:")
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(tidyverse)
theme_set(theme_minimal())
library(coretta2018itapol)
data("kinematics_series")
data("kinematics")
library(mgcv)
library(tidymv)
library(itsadug)
library(lme4)
library(lmerTest)
library(sjPlot)
library(effects)

tra_series <- kinematics_series %>%
  filter(c1_phonation == "voiceless", proportion > 0) %>%
  mutate(
    c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced")),
    trp = -TR_displacement_sm,
    c2_phonation_o = ordered(c2_phonation,  levels = c("voiceless", "voiced")),
    speaker_voicing = interaction(speaker, c2_phonation)
  ) %>%
  mutate_if(is.character, as.factor) %>%
  arrange(rec_date, proportion) %>%
  create_start_event(rec_date)

contrasts(tra_series$c2_phonation_o) <- "contr.treatment"

kines <- kinematics %>%
  filter(c1_phonation == "voiceless", label == "closure_", TR_displacement_sm > 10) %>%
  mutate(
    c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced")),
    trp = -TR_displacement_sm,
    v1_duration_c = v1_duration - mean(v1_duration)
  ) %>%
  mutate_if(is.character, as.factor)

kines_max <- kinematics %>%
  filter(c1_phonation == "voiceless", label %in% c("max_TT", "max_TD"), TR_displacement_sm > 10) %>%
  mutate(
    c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced")),
    trp = -TR_displacement_sm
  ) %>%
  mutate_if(is.character, as.factor)

kines_mean <- kines %>%
  group_by(speaker) %>%
  mutate(trp_z = scale(trp)) %>%
  group_by(speaker, c2_phonation, language) %>%
  summarise(
    trp_mean = mean(trp, na.rm = TRUE),
    trp_sd = sd(trp, na.rm = TRUE),
    trp_z_mean = mean(trp_z, na.rm = TRUE),
    trp_z_sd = sd(trp, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

trp_difference <- kines_mean %>%
  group_by(speaker) %>%
  mutate(
    difference = trp_mean - lag(trp_mean),
    difference_se = sqrt( ((trp_sd^2) / n) + ((lag(trp_sd)^2) / lag(n)) )
    ) %>%
  na.omit()
```

# Introduction

It is well known that voiced stops (stops generally articulated with simultaneous vibration of the vocal folds) are almost universally accompanied by two phonetic correlates: advanced tongue root and preceding longer vowel durations \citep{westbury1983, lisker1974, fowler1992}.
While a lot of work has been done on each of these topic separately, less is known about their relation.
In this exploratory study of the articulatory correlates of stop voicing, it is found that tongue root advancement---a mechanism known to facilitate voicing during stop closure---is initiated during the production of the vowel preceding a stop.
This replicates previous work on tongue root position.
Moreover, the results of this study indicate that the acoustic duration of the vowel is positively correlated with tongue root position, such that longer vowel durations correspond to greater tongue root advancement.

## Tongue root position and voicing

One of the differences in supra-glottal articulation between voiced and voiceless stops concerns the position of the tongue root relative to the front-back dimension of the oral tract.
<!-- note that the actual movement is more forward and down, given it is pulled by the genioglossus -->
The initiation and maintenance of vocal fold vibration (i.e. voicing) requires a difference in air pressure between the cavities below and above the glottis.
Specifically, the sub-glottal pressure needs to be higher than the supra-glottal pressure, in other words, there must be a positive transglottal air pressure differential \citep{berg1958, rothenberg1967}.
This property of voicing is formally known as the Aerodynamic Voicing Constraint \citep{ohala2011}.
When the oral tract is completely occluded during the production of a stop closure, the supra-glottal pressure quickly increases, due to the incoming airstream from the lungs.
Such pressure increase can hinder the ability to sustain vocal fold vibration during closure, to the point voicing ceases.

An articulatory solution to counterbalance the increased pressure is to enlarge the supra-glottal cavity by advancing the root of the tongue.
It has been repeatedly observed that the tongue root is in a more front position in voiced stops compared to voiceless stops \citep{kent1969, perkell1969, westbury1983}.
\citet{rothenberg1967} calculates that the walls of the oral tract can absorb the incoming airflow for 20 to 30 ms by passive expansion, after which the sub- and supra-glottal pressures would equalise and voicing cease.
Based on these estimates, a passive expansion of the pharyngeal walls is thus not sufficient to maintain voicing during the closure of a stop.

Reaching a complete ballistic forward gesture would require the tongue root about 70 to 90 ms \citep{rothenberg1967}.
Given that voiced stop closures are on average shorter than that (the mean duration is about 64 ms in \citet{luce1985}), it is natural that the movement is initiated during the production of the vowel, so that an appreciable amount of advancement is obtained when closure is achieved.
Furthermore, \citet{westbury1983} finds that tongue root advancement is initiated before the achievement of full closure and that there is a forward movement even in some cases of voiceless stops, although the rate and magnitude of the advancement were consistently higher in voiced stops.
Finally, tongue root adjustments seem to target more specifically lingual consonants, while the tongue body is more involved in labials consonants \citet{perkell1969, westbury1983}.

However, the relation between tongue root advancement and voicing is a complex one.
First, tongue root advancement is not the only mechanism for sustaining voicing during a stop \citep{rothenberg1967, westbury1983, ohala2011} and it has a certain degree of idiosyncrasy \citep{ahn2016}.
A wide-spread difference between voiceless and voiced stops concerns their respective closure durations.
The closure of voiced stops are generally longer than those of voiceless stops \citep{lisker1957, umeda1977, van-summers1987, davis1989, de-jong1991}.
A shorter closure favours maintenance of vocal fold vibration by ensuring that the pressure build-up in the oral cavity does not equalise the sub-glottal and supra-glottal pressures (at which point voicing would cease).
<!-- lateral pharyngeal walls -->
Other solutions which help sustaining voicing during closure include larynx lowering \citep{riordan1980}, opening of the velopharyngeal port \citep{yanagihara1966}, and producing a retroflex occlusion \citep{sprouse2008}.

Second, implementation of tongue root advancement can be decoupled from the actual presence of vocal fold vibration.
In \citet{westbury1983}, advancement of the tongue root is found in some productions of voiceless stops.
This is counterintuitive, since tongue root advancement is generally considered to be a feature of voiced stops which require voicing-related pressure adjustments.
Moreover, \citet{ahn2015, ahn2016, ahn2018} looked at utterance-initial stops and found that the tongue root is more advanced in the phonologically voiced stops independent of whether they actually show vocal fold vibration or not.

To summarise, tongue root advancement is a common articulatory solution employed to counterbalance the increase in supra-glottal pressure and maintaining voicing during the production of at least lingual voiced stops.
While this gesture is not exclusive of voiced stops and it can be implemented even in the absence of vocal fold vibration, tongue root advancement seems to be a robust correlate of (phonological) voicing.

## Vowel duration and voicing

The results discussed here are part of a larger study which focusses on the effect of consonant voicing on preceding vowel durations.
A great number of studies showed that, cross-linguistically, vowels tend to be longer when followed by voiced obstruents than when they are followed by voiceless ones \citep{house1953, peterson1960, chen1970, klatt1973, lisker1974, farnetani1986, fowler1992, hussein1994, esposito2002, lampp2004, durvasula2012}.
This so-called 'voicing effect' has been reported in a variety of languages,
including (but not limited to) English, German, Hindi, Russian, Arabic, Korean, Italian, and Polish (see \citealt{maddieson1976} and \citealt{begus2017} for a more comprehensive list).
The latter two offer a good opportunity for the study of ..., given their reported differences in magnitude/presence of the voicing effect and their relative comparability.
While Italian has been consistently reported as a voicing-effect language \citep{caldognetto1979, farnetani1986, esposito2002}, some studies found an effect in Polish \citet{slowiaczek1985, nowak2006, malisz2008, coretta2018j} but others did not \citep{keating1984, jassem1989}.

\citet{coretta2018j} discusses the acoustic data of this study, and argues that the stressed vowels of disyllabic (CV́CV) words in Italian and Polish are 16 ms longer (SE = 4.4) when followed by a voiced stop.
The high degree of intra-speaker variation and backed up by statistical modelling also indicates that these languages possibly behave similarly regarding the voicing effect.
Moreover, \citet{coretta2018j} shows that the temporal distance between two consecutive stop releases in CV́CV words is not affected by the voicing of the second consonant.
The duration of the release to release interval is stable across voicing contexts.
Within this interval, the timing of the onset of the stop closure produces differences in the respective durations of vowel and closure, following a mechanism of temporal compensation \citep{lindblom1967, slis1969a, slis1969, lehiste1970, lehiste1970a}.
A later closure onset results in a long vowel and a short closure, while an earlier closure onset generates a short vowel and a long closure.
Since the closure of voiceless stops is longer than that of voiced stops, it follows that vowels are shorter when followed by the former than when followed by the latter.

## This study

In sum, tongue root advancement and longer vowel durations are two correlates of voicing.
Previous studies have shown that voicing can be maintained by advancing the tongue root during the production of voiced stops and that vowels followed by voiced stops tend to be longer than vowels followed by voiceless stops.
The acoustic data from this exploratory study discussed in \citet{coretta2018j} revealed that the duration of the stop closure bears on the duration of the preceding vowel.
The articulatory part of this study adds to this idea by offering new insights on mechanisms involving the root of the tongue and its relation with vowel duration, and on how these interact with stop closure duration.

# Methodology

## Participants

Participants were recruited in Manchester (UK), and Verbania (Italy)
Eleven native speakers of Italian (5 females, 6 males) and 6 native speakers of Polish (3 females, 3 males) participated in this study.
Most speakers of Italian are originally from the North of Italy, while 3 are from Central Italy.
The Polish speakers came from different parts of Poland (2 from the west, 3 from the centre, and 1 from the east).
This study has been approved by the SALC Ethics committee of the University of Manchester (REF 2016-0099-76).
The participants signed a written consent and received a monetary compensation of £10.

## Equipment

\label{s:equipment}

<!-- \begin{figure*}
    \centering
    \includegraphics[width=.7\textwidth]{../../graphics/uti-setup.pdf}
    \caption{Schematic representation of the equipment setup (\citealt{articulate2011}). The system is described in details in \Cref{s:equipment}.}
    \label{f:uti-setup}
\end{figure*} -->

Simultaneous recordings of audio and ultrasound tongue imaging were obtained in the Phonetics Laboratory at the University of Manchester (UK) and in a quiet room in Verbania (Italy).
An Articulate Instruments Ltd™ system was used for this study.
The system is made of a TELEMED Echo Blaster 128 unit, an Articulate Instruments Ltd™ P-Stretch synchronisation unit, and a FocusRight Scarlett Solo pre-amplifier (see \Cref{f:uti-setup}).
A TELEMED C3.5/20/128Z-3 ultrasonic transducer (20mm radius, 2-4 MHz) and a Movo LV4-O2 Lavalier microphone were used respectively for the acquisition of ultrasonic and audio data.
The ultrasonic probe was placed in contact with the sub-mental triangle, aligned with the mid-sagittal plane.
A metallic headset designed by Articulate Instruments Ltd™ (\citeyear{articulate2008}) was used to hold the probe in a fixed position and inclination relative to the head.
<!-- add figure -->
The acquisition of the mid-sagittal ultrasonic and audio signals was achieved with the software Articulate Assistant Advanced (AAA, v2.17.2) running on a Hawlett-Packard ProBook 6750b laptop with Microsoft Windows 7.
The synchronisation of the ultrasonic and audio signals was performed by AAA after recording by means of a synchronisation signal produced by the P-Stretch unit.
The ranges of the ultrasonic settings were: 43-68 frames per second, 88-114 number of scan lines, 980-988 pixel per scan line, field of view 71-93°, pixel offset 109-263, depth 75-180 mm.
The audio signal was sampled at 22050 Hz (16-bit).

## Materials

Disyllabic words of the form C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{2} were used as targets, where C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/, and V\textsubscript{2} = V\textsubscript{1} (e.g. *pata*, *pada*, *poto*, etc.), giving a total of 12 target words, used both for Italian and Polish.\footnote{Note that stressed vowels in open syllables in Italian are long \citep{renwick2016}. Moreover, /o/ is used here for typographical simplicity to indicate the mid-back vowels of Italian and Polish, although they do differ in quality. See \citet{kramer2009, renwick2016, gussmann2007}.}
The resulting words are nonce words, with a few exceptions, and they were presented in the languages' respective writing conventions (see \Cref{a:targets}).
A labial stop was chosen as the first consonant to reduce possible coarticulation with the following vowel.\footnote{However, note that \citet{westbury1983} and \citet{vazquez-alvarez2007} report tongue body lowering in the context of labial stops.}
Central/back vowels only were included in the target words for two reasons.
First, high and mid front vowels tend to be difficult to image with ultrasound, given their greater distance from the ultrasonic probe when compared with back vowels.
Second, high and mid front vowels usually produce less tongue displacement from and to a stop consonant.
This characteristic can make it more difficult to identify gestural landmarks using the methodology discussed in \Cref{s:process}.
Since the focus of the study was to explore differences in the closing gesture of voiceless and voiced stops, only lingual consonants have been included, since of course the closure of labial stops cannot be imaged with ultrasound.
The sentence *Dico X lentamente* 'I say X slowly' in Italian, and *Mówię X teraz* 'I say X now' for Polish functioned as frames for the test words.\footnote{Due to software constraints, the Polish frame sentence was presented on screen without diacritics. The speakers read them as if spelled correctly.}
Speakers were instructed to read the sentences without pauses and to speak at a comfortable pace.

## Procedure

The participants familiarised themselves with the sentence stimuli at the beginning of the session.
Headset and probe were then fitted on the participant's head.
<!-- A reference for the coordinate system was obtained by imaging the occlusal plane with a metallic bite plate inserted in the participant's mouth \citep{scobbie2011}.
The participant gently bites on the bite plate while pressing the tongue against it to keep it still.
The trace of the metallic plate visible in the ultrasonic image was then used to rotate the tongue data (see \Cref{s:process}). -->
<!-- add image -->
The participant read the sentence stimuli, which were presented on the computer screen in a random order, while the audio and ultrasonic signals were acquired simultaneously.
The random list of sentences was read 6 times consecutively (with the exception of IT02, who repeated the sentences 5 times only).
Due to software constraints, the order of the sentences within participant was kept the same for each of the six repetitions.
The participant could optionally take breaks between one repetition and the other.
Sentences with hesitations or speech errors were immediately discarded and re-recorded.
A total of 1212 tokens (792 from Italian, 420 from Polish) were obtained.

## Data processing and statistical analysis
\label{s:process}

The audio data was subject to force alignment using the SPeech Phonetisation Alignment and Syllabification software (SPPAS, \citealt{bigi2015}).
The outcome of the automatic alignment was then manually corrected, according to the recommendations in \citet{machac2009}.
The onset and offset of V1 in the C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{2} test words were respectively placed in correspondence of the appearance and disappearance of higher formant structure in the spectrogram.
Vowel duration was calculated as the duration of the V1 onset to V1 offset interval.
Speech rate is the number of syllables in the sentence (8 in Italian and 6 in Polish) divided by the duration of the sentence in seconds.
<!-- add image -->

The displacement of the tongue root was obtained from the ultrasonic data according to the procedure used in \citet{kirkham2017}.
Smoothing splines were automatically fitted to the visible tongue contours in AAA.
Manual correction was then applied in cases of clear tracking errors.
A fan-like frame consisting of 42 equidistant radial lines superimposed on the ultrasonic image was used as the coordinate system.
The origin of the 42 fan-lines coincides with the (virtual) origin of the ultrasonic beams, such that each fan-line is parallel to the direction of the ultrasonic scan lines.
Tongue root displacement was thus calculated as the displacement of the fitted spline along a selected vector \citep{strycharczuk2015}.
For each participant, the fan-line with the highest standard deviation of displacement within the area corresponding to the speaker's tongue root was chosen as the tongue root displacement vector.
A Savitzky--Golay smoothing filter (second-order, frame length 75 ms) was applied to the raw displacement.
Displacement values for analysis are taken from the smoothed displacement signal.
Tongue root displacement was obtained from a static time point (the onset of the closure of C2) and along the duration of the vowel.
The displacement values along the vowel duration were extracted at time points corresponding to real ultrasonic video frames.
Given the average frame rate is 55 frames per second, values are sampled about every 20 ms.
<!-- add image -->

Statistical analysis was performed in R v3.5.2 \citep{r-core-team2018}.
Linear mixed-effects models were fitted with lme4 v1.1-19 \citep{bates2015}.
Factor terms were coded with treatment contrasts (the reference level is the first listed for each factor): C2 voicing (voiceless, voiced), vowel (/a/, /o/, /u/).
Speech rate was centred for inclusion in the statistical models, by subtracting the mean speech rate across all speakers from the calculated speech rate values.
Centring ensures the intercepts are interpretable.
*t*-tests with Satterthwaite's approximation to degrees of freedom on the individual terms were used to obtain *p*-values using lmerTest v3.0-1 \citep{kuznetsova2017, luke2017}.
An effect is considered significant if the *p*-value is below the alpha level ($\alpha = 0.05$).
Generalised additive mixed models were fitted with mgcv v1.8-26 \citep{wood2011, wood2017}.
The smooths used thin plate regression splines as basis \citep{wood2003}.
The ordered factor difference smooths method described in \citet{soskuthy2017, wieling2018} was used to model the effect of factor terms in GAMs.
The models were fitted by maximum likelihood (ML) and autoregression in the residuals was controlled with a first-order autoregressive model.

Significance testing of the relevant predictors was achieved by comparing the ML score of the full model with the score of a null model (in which the relevant predictor is dropped), using the `compareML()` function of the itsadug package \citep{van-rij2017}.
A preliminary analysis indicated that including either language or C2 place of articulation as predictors produced respective *p*-values above the alpha level, without affecting the estimates of the other terms.
\Cref{s:idio} further discusses the idiosyncratic behaviour of the tongue root observed between speakers, which does not seem to pattern in any way with their native language.
For these reasons, these variables were not included in the models reported here and will not be discussed.
Future research is warranted to ascertain language-related differences and possible effects of place of articulation.

## Open Science statement

Following recent practices which encourage scientific transparency \citep{cruwell2018, berez-kroeker2018, roettger2019}, data and analysis code are available on the Open Science Framework.

# Results
\label{s:results}

## Tongue root position at C2 closure onset

```{r tra-lm, include=FALSE, cache=TRUE}
tra_lm <- lmer(
  trp ~
    c2_phonation +
    speech_rate_c +
    vowel +
    (1|speaker) +
    (1|item),
  data = kines
)

summary(tra_lm)
```

```{r trp-z-box, include=TRUE, fig.cap = "Fig caption", fig.lp="f:", out.extra="width=\\linewidth"}
kines %>%
  group_by(speaker) %>%
  mutate(trp_z = scale(trp)) %>%
  ggplot(aes(c2_phonation, trp_z)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.2, position = position_jitter(width = 0.2, seed = 878)) +
  labs(x = "C2 voicing", y = "Tongue root position (z-scores)")
```

\Cref{f:trp-z-box} shows raw data points and boxplots of the position of the tongue root at C2 closure onset when C2 is voiceless (left) and voiced (right).
Since the actual value of tongue root position (in mm) is not informative (it depends on the speaker's anatomy and on the probe location), scaled tongue root position is used here (note though that the raw data is used in statistical modelling).
The trend is that, not surprisingly, the position of the tongue root is more advanced if C2 is voiced compared to its position when C2 is voiceless.
A linear mixed-effects model with tongue root position as the outcome variable was fitted with the following predictors: fixed effects for C2 voicing (voiceless, voiced), centred speech rate (as number of syllables per second, centred), vowel (/a/, /o/, /u/); by-speaker and by-word random intercepts (a by-speaker random coefficient for C2 voicing led to singular fit, so was not included in the final model).
The effects of C2 voicing and vowel are significant according to *t*-tests with Satterthwaite's approximation to degrees of freedom.
The tongue root at C2 closure onset is 0.77 mm (SE = 0.35) more front when C2 is voiced, and it is 1.87 mm (SE = 0.42) more retracted if V1 is /o/.

## Tongue root position during V1
\label{s:trp-v1}

```{r tra-gam, include=FALSE, cache=TRUE}
tra_gam <- bam(
  trp ~
    c2_phonation_o +
    s(speech_rate_c) +
    s(proportion) +
    s(proportion, by = c2_phonation_o) +
    ti(proportion, speech_rate_c) +
    s(proportion, speaker, bs = "fs", m = 1),
  data = tra_series
)

rho <- start_value_rho(tra_gam)

tra_gam_ar <- bam(
  trp ~
    c2_phonation_o +
    s(speech_rate_c) +
    s(proportion) +
    s(proportion, by = c2_phonation_o) +
    ti(proportion, speech_rate_c) +
    s(proportion, speaker, bs = "fs", m = 1),
  data = tra_series,
  method = "ML",
  rho = rho,
  AR.start = tra_series$start_event
)

tra_gam_ar_0 <- bam(
  trp ~
    # c2_phonation_o +
    s(speech_rate_c) +
    s(proportion) +
    # s(proportion, by = c2_phonation_o) +
    ti(proportion, speech_rate_c) +
    s(proportion, speaker, bs = "fs", m = 1),
  data = tra_series,
  method = "ML",
  rho = rho,
  AR.start = tra_series$start_event
)

compareML(tra_gam_ar_0, tra_gam_ar)
```

The position of the tongue root during the articulation of V1 was assessed with generalised additive mixed models (GAMM).
A GAMM was fitted to tongue root position with the following terms: C2 voicing as a parametric term; a smooth term over centred speech rate, a smooth term over V1 proportion with a by-C2 voicing difference smooth, a tensor product interaction over V1 proportion and centred speech rate; a factor random smooth over V1 proportion by speaker (penalty order = 1).
A chi-squared test on the ML scores of the full model and model excluding C2 voicing indicates C2 voicing significantly improves fit ($\chi$(3) = 7.758, *p* = 0.001).
\Cref{f:tra-gam-plot} shows that the root advances during the production of the vowel, relative to its position at V1 onset.
Such forward movement can be seen both in the context of a following voiced stop and in the one of a voiceless stop.
However, the magnitude of the movement is greater in the former.
At V1 offset, there is a difference in tongue root position of about 1 mm.

```{r tra-gam-plot, include=TRUE, fig.cap="Fig caption", fig.lp="f:", out.extra="width=\\linewidth", cache=TRUE}
predict_gam(tra_gam_ar, exclude_terms = list("s(proportion,speaker)", "s(speech_rate_c)", "ti(proportion,speech_rate_c)")) %>%
  filter(speaker == "it01") %>% # random effect have been cancelled, so selecting one speaker
  ggplot(aes(proportion, fit)) +
  geom_smooth_ci(c2_phonation_o, size = 1) +
  labs(
    x = "V1 proportion",
    y = "Tongue root position (mm)"
  ) +
  scale_colour_discrete(name = "C2 voicing") +
  scale_linetype_discrete(name = "C2 voicing")
```

## Correlation between tongue root position and V1 duration
\label{s:trp-vdur}

```{r tra-lm-2, include=FALSE, cache=TRUE}
tra_lm_2 <- lmer(
  trp ~
    v1_duration_c +
    speech_rate_c +
    vowel +
    v1_duration_c:vowel +
    (1+v1_duration_c|speaker) +
    (1|item),
  data = kines
)

summary(tra_lm_2)
```

```{r tra-lm-2-plot, include=TRUE, fig.cap="Fig caption", fig.lp="f:", out.extra="width=\\linewidth"}
as_tibble(effect("v1_duration_c:vowel", tra_lm_2)) %>%
  ggplot(aes(v1_duration_c, fit)) +
  geom_point(data = kines, aes(y = trp), alpha = 0.1) +
  geom_line() +
  geom_ribbon(aes(ymax = upper, ymin = lower), alpha = 0.2) +
  geom_rug(data = kines, aes(y = trp), alpha = 0.1) +
  facet_grid(~ vowel) +
  labs(x = "Vowel duration (ms)", y = "Tongue root position (ms)")
```

A second linear mixed regression was fitted to tongue root position to assess the effect of V1 duration on root position.
The following terms were included: centred V1 duration (in milliseconds), centred speech rate (as number of syllables per second), vowel (/a/, /o/, /u/), C2 place of articulation (coronal, velar); an interaction between centred V1 duration and vowel; by-speaker and by-word random intercept and a by-speaker random coefficient for V1 duration.
All predictors and the V1 duration/vowel interaction are significant.
V1 duration and tongue root position are positively correlated: The longer the vowel, the more advanced the tongue root is at V1 offset ($\hat{\beta}$ = 0.056 mm, SE = 0.015).
The effect is stronger with /a/ than with /o/ and /u/ (see \Cref{f:tra-lm-2-plot}).

## Tongue root position during V2 as a function of V1 duration

```{r tra-gam-2, include=FALSE, cache=TRUE}
tra_gam_2 <- bam(
  trp ~
    s(v1_duration) +
    s(proportion) +
    ti(proportion, v1_duration) +
    s(proportion, speaker, bs = "fs", m = 1),
  data = tra_series
)

rho <- start_value_rho(tra_gam_2)

tra_gam_ar_2 <- bam(
  trp ~
    s(v1_duration) +
    s(proportion) +
    ti(proportion, v1_duration) +
    s(proportion, speaker, bs = "fs", m = 1),
  data = tra_series,
  method = "ML",
  rho = rho,
  AR.start = tra_series$start_event
)

tra_gam_ar_0_2 <- bam(
  trp ~
    s(v1_duration) +
    s(proportion) +
    # ti(proportion, v1_duration) +
    s(proportion, speaker, bs = "fs", m = 1),
  data = tra_series,
  method = "ML",
  rho = rho,
  AR.start = tra_series$start_event
)
```

The effect of V1 duration on tongue root position during V1 was modelled by fitting a GAMM with the following terms: tongue root position as the outcome variable, smooth terms over V1 duration and V1 proportion, a tensor product interaction over V1 proportion and V1 duration; a factor random smooth over V1 proportion by speaker (penalty order = 1).
The full model with the tensor product interaction over V1 proportion and V1 duration has better fit according to model comparison with a model without the interaction ($\chi$(3) = 12.559, *p* < 0.001).
\Cref{f:tra-gam-2-plot} shows the estimated root trajectories at four values of vowel duration.
The general trend is that the forward movement of the root during the vowel is greater the longer the duration of the vowel (\Cref{f:tra-gam-2-plot}).
Moreover, the trajectory curvature increases with vowel duration: Shorter vowels have a flatter trajectory of tongue root advancement.

```{r tra-gam-2-plot, include=TRUE, fig.cap="Fig caption", fig.lp="f:", out.extra="width=\\linewidth", cache=TRUE}
predict_gam(tra_gam_ar_2, exclude_terms = "s(proportion,speaker)", values = list(v1_duration = c(50, 100, 145, 200))) %>%
  filter(speaker == "it01") %>% # random effect have been cancelled, so selecting one speaker
  ggplot(aes(proportion, fit)) +
  geom_smooth_ci(v1_duration, size = 1, ci_alpha = 0.05) +
  labs(
    x = "V1 proportion",
    y = "Tongue root position (mm)"
  ) +
  scale_colour_discrete(name = "V1 duration (ms)") +
  scale_linetype_discrete(name = "V1 duration (ms)")
```

# Discussion
\label{s:discussion}

## Voicing, tongue root position and vowel duration

The results of this study of voicing and vowel duration in Italian and Polish revealed a few patterns in the relation between between consonant voicing category, tongue root position, and vowel duration.
Unsurprisingly, the position of the tongue root at vowel offset is more front when the following stop is voiced than when the following stop is voiceless in both surveyed languages.
This finding aligns with the results of previous work on English \citep{kent1969, perkell1969, westbury1983, ahn2018}.
When looking at the position of the tongue root during the vowel, it was found that the root starts advancing during the articulation of the vowel.
\citet{westbury1983} found the same pattern in English.
Moreover, similarly to the results in \citet{westbury1983}, some tongue root advancement during the production of the vowel is found even when C2 is voiceless.

A possible reason for the presence of such a small degree of advancement in voiceless stops is offered by arguments in relation to the general absence of advancement in labials.
\citet{westbury1983} proposes that the articulation of the closure of lingual stops mechanically involves movements of the tongue root, so that, in order to keep a constant oral pressure, the root moves forward while the tongue body moves upward.
On the other hand, the tongue can move freely in labial stops since their closure involves the lips.
This idea is supported by the 'trough effect' \citep{vazquez-alvarez2007}, i.e. VCV sequences involving a labial stop show tongue body lowering, and by the fact that voiced labials tend to resort to tongue body lowering rather than tongue root advancement as a mechanism for voicing maintenance \citep{perkell1969, westbury1983, ahn2018}.
The small degree of advancement in voiceless lingual stops could then as well be a mechanic consequence of the tongue moving upward for producing the stop closure.

The data discussed here also suggest that tongue root position is positively correlated with vowel duration, such that longer vowels show a more advanced tongue root at vowel offset (= closure onset) than shorter vowels.
Said correlation exists independent of the voicing status of the consonant following the vowel (compatible with the finding that even voiceless stops have some degree of advancement).
This correlation could be interpreted as to indicate that the onset of the forward gesture of the root is timed relative to a landmark preceding the closure independent of the duration of the vowel.
The timing of the stop closure along the advancement movement would sanction the degree of advancement found at closure onset.

<!-- there is still the possibility that the velocity profiles are different in voiceless vs. voiced, or they could derive from the curvature differences? -->

The dynamic data of tongue root advancement during the articulation of the vowel indicates that vowels followed by voiced stops have greater tongue root advancement at vowel offset than vowels followed by voiceless stops, in accordance with the results from the static analysis at vowel offset.
Moreover, a significant interaction between vowel duration and the trajectory shape was found.
Shorter vowels have a flatter trajectory, while the curvature of the trajectory in longer vowels has a greater curvature.

When comparing the effects of vowel duration and speech rate on tongue root position, though, we are faced with a paradox.
Both variables have a positive effect on tongue root position, so that longer vowels and higher speech rates imply a more advanced root.
However, speech rate has a negative effect on vowel duration (and segments duration in general), such that higher speech rates are correlated with shorter vowel durations (this holds for this data, see \citealt{coretta2018j}).
If higher speech rates mean shorter vowels and shorter vowels imply a less advanced root, we should also find less advancement with higher speech rates.
However, the results indicate the opposite, and higher speech rates are correlated with more root advancement.
However, a regression model on the position of the tongue root at *vowel onset* suggests that speech rate is positively correlated with tongue root position at vowel onset.
The tongue root is already in a more advanced position at vowel onset when the speech rate is high, so that, if vowel duration is held constant, more advancement is expected at vowel offset with higher speech rates even when higher speech rate has negative effect on vowel duration.

If we take the release of the consonant preceding the vowel as a reference point, a delayed consonant closure could ensure that, by the time closure is made, an appreciable amount of tongue root advancement is achieved, so that the value of the threshold at which transglottal pressure equalisation is reached and voicing ceases is increased (so that more time would be required to reach it).
This mechanism would ensure that voicing can be maintained during the duration of the stop closure.

<!-- rel-rel -->

As seen in this paper, the closure of voiced stops is initiated later (relative to the preceding consonant release) compared to the closure of voiceless stops.
Given the temporal stability of the release to release interval, the delay in producing a full closure seen in the context of voiced stops has thus a double advantage: (1) A greater degree of tongue root advancement is achieved at closure onset, and (2) the stop closure is shorter.
Both these features cooperate to satisfy the aerodynamic voicing constraint.
A more advanced tongue root ensures that the transglottal pressure differential is sufficient, and a shorter closure reduces the pressure build-up during the stop closure.

While the patterns observed in this study are in agreement with previous work, the correlation between tongue root position and vowel duration needs to be replicated in studies which expand the enquired contexts to other types of consonants and vowels.
Investigating the relative phasing of tongue root and body gestures in lingual and labial consonants is also necessary to shed light on the mechanisms that could underlie the timing of stop closures.
<!-- westbury and perkell don't find much TRA in labials, nor Ahn -->
The following sections further discuss in some details the idiosyncratic patterns observed across individual speakers and the estimated value of the difference in tongue root position.

## Individual differences
\label{s:idio}

```{r trp-voicing-plot, include=TRUE, fig.cap="Fig caption", fig.lp="f:", out.width=".49\\linewidth", fig.width=3, fig.height=3, fig.show="hold", fig.align="center"}
kines_mean %>%
  ggplot(aes(c2_phonation, trp_mean, linetype = language)) +
  geom_point() +
  geom_line(aes(group = speaker)) +
  scale_linetype_discrete(name = "Language") +
  labs(
    x = "C2 voicing",
    y = "Mean tongue root position (mm)"
  )

kines_mean %>%
  ggplot(aes(c2_phonation, trp_z_mean, linetype = language)) +
  geom_point() +
  geom_line(aes(group = speaker)) +
  scale_linetype_discrete(name = "Language") +
  labs(
    x = "C2 voicing",
    y = "Mean tongue root position (z-scores)"
  )
```

The results presented in \Cref{s:results} and discussed in \Cref{s:discussion} are general patterns that can be attributed to a population as inferred from the present sample.
However, the data shows a great amount of individual differences.
\Cref{f:trp-voicing-plot} shows two slope plots of mean tongue root position depending on C2 voicing for each speaker.
In each plot, the two means of each speaker are linked by a line that shows the difference (or lack thereof) in means.
Solid lines are Italian speakers, while dashed lines are Polish speakers.
The *y*-axis of left plot is raw mean positions in millimetres, while that of the right plot is the standardised values (z-scores).
An upward-slanted slope line indicates that the mean tongue root position in the voiced condition is higher, while a downward-slanted slope means a decrease in mean root position.
A flat slope suggests there is no difference in means between the voiceless and the voiced condition.
These plots show that all three possibilities of slope direction are found in the data.
The mean value of tongue root position with a voiced C2 relative to the voiceless mean is greater in some speakers, smaller in others, and very similar in yet other speakers.
Moreover, no discernible pattern can be found between the two languages: Speakers of both languages show more or less the same range of variation.
However, as we have seen in \Cref{s:results}, the estimated overall effect of C2 voicing is a more advanced tongue root.
The right plot of \Cref{f:trp-voicing-plot} confirms this point visually.
Two speaker show a declining slope (one is Italian and the other Polish), one has a virtually flat slope, while all the others have an increasing slope at varying degrees.
The individual variation across speakers found in this data is comparable to that in \citet{ahn2018}.

```{r trp-difference, include=TRUE, fig.cap="Fig caption", fig.lp="f:", out.extra="width=\\linewidth", cache=TRUE}
trp_difference %>%
  ggplot(aes(difference, reorder(speaker, difference))) +
  geom_errorbarh(
    aes(
      xmin = difference - difference_se,
      xmax = difference + difference_se,
      linetype = language
    ), height = 0) +
  geom_point(aes(shape = language), size = 4) +
  labs(
    x = "Difference in tongue root position (mm)",
    y = "Speakers"
  ) +
  scale_linetype_discrete(name = "Language") +
  scale_shape_discrete(name = "Language")
```

\Cref{f:trp-difference} shows the mean difference in tongue root position at vowel offset for each speaker.
The horizontal lines indicate the difference standard error.
The bottom 7 speakers (3 Polish, 4 Italian) show either a weak negative difference (the tongue root is slightly more advanced in voiceless stops) or a weak positive difference with wide standard errors which include 0.
The remaining 11 speakers have a more robust positive difference (the tongue root is more advanced in voiced stops).
Finally, speakers of each language do not cluster together, reiterating the observation above that language does not seem to be an informative parameter.

Interesting individual patterns can also be seen in the trajectories of tongue root position.
\Cref{f:tra-gam-s-ar-plot} shows these trajectories for the 17 speakers (note that the *y*-axis of each plot is on a different scale, so magnitude comparisons should not be made visually).
Speakers IT01, IT03, and PL04 in particular have a somewhat categorical distinction in tongue root position during vowels followed by voiceless vs. voiced stops.
Such tongue root distinction is implemented across the total duration of the vowel, rather than towards the end (as suggested by the results from the aggregated data, see \Cref{s:trp-v1}).
The phonological literature reports cases in which the tongue root distinction in vowels is enhanced, leading to phonological alternations or diachronic loss of the voicing distinction with maintenance of the tongue root distinction (see \citet{vaux1996} and references therein).
The ultrasound data from this study offers articulatory evidence for a possible precursor of said phonological patterns.\footnote{All the examples in \citet{vaux1996} are on vowels \textit{following} voiceless vs. voiced stops, rather than preceding, as in the current study. While beyond the scope of this paper, whether this is a systematic gap or not and how this relates to the present findings should be examined in future studies.}

```{r tra-gam-s-ar, cache=TRUE}
tra_gam_s <- bam(
  trp ~
    s(speech_rate_c) +
    s(proportion) +
    ti(proportion, speech_rate_c) +
    s(proportion, speaker_voicing, bs = "fs", m = 1),
  data = tra_series,
  method = "ML"
)

rho <- start_value_rho(tra_gam_s)

tra_gam_s_ar <- bam(
  trp ~
    s(speech_rate_c) +
    s(proportion) +
    ti(proportion, speech_rate_c) +
    s(proportion, speaker_voicing, bs = "fs", m = 1),
  data = tra_series,
  method = "ML",
  rho = rho,
  AR.start = tra_series$start_event
)
```

```{r tra-gam-s-ar-plot, include=TRUE, fig.cap="Fig caption", fig.lp="f:", out.extra="width=\\linewidth", fig.height=8, cache=TRUE, dependson="tra-gam-s-ar"}
predict_gam(tra_gam_s_ar, values = list(speech_rate_c = 0)) %>%
  separate(speaker_voicing, c("speaker", "voicing")) %>%
  mutate(voicing = factor(voicing, levels = c("voiceless", "voiced"))) %>%
  ggplot(aes(proportion, fit)) +
  geom_smooth_ci(voicing, size = 1) +
  facet_wrap(~ speaker, scales = "free", ncol = 3)
```

## Estimates of tongue root displacement

It is worth to briefly discuss the estimated difference in tongue root position between voiceless and voiced stops and its significance.
The estimated magnitude of such difference is 0.77 mm (SE = 0.35).
The 95% confidence interval for the difference is approximately within the range 0-1.5 mm.
\citet{rothenberg1967} argues that the anterior wall of the lower pharynx (corresponding to the tongue root) can move by 5 mm along the antero-posterior axis.
Figure 1 in \citet{kirkham2017} suggests that the tongue root of a Twi speaker is about 4 mm more front in /e/ (a +ATR vowel) than in /ɛ/ (a -ATR vowel).
Given that a difference of 4 mm can produce a substantially different acoustic output in vowels (like two different phonemes), a smaller difference in tongue root position as driven by consonantal factors makes sense.\footnote{Thanks to Sam Kirkham for bringing this point to my attention.}
Moreover, the data presented here indicate that for every millisecond increase in vowel duration there is a 0.056 mm increase in tongue root advancement (see \Cref{s:trp-vdur}).
If a maximal ballistic forward movement of the tongue root takes between 70 and 90 ms \citep{rothenberg1967}, we can estimate a maximum displacement of 3.9 to 5 mm.
These values are well in accordance with the maximum possible root displacement of 5 mm given by Rothenberg.

The results also shed some light on timing aspects of tongue root advancement.
Possibly, the correlation between tongue root position and vowel duration is a consequence of the timing of the advancement gesture.
In order to obtain such correlation, the onset of the gesture (during the articulation of the vowel) should be at a fixed distance from an earlier reference point (like the vowel onset or the preceding consonant offset) such that the timing of consonant closure will create the correlation seen in the data.
Although ideally the timing of the onset of the advancing gesture should be fixed, the velocity of the gesture itself could be different depending on the voicing of the following consonant.
It is possible that the velocity will be greater in the context of voiced stops, especially if the advancing gesture in this context is executed with greater muscular force.
Unfortunately, a preliminary screening of the current data on this matter was inconclusive due to the difficulty in identifying the onset of the advancing gesture.
Further data should be collected with the aim of testing the hypothesis that the timing of the gesture onset should be the same in voiceless and voiced contexts, while the velocity of the gesture should differ in the two contexts.

# Conclusion



\appendix

# Target words

\label{a:targets}

See \Cref{t:targets}.

\ctable[caption = The list of Italian and Polish target words. An asterisk indicates a real word.,
label = t:targets,
star,
doinside = \footnotesize
]{lllllll}{}{
\FL
Italian   &        &      & \hspace{0.5cm} & Polish &      &      \NN
\cmidrule{1-3}\cmidrule{5-7}
pata      & poto*  & putu & & pata   & poto & putu \NN
pada      & podo   & pudu & & pada*  & podo & pudu \NN
paca*     & poco*  & pucu & & paka*  & poko & puku \NN
paga*     & pogo   & pugu & & paga   & pogo & pugu \LL
}
