---
title: "Appendix with statistical output, participants’ sociolinguistic information, and list of target words"
shorttitle: "Appendix"
author:
- name: Stefano Coretta
  affiliation: The University of Manchester
  address: Oxford Road, Manchester M13 9PL  stefano.coretta\@manchester.ac.uk
output:
  glossaarticle::glossa_article:
    citation_package: natbib
    fig_caption: yes
bibliography: linguistics
classoption:
  - charis
header-includes:
- \let\tnote\relax
- \usepackage{cleveref}
- \usepackage{ctable}
- \usepackage{enumerate}
---

```{r setup-1, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.path = "./", fig.lp = "f:")
knitr::opts_knit$set(root.dir = here::here())
library(knitr)
```


```{r read, message=FALSE, results='hide'}
knit("./papers/2018-relrel/2018-relrel.Rmd", tangle = TRUE, output = "./papers/2018-relrel/2018-relrel.R")
source("./papers/2018-relrel/2018-relrel.R")
```

```{r vow-table}
kable(tidy(vow_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect) %>% mutate(term = c("Intercept", "Voicing = voiced", "Vowel = /o/", "Vowel = /u/", "Lang = Polish", "Place = velar", "Speech rate", "Voiced × /o/", "Voiced × /u/", "Voiced × Polish", "/o/ × Polish", "/u/ × Polish", "Voiced × /o/ × Polish", "Voiced × /u/ × Polish")) %>% rename(Predictor = term, Estimate = estimate, SE = std.error, `t-value` = statistic, `p-value` = p.value, `CI low` = conf.low, `CI up` = conf.high) %>% select(Predictor, Estimate, SE, `CI low`, `CI up`, df, `t-value`, `p-value`) %>% mutate(`< α` = ifelse(`p-value` < 0.05, "*", "")), format = "latex", booktabs = TRUE, linesep = "", digits = 2, caption = "Summary of the linear mixed-effects model fitted to vowel duration (Section 3.1 of the paper)") %>%
kable_styling(font_size = 8, latex_options = "hold_position")
```

```{r clo-table}
kable(tidy(clo_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect) %>% mutate(term = c("Intercept", "Voicing = voiced", "Vowel = /o/", "Vowel = /u/", "Lang = Polish", "Place = velar", "Speech rate", "Voiced × /o/", "Voiced × /u/", "Voiced × Polish", "/o/ × Polish", "/u/ × Polish", "Voiced × /o/ × Polish", "Voiced × /u/ × Polish")) %>% rename(Predictor = term, Estimate = estimate, SE = std.error, `t-value` = statistic, `p-value` = p.value, `CI low` = conf.low, `CI up` = conf.high) %>% select(Predictor, Estimate, SE, `CI low`, `CI up`, df, `t-value`, `p-value`) %>% mutate(`< α` = ifelse(`p-value` < 0.05, "*", "")), format = "latex", booktabs = TRUE, linesep = "", digits = 2, caption = "Summary of the linear mixed-effects model fitted to closure duration (Section 3.2 of the paper)") %>%
kable_styling(font_size = 8, latex_options = "hold_position")
```

```{r vow-clo-table}
kable(tidy(vow_clo_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect) %>% mutate(term = c("Intercept", "Closure dur.", "Vowel = /o/", "Vowel = /u/", "Speech rate", "Closure × /o/", "Closure × /u/", "C2 closure × sp. rate", "/o/ × sp. rate", "/u/ × sp. rate", "Closure × /o/ × sp. rate", "Closure × /u/ × sp. rate")) %>% rename(Predictor = term, Estimate = estimate, SE = std.error, `t-value` = statistic, `p-value` = p.value, `CI low` = conf.low, `CI up` = conf.high) %>% select(Predictor, Estimate, SE, `CI low`, `CI up`, df, `t-value`, `p-value`) %>% mutate(`< α` = ifelse(`p-value` < 0.05, "*", "")), format = "latex", booktabs = TRUE, linesep = "", digits = 2, caption = "Summary of the linear mixed-effects model for testing the correlation between vowel and closure duration (Section 3.3 of the paper)") %>%
kable_styling(font_size = 8, latex_options = "hold_position")
```

\ctable[caption = Participants' sociolinguistic information. The column 'Spent most time in' gives the city in which the participant spent most of their life. The last column ('> 6 mo') indicates whether the participant has spent more than 6 months abroad,
label = t:socio,
width=\textwidth,
star,
pos=h,
doinside = \scriptsize
]{llll>{\raggedright}p{3cm}l>{\raggedright}p{2cm}l}{}{
\FL
ID & Age & Sex    & Native L & Other Ls& City of birth & Spent most time in & > 6 mo \ML
IT01 & 29 & Male   & Italian & English, Spanish                          & Verbania            & Verbania        & Yes \NN
IT02 & 26 & Male   & Italian & Friulian, English, Ladin-Venetan          & Udine               & Tricesimo & Yes \NN
IT03 & 28 & Female & Italian & English, German                           & Verbania            & Verbania        & No  \NN
IT04 & 54 & Female & Italian & Calabrese                                 & Verbania            & Verbania        & No  \NN
IT05 & 28 & Female & Italian & English                                   & Verbania            & Verbania        & No  \NN
IT09 & 35 & Female & Italian & English                                   & Vignola & Vignola         & Yes \NN
IT11 & 24 & Male   & Italian & English                                   & Monza               & Monza           & Yes \NN
IT12 & 26 & Male   & Italian & English                                   & Rome               & Rome           & Yes \NN
IT13 & 20 & Female & Italian & English, French, Arabic, Farsi            & Ancona              & Chiaravalle     & Yes \NN
IT14 & 32 & Male & Italian & English, Spanish & Frosinone & Frosinone & Yes \NN
PL02 & 32 & Female & Polish  & English, Norwegian, French, German, Dutch & Koło                & Poznań          & Yes \NN
PL03 & 26 & Male   & Polish  & Russian, English, French, German          & Nowa Sol            & Poznań          & Yes \NN
PL04 & 34 & Female & Polish  & Spanish, English, French                  & Warsaw              & Warsaw          & No  \NN
PL05 & 42 & Male   & Polish  & English, French                           & Przasnysz           & Warsaw        & No  \NN
PL06 & 33 & Male   & Polish  & English                                   & Zgierz              & Zgierz          & Yes \NN
PL07 & 32 & Female & Polish  & English, Russian                          & Bielsk Podlaski     & Bielsk Podlaski & Yes \LL
}

\ctable[caption = Target words. Asterisks indicate real words,
label = t:targets,
star,
pos=h,
doinside = \footnotesize
]{lllllll}{}{
\FL
Italian   &        &      & \hspace{0.5cm} & Polish &      &      \NN
\cmidrule{1-3}\cmidrule{5-7}
pata      & poto*  & putu & & pata   & poto & putu \NN
pada      & podo   & pudu & & pada*  & podo & pudu \NN
paca*     & poco*  & pucu & & paka*  & poko & puku \NN
paga*     & pogo   & pugu & & paga   & pogo & pugu \LL
}
