---
title: "An exploratory study of voicing-related differences in vowel duration as compensatory temporal adjustment in Italian and Polish"
shorttitle: Voicing-related differences in vowel duration as compensatory temporal adjustment
author:
- name: Stefano Coretta
  affiliation: Linguistics and English Language, University of Manchester, Oxford Road, Manchester, M13 9PL, United Kingdom
  email: stefano.coretta\@manchester.ac.uk
output:
  jasaarticle::jasa_article:
    citation_package: natbib
    fig_caption: yes
abstract: |
  Over a century of phonetic research has established the cross-linguistic existence of the so called 'voicing effect', by which vowels tend to be shorter when followed by voiceless stops and longer when the following stop is voiced.
  However, no agreement is found among scholars regarding the source of this effect, and several causal accounts have been advanced.
  A notable one is the compensatory temporal adjustment account, according to which the duration of the vowel is inversely correlated with the stop closure duration (voiceless stops having longer closure durations than voiced stops).
  The compensatory account has been criticised due to lack of empirical support and its vagueness regarding the temporal interval within which compensation is implemented.
  The results from this exploratory study of Italian and Polish suggest that the duration of the interval between two consecutive stop releases in CVCV words in these languages is not affected by the voicing of the second stop.
  The durational difference of the first vowel then follows from differences in closure durations of the following stop.
  It is proposed that these findings agree with a gestural organisation in which the right edge of the consonant is in a constant relationship with the preceding vowel.
bibliography: linguistics.bib
header-includes:
- \usepackage{cleveref}
- \usepackage{ctable}
- \usepackage[caption=false]{subfig}
- \usepackage{enumerate}
mainfont: "Times New Roman"
classoption: preprint
params:
  version: 1.0
  version-date: "2018/08/06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.lp = "f:")
library(tidyverse)
theme_set(theme_minimal())
library(lme4)
library(lmerTest)
library(effects)
library(broom)
library(broom.mixed)
library(knitr)
```

```{r read-data, cache=TRUE, include=FALSE, message=FALSE}
speakers <- read_csv("./datasets/speakers.csv")
stimuli <- read_csv("./datasets/stimuli.csv")

durations <- list.files(
  path = "./datasets",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(
    c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced")),
    language = recode(language, italian = "Italian", polish = "Polish"),
    syl_rate = ifelse(
      language == "Italian",
      8 / sentence_duration,
      6 / sentence_duration
    )
  ) %>%
  filter(c1_phonation == "voiceless")

durations_filtered <- durations %>%
  group_by(speaker) %>%
  mutate(
    vowel_duration_z = scale(vowel_duration),
    closure_duration_z = scale(closure_duration)
  ) %>%
  filter(
    vowel_duration_z < 3, vowel_duration_z > -3,
    closure_duration_z < 3, closure_duration_z > -3
  )

word_filtered <- durations %>%
    group_by(speaker) %>%
  mutate(
    word_duration_z = scale(word_duration)
  ) %>%
  filter(
    word_duration_z < 3, word_duration_z > -3
  )

rel_filtered <- durations %>%
    group_by(speaker) %>%
  mutate(
    rel_rel_z = scale(rel_rel)
  ) %>%
  filter(
    rel_rel_z < 3, rel_rel_z > -3
  )
```

# Introduction
\label{s:intro}

<!-- it might be interesting to say that the VE applies even in cases when voicing is not actually present -->

Almost a hundred years of research have consistently shown that consonantal voicing has an effect on preceding vowel duration: vowels followed by voiced obstruents are longer than when followed by voiceless ones [@meyer1904; @heffner1937; @house1953; @belasco1953; @peterson1960; @halle1967; @chen1970; @klatt1973; @lisker1974; @raphael1975; @javkin1976; @maddieson1976; @farnetani1986; @kluender1988; @laeufer1992; @fowler1992; @hussein1994; @esposito2002; @lampp2004; @warren2005; @durvasula2012].
This so called 'voicing effect' has been found in a considerable variety of languages, including (but not limited to) English, German, Hindi, Russian, Italian, Arabic, and Korean (see \citealt{maddieson1976} for a more comprehensive, but still not exhaustive list).
Despite of the plethora of evidence in support of the *existence* of the voicing effect, agreement hasn't been reached regarding its *source*.

Several proposals have been put forward in relation to the possible source of the voicing effect (see \citealt{soskuthy2013} and \citealt{begus2017} for an overview).
The majority of the proposed accounts place the source of the voicing effect in properties of speech production.\footnote{Two accounts that posit a perceptual cause are \citet{javkin1976} and \citet{kluender1988}. To the best of my knowledge, Javkin's (1976) proposal remains to be empirically tested, while see \citet{fowler1992} for arguments against \citet{kluender1988}.}
A notable production account, which will be the focus of this study, is the compensatory temporal adjustment account [@lindblom1967; @slis1969a; @slis1969; @lehiste1970; @lehiste1970a].
According to this account, the voicing effect follows from the reorganisation of gestures within a unit of speech that is not affected by stop voicing.
The duration of such unit is held constant across voicing contexts, while the duration of voiceless and voiced obstruents differs.
The closure of voiceless stops is longer than that of voiced stops [@lisker1957; @van-summers1987; @davis1989; @de-jong1991].
As a consequence, vowels followed by voiceless stops (which have a long closure) are shorter than vowels followed by voiced stops (which have a short closure).
Advocates of the compensatory account propose two prosodic units as the scope of the temporal adjustment: the syllable (or, more neutrally, the VC sequence, \citealt{lindblom1967}), and the word [@slis1969a; @slis1969; @lehiste1970; @lehiste1970a].
However, the compensatory temporal adjustment account has been criticised in subsequent work.

Empirical evidence and logic challenge the proposal that the syllable or the word have a constant duration and hence drive compensation.
First, Lindblom's \citeyear{lindblom1967} argument that the duration of the syllable is constant is not supported by the findings in @chen1970 and @jacewicz2009.
@chen1970 rejects a syllable-based compensatory account on the light of the fact that the duration of the syllable is affected by consonant voicing.
@jacewicz2009 further show that the duration of monosyllabic words in American English changes depending on the voicing of the coda consonant.
Second, although the results in @slis1969 suggest that the duration of disyllabic words in Dutch is constant whether the second stop is voiceless or voiced, it does not follow from this fact that compensation should necessarily target the vowel preceding the stop.
Indeed, it is logically possible that the following unstressed vowel could be the target of the compensation, therefore differences in preceding vowel duration still call for an explanation.

The compensatory temporal adjustment account has been further challenged on the basis of the so called 'aspiration effect' [@maddieson1976], by which vowels are longer when followed by aspirated stops than when followed by unaspirated stops.
In Hindi, vowels before voiceless unaspirated stops are the shortest, followed by vowels before voiced unaspirated and voiceless aspirated stops, which have similar duration, followed by vowels before voiced aspirated stops, which are the longest.
@maddieson1976 find no compensatory pattern between vowel and consonant duration: the consonant /t/, which has the shortest duration, is preceded by the shortest vowel, and vowels before /d/ and /tʰ/ have the same duration although the durations of the two consonant are different.
@maddieson1976 argue that a compensatory explanation for differences in vowel duration cannot be maintained.

However, a re-evaluation of the way consonant duration is measured in @maddieson1976 might actually turn their findings in favour of a compensatory account.
Due to difficulties in detecting the release of the consonant of interest, consonant duration in @maddieson1976 is measured from the closure of the relevant consonant to the release of the following, (e.g., in *ab sāth kaho*, the duration of /tʰ/ in *sāth* is calculated as the interval between the closure of /tʰ/ and the release of /k/).
This measure includes the burst and aspiration (if present) of the consonant following the target vowel.
@slis1969a, however, state that the inverse correlation between vowel duration and the following consonant applies to *closure* duration, and not to the entire *consonant* duration.
If a correlation exists between vowel and closure duration, the inclusion of burst and/or aspiration clearly alters this relationship.

Indeed, the study on Hindi voicing and aspiration effects conducted by @durvasula2012 indicates that closure duration, measured from closure onset to closure offset, decreases according to the hierarchy voiceless unaspirated > voiced unaspirated > voiceless aspirated > voiced aspirated, which closely resembles the order of increasing vowel duration in @maddieson1976.
Nonetheless, \citet{durvasula2012} do not find a negative correlation between vowel duration and consonant closure duration, but rather a (small) *positive effect*.
Vowel duration increases with closure duration when voicing and aspiration are taken into account.
<!-- De Jong 1991 finds negative correlation -->
However, as noted in @begus2017, it is likely that this result is a consequence of not controlling for speech rate.
A small negative effect of closure duration can turn positive if the effect of speech rate (which is positive) is greater, given the cumulative nature of these effects \citep[p. 2177]{begus2017}.

Further evidence for a compensatory account comes from the effect of a third type of consonants, namely ejectives.
@begus2017 finds that in Georgian (which contrasts aspirated, voiced, and ejective consonants) vowels are short when followed by voiceless aspirated stops, longer before ejective stops, and longest when followed by voiced stops.
Crucially, stop closure duration follows the reversed pattern: closure duration is short in voiced stops, longer in ejectives, and longest in voiceless aspirated stops.
Moreover, vowel duration is inversely correlated with closure duration across the three phonation types.
@begus2017 argues that these findings support a temporal compensation account (although not univocally, see \citealt[Section V]{begus2017}).

To summarise, a compensatory temporal adjustment account has been proposed to explain the voicing effect.
According to such account, the difference in vowel duration before consonants varying in voicing (and possibly other phonation types) is the outcome of a compensation between vowel and closure duration.
After a careful review of the critiques advanced by @chen1970 and @maddieson1976, and in face of the results in @slis1969 and @begus2017,
a compensatory account gains credibility.
However, issues about the actual implementation of the compensation mechanism still remain.
In conclusion, while the compensatory temporal adjustment account is plausible on the light of the reviewed literature, we are left with the necessity of finding a constant speech interval within which compensation is logically implemented.
<!-- the good candidate -->

## The present study

This paper reports on selected results from a broader exploratory study that investigates the relationship between vowel duration and consonant voicing from both an acoustic and articulatory perspective.
Synchronised recordings of audio, ultrasound tongue imaging, and electroglottography were carried out to enable a data-driven approach to the analysis of features related to the voicing effect in the context of disyllabic (CV́CV) words in Italian and Polish.\footnote{As per \citet{cysouw2013}, the glossonyms \textit{Italian} and \textit{Polish} as used here to refer, respectively, to the languoids Italian [\textsc{Glottocode}: \texttt{ital1282}] and Polish [\textsc{Glottocode}: \texttt{poli1260}].}
Given its exploratory nature, this study was not devised to test the compensatory account, but rather to collect articulatory and acoustic data on the voicing effect.\footnote{To the best of my knowledge, this is the first attempt to gather synchronised acoustic, tongue imaging and electroglottographic data in relation to the voicing effect.}
Moreover, the design of the study has been constrained by the use of such articulatory techniques (see \Cref{s:method}).
Since the tongue imaging and electroglottographic data don't bear on the main argument put forward here, only the results from acoustics will be discussed.

Italian and Polish reportedly differ in the magnitude of the effect of stop voicing on vowel duration.
<!-- add more references -->
Italian has been unanimously reported as a voicing-effect language [@caldognetto1979; @farnetani1986; @esposito2002].
The mean difference in vowel duration when followed by voiceless vs. voiced consonants ranges between 22 and 24 ms, with longer vowels followed by voiced consonants.\footnote{The mean differences are based on 3 speakers in \citealt{farnetani1986} and 7 speakers in \citealt{esposito2002}. \citealt{caldognetto1979} does not report estimates of vowel duration, but the study is based on 10 speakers.}
On the other hand, the results regarding the presence and magnitude of the effect in Polish are mixed.
While @keating1984 reports no effect of voicing on vowel duration in data from 24 speakers, @nowak2006 finds that vowels followed by voiced stops are 4.5 ms longer in the 4 speakers recorded.
Moreover, @malisz2008 argue based on data from 40 speakers that the magnitude of the voicing effect in Polish is highly idiosyncratic, and claim their results are inconclusive on this matter.
<!-- Comparing Italian and Polish could reveal differences in the implementation of voicing, which could in turn indicate a possible source for the voicing effect. -->

The acoustic data from the exploratory study discussed here confirms the existence of a voicing effect in Italian and Polish, and suggests that the duration of the interval between two consecutive stop releases (the Release to Release interval) is not affected by the voicing of the second consonant in both languages.
This finding is compatible with a compensatory temporal adjustment account by which the timing of the stop closure onset within said interval determines the respective durations of vowel and closure.
I further propose that the constant duration of the Release to Release interval is congruent with current views on gestural timing [@goldstein2014] and I discuss the insights such an account provides in relation to our understanding of the gestural organisation of speech.

# Method
\label{s:method}

## Participants

Seventeen subjects in total participated in this exploratory study.
Eleven subjects are native speakers of Italian (5 female, 6 male), while six are native speakers of Polish (3 female, 3 male).
The Italian speakers are from the North and Centre of Italy (8 speakers from Northern Italy, 3 from Central Italy).
The Polish group has 2 speakers from Western Poland, 3 speakers from Central Poland, and 1 speaker from Eastern Poland.
For more information on the sociolinguistic details of the speakers, see \Cref{a:socioling}.
Ethical clearance for this study was obtained from the University of Manchester (REF 2016-0099-76).
The participants signed a written consent and received a monetary compensation of £10.

## Equipment

The acquisition of the audio signal was achieved with the software Articulate Assistant Advanced™ (AAA, v2.17.2, \citealt{articulate2011}) running on a Hawlett-Packard ProBook 6750b laptop with Microsoft Windows 7.
Audio recordings were sampled at 22050 Hz (16-bit), in a proprietary format (`.aa0`).
A FocusRight Scarlett Solo pre-amplifier and a Movo LV4-O2 Lavalier microphone were used for audio recording.
The microphone was placed at the level of the participant's mouth on one side, at a distance of about 10 cm.\footnote{The microphone was clipped onto a metal headset wore by the participant, which is part of the ultrasound equipment.}

## Materials
\label{s:materials}

<!-- maybe note on /u/ in italian is phonotactically not good -->

The target stimuli were disyllabic words with C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{2} structure, where C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/, and V\textsubscript{2} = V\textsubscript{1} (e.g. /pata/, /pada/, /poto/, etc.).\footnote{Italian has both a mid-low [ɔ] and a mid-high [o] back vowel in its vowel inventory. These vowels are traditionally described as two distinct phonemes \citep{kramer2009}, although both their phonemic status and their phonetic substance are subject to a high degree of geographical and idiosyncratic variability \citep{renwick2016}. As a rule of thumb, stressed open syllables in Italian (like the ones used in this study) have [ɔː] (vowels in penultimate stressed open syllables are long) rather than [oː] \citep{renwick2016}. On the other hand, Polish has only a mid-low back vowel phoneme /ɔ/ \citep{gussmann2007}. For sake of typographical simplicity, the symbol /o/ will be used here for both languages.}
Most are nonce words, although inevitably some combinations produce real words both in Italian (4 words) and Polish (2 words, see \Cref{a:targets}).
The lexical stress of the target words was placed by speakers of both Italian and Polish on V\textsubscript{1}, as intended.

The make-up of the target words was constrained by the design of the experiment, which included ultrasound tongue imaging (UTI).
Front vowels are difficult to image with UTI, since their articulation involves tongue positions which are particularly far from the ultrasonic probe, hence reducing the visibility of the tongue contour.
For this reason, only central and back vowels were included.
Since one of the variables of interest in the exploratory study was the closing gesture of C\textsubscript{2}, only lingual consonants were used.
A labial stop was chosen as the first consonant to reduce possible coarticulation with the following vowel (although see \citealt{vazquez-alvarez2007}).

The target words were embedded in a frame sentence, *Dico X lentamente* 'I say X slowly' in Italian [following @hajek2008], and *Mówię X teraz* 'I say X now' in Polish, and presented according to the respective writing conventions.
These sentences were chosen in order to keep the placement of stress and emphasis similar across languages.

## Procedure

The participant was asked to read the sentences with the target words which were sequentially presented on the computer screen.
The order of the sentences was randomised for each participant.
Participants read the list of randomised sentence stimuli 6 times.
Due to software constraints, the order of the list was kept the same across the six repetitions within each participant.
The reading task lasted between 15 and 20 minutes, with optional short breaks between one repetition and the other.
The total experiment time lasted around 45 minutes.
Each speaker read a total of 12 sentences for 6 times (with the exceptions of IT02, who repeated the 12 sentences 5 times, and IT07, with whom words containing /u/ were not recorded due to technical difficulties relating to ultrasound data collection), which yields to a grand total of 1224 tokens (792 from Italian, 432 from Polish).\footnote{IT01 and IT02 (the first two participants of this study) also read sentences with words starting with /b/, which were later excluded from the experimental design. The data from /b/-initial words are not included in the analysis reported in this paper.}

The experiment was carried out in two locations: in the sound attenuated booth of the Phonetics Laboratory at the University of Manchester, and in a quiet room in a field location in Italy (Verbania, Northern Italy).
In both locations the equipment and procedures were the same.
Data collection started in December 2016 and ended in March 2018.

## Data processing and measurements

\ctable[caption = List of measurements as extracted from acoustics.,
label = t:dur-measures,
width=\textwidth,
star
]{ll>{\raggedright}p{9cm}}{}{
\FL
\textbf{landmark}               &                  & \textbf{criteria}                                                                                    \ML
vowel onset           & (V1 onset)         & appearance of higher formants in the spectrogram following the release of /p/ (C1)            \NN
vowel offset          & (V1 offset)        & disappearance of the higher formants in the spectrogram preceding the target consonant (C2) \NN
consonant onset       & (C2 onset)         & corresponds to V1 offset                                                                    \NN
closure onset         & (C2 closure onset) & corresponds to V1 offset                                                                    \NN
consonant offset      & (C2 offset)        & appearance of higher formants of the vowel following C2 (V2); corresponds to V2 onset                                \NN
consonant release & (C1/C2 release)         & automatic detection + manual correction \citep{ananthapadmanabha2014}                                           \LL
}

The audio recordings were exported from AAA in the `.wav` format at the same sample and bit rate for further processing.
A forced aligned transcription was accomplished through the SPeech Phonetisation Alignment and Syllabification software (SPPAS) [@bigi2015].
The outcome of the automatic annotation was manually corrected, according to the criteria in \Cref{t:dur-measures}.
The releases of C1 and C2 were detected automatically by means of a Praat scripting implementation of the algorithm described in @ananthapadmanabha2014.
The durations in milliseconds of the following intervals were extracted with Praat scripting from the annotated acoustic landmarks: word duration, vowel duration (V1 onset to V1 offset), consonant closure duration (V1 offset to C2 release), and Release to Release duration (C1 release to C2 release).
Sentence duration was measured in seconds.
\Cref{f:segmentation} shows an example of the segmentation of /pata/ (a) and /pada/ (b) from an Italian speaker.
Syllable rate (syllables per second) was used as a proxy to speech rate [@plug2018], and was calculated as the number of syllables divided by the duration of the sentence in seconds (8 syllables in Italian, 6 in Polish).
All further data processing and visualisation was done in R v3.5.0 [@r-core-team2018; @wickham2017].

\begin{figure*}
  \figline{
    \fig{img/annotation-1.pdf}{0.5\textwidth}{(a)}
    \fig{img/annotation-2.pdf}{0.5\textwidth}{(b)}
  }
  \caption{Segmentation example of the words \textit{pata} and \textit{pada} uttered by an Italian speaker.}
  \label{f:segmentation}
\end{figure*}

## Statistical analysis

<!-- add power analysis? -->

Given the data-driven nature of the study, all statistical analyses reported here are to be considered exploratory (hypothesis-generating) rather than confirmatory (hypothesis-driven) [@kerr1998; @gelman2013; @roettger2018].
The durational measurements were analysed with linear mixed-effects models using `lme4` v1.1-17 in R [@bates2015], and model estimates were extracted with the `effects` package v4.0-2 [@fox2003].
All factors were coded with treatment contrasts and the following reference levels: voiceless (vs. voiced), /a/ (vs. /o/, /u/), coronal (vs. velar), Italian (vs. Polish).
The models were fitted by Restricted Maximum Likelihood estimation (REML).
The estimates in the results section refer to these reference levels unless interactions are discussed.
*P*-values for the individual terms were obtained with `lmerTest` v3.0-1, which uses the Satterthwaite's approximation to degrees of freedom [@kuznetsova2017; @luke2017].
A result is considered significant if the *p*-value is below the alpha level ($\alpha = 0.05$).\footnote{\citet{luke2017} argues that the common approach of using likelihood ratio tests for statistical inference with mixed models leads to inflated Type I error rates. \citet[1501]{luke2017} also warns that `results should be interpreted with caution, regardless of the method adopted for obtaining \textit{p}-values'.}

Bayes factors were used to specifically test the null hypotheses that word and Release to Release duration are not affected by C2 voicing (i.e., the effect of C2 voicing on duration is `0`).
For each set of null/alternative hypotheses, a full model (with the predictor of interest) and a null model (excluding it) were fitted separately using the Maximum Likelihood estimation (ML) [@bates2015, p. 34].
The BIC approximation was then used to obtain Bayes factors [@raftery1995; @raftery1999; @wagenmakers2007; @jarosz2014].
The approximation is calculated according to the equation in \ref{eq:bayes} [@wagenmakers2007, p. 796].

\begin{equation}
\label{eq:bayes}
BF_{01} \approx exp(\Delta{}BIC_{10}/2)
\end{equation}

where $\Delta{}BIC_{10} = BIC_1 - BIC_0$, $BIC_1$ is the BIC of the full model, and $BIC_0$ is the BIC of the null model.
Values of $BF_{01} > 1$ indicate a preference of H\textsubscript{0} over H\textsubscript{1}.
The interpretation of the Bayes factors follows the recommendations in @raftery1995 [p. 139].

The extracted measurements were filtered before statistical analysis.
Measures of vowel duration, closure duration, word duration, and Release to Release duration that are 3 standard deviations lower or higher than the respective means were excluded from the final dataset.
This operation (which generally corresponds to a loss of around 2.5\% of the data) yields a total of 920 tokens of vowel and closure durations, 1176 tokens of word duration, and 848 tokens of Release to Release duration.

## Reproducibility statement

Following the recommendations in @berez-kroeker2018, the data and code used to produce the analyses discussed in this paper are available on the Open Science Framework at \url{https://osf.io/bfyhr/?view_only=391ef2dcc2834039a90f739ddb6f137a} [@coretta2018g].

# Results

<!-- I talk about voiceless and voiced, should I mention that I mean underlyingly voiceless and voiced? I added it now in the results summary and I I say it also in the discussion towards the end -->

The following sections report the results of the study in relation to the durations of vowels, consonant closure, word, and the Release to Release interval.
When discussing the output of statistical modelling, only the relevant predictors and interactions will be presented.
To avoid the visual cluttering of parameters tables and alleviate the burden of the reader, the full output of statistical models (including confidence intervals and *p*-values) are given in \Cref{a:stats}.

## Vowel duration

```{r vow-lm}
vow_lm <- lmer(
  vowel_duration ~
    c2_phonation *
    vowel *
    language +
    c2_place +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered
)
```

```{r vowels-plot, include=TRUE, fig.cap = "Raw data and boxplots of the duration in milliseconds of vowels in Italian (top row) and Polish (bottom row), for the vowels /a, o, u/ when followed by a voiceless or voiced stop.", fig.lp="f:", out.extra="width=\\linewidth", cache=TRUE, dependson="read-data"}
durations_filtered %>%
  ggplot(aes(c2_phonation, vowel_duration)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(alpha = 0.1, width = 0.2) +
  facet_grid(language ~ vowel) +
  labs(x = "C2 voicing", y = "Vowel duration (ms)")
```

\Cref{f:vowels-plot} shows boxplots and raw data of vowel duration for the three vowels /a, o, u/ when followed by voiceless or voiced stops in Italian and Polish.
Vowel tend to be longer when followed by a voiced stop in both languages.
The effect appears to be greater in Italian than in Polish, especially for the vowels /a/ and /o/.
There is no evident effect of C2 voicing in /u/ in Italian (as in \citealt{ferrero1978}), but the effect is discernible in Polish /u/.
In Italian, vowels have a mean duration of 106 ms (sd = 27) before voiceless stops, and a mean duration of 118 ms (sd = 33) before voiced stops.
Polish vowels are on average 75 ms long (sd = 16) when followed by a voiceless stop, and 83 ms long (sd = 19) if a voiced stop follows.
The difference in vowel duration based on the raw means is 12 ms in Italian and 8 ms in Polish.

A linear mixed-effects model with vowel duration as the outcome variable was fitted with the following predictors: fixed effects for C2 voicing (voiceless, voiced), C2 place of articulation (coronal, velar), vowel (a, o, u), language (Italian, Polish), and speech rate (as syllables per second); by-speaker and by-word random intercepts with by-speaker random slopes for C2 voicing.
All possible interactions between C2 voicing, vowel, and language were included.
The following terms are significant according to *t*-tests with Satterthwaite's approximation to degrees of freedom: C2 voicing, vowel, language, and speech rate.
Only the interaction between C2 voicing and vowel is significant.
Vowels are 19 ms longer (se = 4.4) when followed by a voiced stop (C2 voicing).
The effect of C2 voicing is smaller with /u/ (around 5 ms, $\hat{\beta}$ = -14.4 ms, se = 6).
Polish has on average shorter vowels than Italian ($\hat{\beta}$ = -28 ms, se = 8), and the effect of voicing is estimated to be about 11 ms (although note that the interaction between language and C2 voicing is not significant).
Speech rate has a negative effect on vowel duration, such that faster rates correlate with shorter vowel durations ($\hat{\beta}$ = -15 ms, se = 1).

## Consonant closure duration

```{r clo-lm}
clo_lm <- lmer(
  closure_duration ~
    c2_phonation *
    vowel *
    language +
    c2_place +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered
)
```

```{r closure-plot, echo=FALSE, include=TRUE, fig.cap = "Raw data and boxplots of closure duration in milliseconds of voiceless and voiced stops in Italian (top row) and Polish (bottom row) when preceded by the vowels /a, o, u/.", fig.lp="f:", out.extra="width=\\linewidth", cache=TRUE, dependson="read-data"}
durations_filtered %>%
  ggplot(aes(c2_phonation, closure_duration)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(alpha = 0.1, width = 0.2) +
  facet_grid(language ~ vowel) +
  labs(x = "C2 voicing", y = "Vowel duration (ms)")
```

\Cref{f:closure-plot} illustrates stop closure durations with boxplots and individual raw data points.
A pattern opposite to that with vowel duration can be noticed: closure duration is shorter for voiced than for voiceless stops.
The closure of voiceless stops in Italian is 77 ms long (sd = 20), while the voiced stops have a mean closure duration of 63 ms (sd = 15).
In Polish, the closure duration is 69 ms (sd = 12) in voiceless stops and 58 ms (sd = 13) in voiced stops.
The difference in closure duration based on the raw means is 14 ms in Italian and 11 ms in Polish.
The same model specification as with vowel duration has been fitted with consonant closure duration as the outcome variable.
C2 voicing, C2 place, and speech rate are significant.
Stop closure is 16.5 ms shorter (se = 3) if the stop is voiced and 3.5 ms longer (se = 1.5) if velar.
Finally, faster speech rates correlate with shorter closure durations ($\hat{\beta}$ = -8.5 ms, se = 1 ms).

## Vowel and closure duration

```{r vow-clo-lm, include=FALSE}
vow_clo_lm <- lmer(
  vowel_duration ~
    closure_duration *
    vowel +
    syl_rate +
    (1+closure_duration|speaker) +
    (1|item),
  data = durations_filtered
)
summary(vow_clo_lm)
```

```{r vow-clo-plot, include=TRUE, fig.cap = "Raw data and estimated regression lines of the effect of closure duration on vowel duration for the vowels /a, o, u/ (from a mixed-effects model fitted to data pooled from Italian and Polish).", fig.lp="f:", out.extra="width=\\linewidth", cache=TRUE, dependson="read-data"}
as_tibble(effect("closure_duration:vowel", vow_clo_lm)) %>%
  ggplot(aes(closure_duration, fit)) +
  geom_point(data = durations_filtered, aes(y = vowel_duration), alpha = 0.1) +
  geom_line() +
  geom_ribbon(aes(ymax = upper, ymin = lower), alpha = 0.2) +
  geom_rug(data = durations_filtered, aes(y = vowel_duration), alpha = 0.1) +
  facet_grid(~ vowel) +
  labs(x = "Closure duration (ms)", y = "Vowel duration (ms)")
```

A model addressing the relationship between vowel and stop closure duration was fitted with the following terms and interactions: vowel duration as the outcome variable; as fixed effects, closure duration, vowel, speech rate; an interaction between closure duration and vowel; by-speaker and by-word random intercepts, and by-speaker random slopes for C2 voicing.
Closure duration has a significant effect on vowel duration ($\hat{\beta}$ = -0.15 ms, se = 0.06 ms).
The effect with /u/ is greater than with /a/ and /o/ ($\hat{\beta}$ = -0.35 ms, se = 0.06 ms).
In general, closure duration is inversely proportional to vowel duration.
However, such correlation is quite weak, as shown by the small estimates.
A 1 ms increase in closure duration corresponds to a 0.2--0.5 ms decrease in vowel duration.
\Cref{f:vow-clo-plot} shows for each vowel /a, o, u/ the individual data points and the regression lines with confidence intervals extracted from the mixed-effects model.

## Word duration

```{r word-bf}
word_lm <- lme4::lmer(
  word_duration ~
    c2_phonation +
    vowel +
    c2_place +
    language +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered,
  REML = FALSE
)

word_lm_null <- lme4::lmer(
  word_duration ~
    # c2_phonation +
    vowel +
    c2_place +
    language +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered,
  REML = FALSE
)

word_bf <- round(exp((BIC(word_lm) - BIC(word_lm_null)) / 2))
```
Words with a voiceless C2 are on average 397 ms long (sd = 81) in Italian and 356 ms long (sd = 39) in Polish.
Words with a voiced stop have a mean duration of 396 ms (sd = 72) in Italian and 362 ms (sd = 39) in Polish.
The following full and null models were fitted to test the effect of C2 voicing on word duration.
The full model is made up of the following fixed effects: C2 voicing, C2 place, vowel, language, and speech rate.
The model also includes by-speaker and by-word random intercepts, and a by-speaker random slope for C2 voicing.
The null model is the same as the full model with the exclusion of the fixed effect of C2 voicing.
The Bayes factor of the null against the full model is `r word_bf`.
Thus, the null model (in which there is no effect of C2 voicing, $\beta$ = 0) is `r word_bf` times more likely under the observed data than the full model.
This indicates that there is strong evidence for a null effect of C2 voicing on word duration.

## Release to Release interval duration

```{r rr-plot, include=TRUE, fig.cap="Raw data and boxplots of the duration in milliseconds of the Release to Release interval in Italian (left) and Polish (right) when C2 is voiceless or voiced.", out.extra="width=\\linewidth", cache=TRUE, dependson="read-data", warning=FALSE}
rel_filtered %>%
  ggplot(aes(c2_phonation, rel_rel)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(alpha = 0.1, width = 0.3) +
  facet_grid(~ language) +
  labs(x = "C2 voicing", y = "Release to Release duration (ms)") +
  theme(legend.position = "none")
```

```{r rr-bf}
rr_lm <- lme4::lmer(
  rel_rel ~
    c2_phonation +
    vowel +
    c2_place +
    language +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered,
  REML = FALSE
)

rr_lm_null <- lme4::lmer(
  rel_rel ~
    # c2_phonation +
    vowel +
    c2_place +
    language +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered,
  REML = FALSE
)

rr_bf <- round(exp((BIC(rr_lm) - BIC(rr_lm_null)) / 2))
```

In \Cref{f:rr-plot}, boxplots and raw data points show the duration of the Release to Release interval in words with a voiceless vs. a voiced C2 stop, in Italian and Polish.
It can be seen that the distributions, medians, and quartiles of the durations in the voiceless and voiced condition do not differ much in either language.
In Italian, the mean duration of the Release to Release interval is 210 ms (sd = 44) if C2 is voiceless, and 209 ms (sd = 41) if voiced.
In Polish, the mean durations are respectively 173 (sd = 22) and 172 (sd = 21) ms.
The specifications of the null and full models for the Release to Release duration are the same as for word duration.
The Bayes factor of the null model against the full model is `r rr_bf`, which means that the null model (without C2 voicing) is `r rr_bf` times more likely than the model with C2 voicing as a predictor.
The Bayes factor suggests there is positive evidence that duration of the Release to Release interval is not affected by C2 voicing.

## Summary

Seventeen participants were recorded while reading CV́CV words embedded in a frame sentence.
The stressed vowel was either /a, o, u/, and C2 was one of /t, d, k, g/.
Of the seventeen participants, 11 are native speakers of Italian and 6 of Polish.
The durations of the vowel, stop closure, word, and Release to Release interval were measured from the acoustic signal.
The analyses of the durational data suggest that:

\begin{enumerate}[(a)]
  \item Stressed vowels in C\textsubscript{1}V́C\textsubscript{2}V words in Italian and Polish are 19 ms longer (se = 4.4) when C2 is voiced.
  \item C2 closure is 16.5 ms shorter (se = 3) if the stop is voiced.
  \item Vowel duration negatively correlates with closure duration, such that shorter closures correspond to longer vowels.
  \item Both word duration and Release to Release duration are not affected by the underlying voicing specification of C2.
\end{enumerate}

# Discussion

The data and statistical analyses of this exploratory study suggest that the duration of the interval between the releases of two consecutive consonants in CV́CV words (the Release to Release interval) is insensitive to the phonological voicing of the second consonant (C2) in Italian and Polish.
In accordance with a compensatory temporal adjustment account [@slis1969; @lehiste1970], the difference in vowel duration before voiceless vs. voiced stops can be seen as the outcome of differences in stop closure duration.
More specifically, the timing of the closure onset of C2 within the invariant Release to Release interval determines the duration of the preceding vowel.
An earlier closure onset relative to the onset of the preceding vowel (like in the case of voiceless stops) causes the vowel to be shorter.
On the other hand, a later closure onset (like with voiced stops) produces a longer vowel.
\Cref{f:compensatory} illustrates this mechanism.

\begin{figure}
  \includegraphics{img/compensatory.pdf}
  \caption{A schematic representation of the voicing effect as a compensatory temporal adjustment phenomenon. The schematic shows the gestural unfolding of a CV́C sequence when C2 is voiceless (top panel), or voiced (bottom panel). Oral cavity aperture (on the \textit{y}-axis, as the inverse of oral constriction) through time (on the \textit{x}-axis) is represented by the black line. Lower values represent a more constricted oral tract (a contoid configuration), while higher values indicate a more open oral tract (a vocoid configuration). The black bars below the time axis represent voicing (vocal fold vibration). Various landmarks and intervals are indicated in the schematic. Design based on \citet{esposito2002}.}
  \label{f:compensatory}
\end{figure}

The invariance of the Release to Release interval allows us to refine the logistics of the compensatory account by narrowing the scope of the temporal adjustment action.
A limitation of this account, as proposed by @slis1969 and @lehiste1970, is the lack of a precise identification of the word-internal mechanics of compensation.
As already discussed in \Cref{s:intro}, it is not clear, for example, why the adjustment should target the preceding stressed vowel, rather then the following unstressed vowel or any other segment in the word.
Since the Release to Release interval includes just the vowel (broadly defined as a vocoid gesture) and the consonant closure, it follows that differences in closure duration must be reflected in differences in the duration of the preceding vowel.

On the one hand, the voicing effect can be re-interpreted as a by-product of gestural timing, rather then a consequence of intrinsic features of voicing *per se*, with a constant Release to Release interval as the explanans.
On the other hand, the Release to Release invariance is in turn an explanandum.
In the following section, I offer a gestural organisation account that allows the invariance of such interval to follow from the relative timing of the articulatory gestures in a CV́C sequence.

## Gestural alignment
\label{s:gestural}

<!-- people have shown to be links between articulation and acoustics in this sense (somebody cited somewhere) -->
<!-- I might have to explicitly say that release = right-edge sort of -->

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/gorganisation.pdf}
  \caption{Gestural organisation patterns for onsets (a), codas (b), heterosyllabic onsets (c). C = consonant, V = vowel, T = voiceless stop, D = voiced stop. See \Cref{s:gestural} for details. Based on \citet{marin2010}.}
  \label{f:gorganisation}
\end{figure}

According to the coupled oscillator model of syllabic structure [@browman1988; @browman2000; @goldstein2006; @goldstein2014], articulatory gestures can be timed according to two coupling modes: in-phase (synchronous) mode, by which two gestures start in synchrony, or anti-phase (sequential) mode, in which a gesture starts when the preceding one has reached its target.
@marin2010 showed that onset consonants in American English are in-phase with respect to the vowel nucleus and anti-phase with each other.
Such phasing pattern establishes a stable relationship between the centre of the consonant (or consonants in a cluster) and the following vowel.
Independent of the number of onset consonants, the temporal midpoint of the onset (the so-called 'C-centre') is maintained at a fixed distance from the vowel, such that an increasing number of consonants in the onset does not change the distance between the vowel and the onset C-centre (\Cref{f:gorganisation}a).
On the other hand, coda consonants are timed anti-phase with the preceding vowel and between themselves.
Temporal stability in codas is found in the lag between the vowel and the left-most edge of the coda, which is not affected by the number of coda consonants (\Cref{f:gorganisation}b).
Other studies found further evidence for the synchronous and sequential coupling modes (see extensive review in \citealt{marin2010} and \citealt{marin2014}), although the use of one mode over the other depends on the language and the consonants under study.

Consonants can thus be said to follow either a C-centre or a left-edge organisation pattern.
In both cases, of course, the pattern is relative to the tautosyllabic vowel (the following vowel for onsets, the preceding vowel for codas).
To the best of my knowledge, no study has reported the timing of onset consonants relative to the *preceding* (heterosyllabic) vowel.
The results from this acoustic study on Italian and Polish are compatible with a right-edge organisation pattern for onset consonants relative to the preceding stressed vowel (\Cref{f:gorganisation}c).
In CV́CV words, the timing of the C2 release (the acoustic parallel of the articulatory right edge of C2) is fixed relative to V1.
<!-- problem of what is the anchor in V1? also, what about phase and anti-phase? is right-edge a kind of anti-phase? -->

A consequence of a right-edge organisation pattern of C2 relative to V1 in CV́CV words is that differences in C2 closure duration do not affect the lag between V1 and the release of C2, as shown by the results of this study.
The invariance of the lag between the release of C1 and that of C2 then can be seen to follow from the invariance in timing between, on the one hand, C1 (which is always /p/ in this study) and V1, and, on the other, between V1 and the right edge of C2.

<!-- maybe footnote on circularity of argument of syllabic structure and gestural timing -->
<!-- also, all data so far is stressed vowels? -->
<!-- indeed we don't know much about how syllables are concatenated? -->

A right-edge organisation account is compatible with findings from a variety of sources.
@celata2018 show with ultrasound tongue imaging data that, in Italian, vowels followed by single consonants are longer than when followed by geminates (for example, /ba.ta/ vs. /bat.ta/).
However, vowels followed by a tautosyllabic cluster have the same duration as vowels followed by a heterosyllabic cluster (/pa.tron/ vs. /bat.man/).
@celata2018 argue that these results corroborate a rhythmic account in which the relevant unit is not the traditional syllable, but rather the the VC(C) sequence.
The importance of the VC(C) sequence as relevant speech unit has been previously recognised in the work of @farnetani1986 (who called it the 'rhythmic syllable') and @steriade2012 (who uses the term 'vowel to vowel interval', see also \citealt{hirsch2014} and \citealt{lunden2017}).
The duration of the rhythmic syllable or interval is constant across the phonological contexts, while the duration of the traditional syllable is not.
This reflects a gestural organisation in which the timing of the right edge of the consonant is fixed relative to the vowel.

@de-jong1991 reports that the closing gesture of voiceless stops (following stressed vowels) is faster than that of voiced stops, and that also it is timed earlier with respect to the opening gesture of the stressed vowel.
According to @de-jong1991, the differences in vowel duration are driven by the timing of the consonantal closing gesture relative to the vocalic opening gesture (also see \citealt{hertrich1997}).
Moreover, the data in @de-jong1991 show that the final portion of the opening gesture is prolonged before voiced stops.

This pattern fits the one reported in an electromyographic study by @raphael1975.
The electromyographic signal corresponding to the vocalic gesture reaches its plateau at the same time relative to the preceding consonant in the voiceless and voiced context, but the plateau is held for longer in the case of vowels followed by voiced stops.
This indicates that muscular activation in the vocalic gesture before voiced stops is held for longer.
@raphael1975 further notes that the durational difference in muscular activation corresponds to the difference in the acoustic duration of vowels before voiceless vs. voiced stops (see also \citealt{warren2005}).

The results of the studies just discussed, together with the results from this study, bring support to a view in which two aspects of gestural organisation contribute to the difference in vowel duration observed before consonants varying in their voicing specification.
These aspects are: (1) the *right-edge alignment* of coda consonants following a stressed vowel relative to the latter, and (2) the *differential timing of the closing gesture onset* for voiceless vs. voiced stops.
The interplay of these two aspects can be synthesised into a compensatory temporal adjustment account, which requires a temporally constant interval, produced by (1), and a temporal reorganisation, brought about by (2).

## Limitations and future work

The generalisations put forward in this paper strictly apply to disyllabic words with a stressed vowel in the first syllable.
It is possible that the organisation pattern found in this context does not occur in sequences including an unstressed vowel.
For example, it is known that the difference in closure duration between voiceless and voiced stops is not stable when the stops precede a stressed vowel, although vowels preceding pre-stress stops have slightly different durations [@davis1989].
According to the gestural interpretation given here, the absence of differences in closure duration should correspond to no difference in vowel duration.
Data from different contexts and different languages is thus needed to assess the generality of the claims put forward in this paper.

The constraints on experimental material enforced by the use of ultrasound tongue imaging have been previously mentioned in \Cref{s:materials}.
Given these constraints, temporal information from other vowels (like front vowels) and places of articulation is a desideratum.
\Cref{s:gestural} discusses the interpretation of the Release to Release invariance in CV́CV words as a consequence of the timing of C2 rather than of a holistic CV́C motor plan in which the Release to Release interval is held constant.
Although beyond the scope of this paper, disambiguating between these two interpretations on articulatory grounds is fundamental for a general understanding of a theory of gestural organisation (a promising venue of research might be the activation-spin model by \citealt{tilsen2013}).

The compensatory temporal adjustment account presented here extends to other durational effects discussed in the literature.
In particular, the account bears predictions on the direction of the durational difference led by phonation types different from voicing, like aspiration and ejection.
For example, the mix of results with regard to the effect of aspiration [@durvasula2012] suggests that the conditions for a temporal adjustment might differ across the contexts and languages studied.
In light of the results in @begus2017, future studies will have to investigate the durational invariance of speech intervals in relation to a variety of phonation contrasts.

# Conclusions

The results of an exploratory study on the effect of voicing on vowel duration bring support for a compensatory temporal adjustment account of such effect.
Acoustic data from seventeen speakers of Italian and Polish show that the temporal distance between two consecutive stop releases is not affected by the voicing of the second stop in CV́CV words.
The temporal invariance of the Release to Release interval, together with a difference in stop closure duration of voiceless and voiced stops, causes vowels to be shorter when followed by voiceless stops (which have a long closure) and longer when followed by voiced stops (the closure of which is short).
I proposed that the Release to Release invariance is a consequence of the gestural organisation of the CV́C sequence, in which the lag between the right-edge of the second consonant and the preceding stressed vowel is fixed.


\begin{acknowledgments}
I am grateful to Dr. Ricardo Bermúdez-Otero and Dr. Patrycja Strycharczuk for their immense support and patience in providing feedback on this project.
I also want to thank the audience at the 16th Laboratory Phonology conference (LabPhon16) for their invaluable input.
Thanks also go to my colleagues at the Phonetics Laboratory of the University of Manchester, who provided help in different ways.
Any remaining errors are my own.
This project has been funded by the School of Arts, Languages, and Cultures Graduate School at the University of Manchester.
\end{acknowledgments}

\appendix

# Output of statistical models
\label{a:stats}

See \Cref{tab:vow-table}, \Cref{tab:clo-table}, and \Cref{tab:vow-clo-table}.

```{r vow-table}
kable(tidy(vow_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect) %>% mutate(term = c("Intercept", "C2 voi: voiced", "Vow: /o/", "Vow: /u/", "Lang: Polish", "C2 place: velar", "Syl. rate", "Voiced, /o/", "Voiced, /u/", "Voiced, Polish", "/o/, Polish", "/u/, Polish", "Voiced, /o/, Polish", "Voiced, /u/, Polish")) %>% rename(Estimate = estimate, SE = std.error, `t-value` = statistic, `p-value` = p.value, `CI low` = conf.low, `CI up` = conf.high) %>% select(term, Estimate, SE, `CI low`, `CI up`, df, `t-value`, `p-value`) %>% mutate(`< α` = ifelse(`p-value` < 0.05, "*", "")), format = "latex", booktabs = TRUE, digits = 4, caption = "Summary of the linear mixed-effects model fitted to vowel duration.")
```

```{r clo-table}
kable(tidy(clo_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect) %>% mutate(term = c("Intercept", "C2 voi: voiced", "Vow: /o/", "Vow: /u/", "Lang: Polish", "C2 place: velar", "Syl. rate", "Voiced, /o/", "Voiced, /u/", "Voiced, Polish", "/o/, Polish", "/u/, Polish", "Voiced, /o/, Polish", "Voiced, /u/, Polish")) %>% rename(Estimate = estimate, SE = std.error, `t-value` = statistic, `p-value` = p.value, `CI low` = conf.low, `CI up` = conf.high) %>% select(term, Estimate, SE, `CI low`, `CI up`, df, `t-value`, `p-value`) %>% mutate(`< α` = ifelse(`p-value` < 0.05, "*", "")), format = "latex", booktabs = TRUE, digits = 4, caption = "Summary of a linear mixed-effects model fitted to closure duration.")
```

```{r vow-clo-table}
kable(tidy(vow_clo_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect) %>% mutate(term = c("Intercept", "C2 closure", "Vow: /o/", "Vow: /u/", "Syl. rate", "C2 closure, /o/", "C2 closure, /u/")) %>% rename(Estimate = estimate, SE = std.error, `t-value` = statistic, `p-value` = p.value, `CI low` = conf.low, `CI up` = conf.high) %>% select(term, Estimate, SE, `CI low`, `CI up`, df, `t-value`, `p-value`) %>% mutate(`< α` = ifelse(`p-value` < 0.05, "*", "")), format = "latex", booktabs = TRUE, digits = 4, caption = "Summary of a linear mixed-effects model fitted to vowel duration with closure duration as predictor.")
```

# Socio-linguistic information of participants
\label{a:socioling}

See \Cref{t:socio}.

<!-- say why there are missing IDs -->

\ctable[caption = Participants' sociolinguistic information.,
label = t:socio,
width=\textwidth,
star,
doinside = \footnotesize
]{llll>{\raggedright}p{5cm}lll}{}{
\FL
ID & Age & Sex    & Native L & Other Ls& City of birth & Spent most time in & > 6 mo \ML
it01 & 29 & Male   & Italian & English, Spanish                          & Verbania            & Verbania        & Yes \NN
it02 & 26 & Male   & Italian & Friulian, English, Ladin-Venetan          & Udine               & Tricesimo & Yes \NN
it03 & 28 & Female & Italian & English, German                           & Verbania            & Verbania        & No  \NN
it04 & 54 & Female & Italian & Calabrese                                 & Verbania            & Verbania        & No  \NN
it05 & 28 & Female & Italian & English                                   & Verbania            & Verbania        & No  \NN
it09 & 35 & Female & Italian & English                                   & Vignola & Vignola         & Yes \NN
it11 & 24 & Male   & Italian & English                                   & Monza               & Monza           & Yes \NN
it13 & 20 & Female & Italian & English, French, Arabic, Farsi            & Ancona              & Chiaravalle     & Yes \NN
it14 & 32 & Male & Italian & English, Spanish & Frosinone & Frosinone & Yes \NN
pl02 & 32 & Female & Polish  & English, Norwegian, French, German, Dutch & Koło                & Poznań          & Yes \NN
pl03 & 26 & Male   & Polish  & Russian, English, French, German          & Nowa Sol            & Poznań          & Yes \NN
pl04 & 34 & Female & Polish  & Spanish, English, French                  & Warsaw              & Warsaw          & No  \NN
pl05 & 42 & Male   & Polish  & English, French                           & Przasnysz           & Warsaw        & No  \NN
pl06 & 33 & Male   & Polish  & English                                   & Zgierz              & Zgierz          & Yes \NN
pl07 & 32 & Female & Polish  & English, Russian                          & Bielsk Podlaski     & Bielsk Podlaski & Yes \LL
}

# Target words
\label{a:targets}

See \Cref{t:targets}.

\ctable[caption = Target words. An asterisk indicates a real words.,
label = t:targets,
star,
doinside = \footnotesize
]{lllllll}{}{
\FL
Italian   &        &      & \hspace{0.5cm} & Polish &      &      \NN
\cmidrule{1-3}\cmidrule{5-7}
pata      & poto*  & putu & & pata   & poto & putu \NN
pada      & podo   & pudu & & pada*  & podo & pudu \NN
paca*     & poco*  & pucu & & paka*  & poko & puku \NN
paga*     & pogo   & pugu & & paga   & pogo & pugu \LL
}
