---
title: This is a title and this is too
shorttitle: A subtitle goes on another line
author:
- name: Stefano Coretta
  affiliation: The University of Manchester
  email: stefano.coretta\@manchester.ac.uk
  thanks: other info
output:
  jasaarticle::jasa_article:
    citation_package: natbib
    fig_caption: yes
abstract: |
  Put your abstract here.
bibliography: linguistics.bib
mainfont: "FreeSerif"
header-includes:
- \usepackage{cleveref}
- \usepackage{ctable}
- \usepackage[caption=false]{subfig}
classoption: preprint
params:
  version: 0.9
  version-date: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.lp="f:")
library(tidyverse)
theme_set(theme_minimal())
library(lme4)
library(lmerTest)
library(effects)
library(broom)
library(broom.mixed)
library(knitr)
```

```{r read-data, cache=TRUE, include=FALSE, message=FALSE}
speakers <- read_csv("./datasets/speakers.csv")
stimuli <- read_csv("./datasets/stimuli.csv")

durations <- list.files(
  path = "./datasets",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(
    c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced")),
    language = recode(language, italian = "Italian", polish = "Polish"),
    syl_rate = ifelse(
      language == "Italian",
      8 / sentence_duration,
      6 / sentence_duration
    )
  ) %>%
  filter(c1_phonation == "voiceless")

durations_filtered <- durations %>%
  group_by(speaker) %>%
  mutate(
    vowel_duration_z = scale(vowel_duration),
    closure_duration_z = scale(closure_duration)
  ) %>%
  filter(
    vowel_duration_z < 3, vowel_duration_z > -3,
    closure_duration_z < 3, closure_duration_z > -3
  )

word_filtered <- durations %>%
    group_by(speaker) %>%
  mutate(
    word_duration_z = scale(word_duration)
  ) %>%
  filter(
    word_duration_z < 3, word_duration_z > -3
  )

rel_filtered <- durations %>%
    group_by(speaker) %>%
  mutate(
    rel_rel_z = scale(rel_rel)
  ) %>%
  filter(
    rel_rel_z < 3, rel_rel_z > -3
  )
```

# Introduction

<!-- need to include Begus 2017 -->

Almost 100 years of research have repeatedly shown that consonantal voicing has an effect on preceding vowel duration: vowels followed by voiced obstruents are longer than when followed by voiceless ones [@heffner1937; @house1953; @belasco1953; @peterson1960; @halle1967; @chen1970; @klatt1973; @lisker1974; @raphael1975; @javkin1976; @maddieson1976; @farnetani1986; @kluender1988; @laeufer1992; @fowler1992; @hussein1994; @esposito2002; @lampp2004; @warren2005; @durvasula2012].
Evidence for such so called 'voicing effect' has been found in a variety of languages, including (but not limited to) English, German, Hindi, Russian, Italian, Arabic, and Korean [see @maddieson1976 for a more comprehensive, but still not exhaustive list].
Despite of the plethora of evidence in support of the *existence* of the voicing effect, still after 100 years agreement hasn't been reached regarding the source of this effect.
<!-- Place here catchy sentence that says what I found -->

<!-- most accounts place source in articulation \footnote{but see bla bla for perceptual and a discussion in soskuthy2013}, then say that a notavle one is durational trade-off but that is discarded by M&G and that it is not clear what the interval is. Then say that you will give data to support it and offer a way of reinterpreting the data in M&G -->

Several proposal have been put forward as to where to look for the possible cause of the voicing effect [see @soskuthy2013 and @begus2017 for an overview].
Most of the proposed accounts place the source of the voicing effect in properties of speech production.\footnote{Two accounts that point to perceptual features are \citet{javkin1976} and \citet{kluender1988}. To the best of my knowledge, \citet{javkin1976}'s proposal remains empirically untested, while see \citet{fowler1992} for arguments against \citet{kluender1988}.}
One of these production accounts, which will be the focus of this study, relates the voicing effect to some constant property of speech that is held constant across contexts while the local property of voiceless vs. voiced obstruents varies, thus creating a trade-off solution within the constant property.
@lindblom1967, @slis1969, and @lehiste1970 (among others) argue that the relevant invariant property of speech is a constant durational interval within which segments of different duration results in different duration of other segments.
Both the syllable/VC sequence [@lindblom1967] and the word [@slis1969a; @slis1969; @lehiste1970; @lehiste1970a] has been proposed as the fixed interval.
<!-- maybe mention later -->
The closure of voiced stops is shorter than that of voiceless stops.
It follows that vowels followed by shorter closures (like in the case of voiced stops) are longer than vowels followed by longer closures (like in the case of voiceless stops).
<!-- This account also fits with the finding that vowel duration is inversely correlated with consonant closure duration. -->
However, the compensatory temporal adjustment account has been criticised in several occasions.

The proposal of the syllable or the word as the targets for compensation and encounter difficulties when confronted with empirical evidence and when scrutinised by logic.
First, Lindblom's -@lindblom1967 argument that the syllable is the interval within which compensation happens is not supported by the findings in @chen1970 and @jacewicz2009.
@chen1970 rejects a syllable-based compensatory account on the light of the fact that the duration of the syllable is affected by consonant voicing.
More recently, @jacewicz2009 further shows that the duration of monosyllabic words in American English does not change dependent on the voicing of the coda consonant.
Second, although the results in @slis1969 confirm that the word does not change in duration whether the stop following the stressed vowel is voiceless or voiced, it does not follow from this fact that compensation should necessarily target the stressed vowel.
Indeed, it is possible that, for example, the following unstressed vowel could be the target of the compensation.

@maddieson1976 reject any compensatory account for the voicing effect based on data from Hindi on the so called 'aspiration effect', by which vowels are longer when followed by aspirated stops than when followed by non-aspirated stops.
Vowels before voiceless unaspirated stops are the shortest, followed by vowels before voiced unaspirated and voiceless aspirated stops, which have similar duration between each other, followed by vowels before voiced aspirated stops, which are the longest.
@maddieson1976 find no compensatory pattern between vowel and consonant duration: the consonant /t/, which has the shortest duration, is preceded by the shortest vowel, and vowels before /d/ and /tʰ/ have the same duration although the durations of the two consonant are different.

However, an reevaluation of the way consonant duration is measured in @maddieson1976 might actually turn the situation in favour of a compensatory account.
Consonant duration is in fact measured from the closure of the relevant consonant to the release of the following consonant, due to difficulties in detecting the release of the consonant of interest (e.g., in *ab sāth kaho*, the duration of /tʰ/ in *sāth* was calculated as the interval between the closure of /tʰ/ and the release of /k/).
This measure includes the burst and (eventual) aspiration of the consonant.
@slis1969a, however, states that the inverse correlation between vowel duration and the following consonant raises when consonant *closure* duration is taken into account, and not the entire *consonant* duration.
If the correlation exists between vowel and closure duration, the inclusion of burst/aspiration duration clearly alters this relationship.
Indeed, the data in @durvasula2012 show that closure duration, appropriately measured, decreases from voiceless unaspirated > voiced > voiceless aspirated > voiced aspirated, which closely resembles the order of increasing vowel duration in @maddieson1976.\footnote{\citet{durvasula2012} does not find a negative correlation between vowel duration and consonant closure duration, but rather a (small) positive effect: vowel duration increases with closure duration when including an voicing and aspiration as fixed effects. However, it is likely that this result is a consequence of not controlling for speech rate, so it will not be discussed here.}

More recently, @begus2017 investigated the effect of three phonation types on vowel durations in Georgian and finds that vowels are short when followed by voiceless aspirated stops, longer before ejective stops, and longest when followed by voiced stops.
Crucially, stop closure duration follows a symmetrical pattern: Closure duration is short in voiced stops, longer in ejectives, and longest in voiceless aspirated stops.
@begus2017 argues that these findings support a temporal compensation account, although not univocally.

To summarise, a compensatory temporal adjustment account of the voicing effect remains possible after a careful review of the critiques advanced by @chen1970 and @maddieson1976, and in face of the results in @begus2017, although issues about the actual implementation of the compensation still persist.
In conclusion, for the compensatory account to gain plausibility, an invariant interval within which compensation is implemented needs to be better defined, on the light of empirical data.
<!-- we need to find a good candidate -->

## The present study

<!-- be more explicit about bigger project! you just mention it here briefly to explain constraints -->

This paper reports on results from a broader exploratory study that investigates the relationship between vowel duration and consonant voicing from an articulatory perspective.
Synchronised recordings of audio, ultrasound tongue imaging, and electroglottography were carried out to enable a data-driven approach to the analysis of features related to the voicing effect in the context of disyllabic (CVCV) words in Italian and Polish.
The design of the study has been constrained by the use of these articulatory techniques (see \Cref{s:method}).
Moreover, given the exploratory nature of the study, the experimental design was not implemented to directly test the compensatory account.
Here, only the results from acoustic will be discussed.

Italian and Polish reportedly differ in the magnitude of the voicing effect.
<!-- add more references -->
Italian has been unanimously reported as a voicing effect language [@caldognetto1979; @farnetani1986; @esposito2002].
<!-- expand footnote, include means as reported in cited works -->
The mean difference in vowel duration when followed by voiceless vs. voiced consonants ranges between 22 and 24 ms [with longer vowels followed by voiced consonants, @farnetani1986; @esposito2002].\footnote{These estimates should be taken as a gross approximation.
There are several issues: number of speakers, different contexts, statistical modelling.}
On the other hand, the results regarding the presence and magnitude of the effect in Polished are mixed.
While @keating1984 reports no effect of voicing on vowel duration in data from 24 speakers, @nowak2006 finds that vowels followed by voiced stops are 4.5 ms longer in the 4 speakers recorded.
Moreover, @malisz2008 argue based on data from 40 speakers that the magnitude of the voicing effect in Polish is highly idiosyncratic, and claim their results to be inconclusive on this matter.
The difference in presence or magnitude of the voicing effect in Italian vs. Polish should enable us to find an underlying property that differs in the two languages and that might indicate a possible source for the voicing effect.

The acoustic data from the exploratory study reported here reveal that the duration of the interval between the releases of the two consonants in CVCV words (the Release to Release interval) is not affected by the voicing of the second consonant.
This finding is compatible with a compensatory temporal adjustment account by which the timing of the stop closure onset within said interval determines the respective durations of the vowel and the stop closure.
I further propose that the invariant duration of the Release to Release interval is congruent with current views on gestural timing [@goldstein2014] and I discuss the insights it provides in relation to our understanding of gestural organisation in speech.

# Method
\label{s:method}

## Participants

Seventeen subjects in total participated to this exploratory study.
Eleven participants were native speakers of Italian (5 female, 6 male), while six were native speakers of Polish (3 female, 3 male).
The Italian speakers were from the North and Centre of Italy (8 speakers from Northern Italy, 3 from Central Italy).
The Polish group had 2 speakers from Poznań and 4 speakers from Eastern Poland.
For more information on the speakers, see \Cref{a:socioling}.
Ethical clearance was obtained for this study from the University of Manchester (REF 2016-0099-76).
The participants signed a written consent and received a monetary compensation of £10.

## Equipment

The acquisition of the audio signal was achieved with the software Articulate Assistant Advanced™ (AAA, v2.17.2) running on a Hawlett-Packard ProBook 6750b laptop with Microsoft Windows 7, with a sample rate of 22050 MHz (16-bit) in a proprietary format.
A FocusRight Scarlett Solo pre-amplifier and a Movo LV4-O2 Lavalier microphone were used for audio recording.

## Materials
\label{s:materials}

<!-- be more explicit about bigger project! you just mention it here briefly to explain constraints -->
<!-- say that some words are real -->
<!-- maybe note on /u/ in italian is phonotactically not good -->

The target stimuli were disyllabic words with C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{2} structure, where C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/, and V\textsubscript{2} = V\textsubscript{1} (e.g. /pata/, /pada/, /poto/, etc.).
<!-- footnote about /o/ notation -->
Most are nonce words, although inevitably some combinations lead to real words both in Italian (4 words) and Polish (2 words, see \Cref{a:targets}).
The lexical stress of the target words was placed by speakers of both Italian and Polish on V\textsubscript{1}, as intended.
The make-up of the target words was constrained by the design of the experiment, which included ultrasound tongue imaging (UTI).
Front vowels are difficult to image with UTI, since their articulation involves tongue positions which are particularly far from the ultrasonic probe, hence reducing the visibility of the tongue contour.
For this reason, only central and back vowels were included.
Since one of the variables of interest in the exploratory study was the closing gesture of C\textsubscript{2}, only lingual consonants were used.
A labial stop was chosen as the first consonant to reduce possible coarticulation with the following vowel (although see \citealt{vazquez-alvarez2007}).
The target words were embedded in a frame sentence, *Dico X lentamente* 'I say X slowly' in Italian [following @hajek2008], and *Mówię X teraz* 'I say X now' in Polish, and presented according to the respective writing conventions.
These sentences were chosen in order to keep the placement of stress and emphasis similar across languages, so to ensure comparability of results.

## Procedure

The participant was asked to read the sentences with the target words which were sequentially presented on the computer screen.
The order of the sentence stimuli was randomised for each participant.
Each participant read the list of randomised sentence stimuli 6 times.
Due to software constraints, the order of the list was kept the same across the six repetitions within each participant.
Each speaker read a total of 12 sentences for 6 times (with the exceptions of IT02, who repeated the 12 sentences 5 times, and IT07, with whom words containing /u/ were not recorded due to technical difficulties relating to the ultrasound data collection).\footnote{IT01 and IT02 (the first two participants of this study) read also sentences with words starting with /b/, which were later excluded from the experimental design. The data from /b/-initial words are not included in the analysis reported in this paper.}
with a grand total of 1224 tokens (792 from Italian, 432 from Polish).
The reading task lasted between 15 and 20 minutes, with optional short breaks between one repetition and the other.

## Data processing and measurements

\ctable[caption = List of measurements as extracted from acoustics.,
label = t:dur-measures,
width=\textwidth,
star
]{ll>{\raggedright}p{9cm}}{}{
\FL
\textbf{landmark}               &                  & \textbf{criteria}                                                                                    \ML
vowel onset           & (V1 onset)         & appearance of higher formants in the spectrogram following the burst of /p/ (C1)            \NN
vowel offset          & (V1 offset)        & disappearance of the higher formants in the spectrogram preceding the target consonant (C2) \NN
consonant onset       & (C2 onset)         & corresponds to V1 offset                                                                    \NN
closure onset         & (C2 closure onset) & corresponds to V1 offset                                                                    \NN
consonant offset      & (C2 offset)        & appearance of higher formants of the vowel following C2 (V2); corresponds to V2 onset                                \NN
consonant release & (C1/C2 release)         & automatic detection + manual correction \citep{ananthapadmanabha2014}                                           \LL
}

The audio recordings were exported from AAA in `.wav` format for further processing.
A forced aligned transcription was accomplished through the SPeech Phonetisation Alignment and Syllabification software (SPPAS) [@bigi2015].
The outcome of the automatic annotation was manually corrected when necessary, according to the criteria in \Cref{t:dur-measures}.
The releases of C1 and C2 were detected automatically by means of a Praat scripting implementation of the algorithm described in @ananthapadmanabha2014.
The durations in milliseconds of the following intervals were extracted from the annotated acoustic landmarks with Praat scripting: sentence duration, word duration, vowel duration (V1 onset to V1 offset), consonant closure duration (V1 offset to C2 burst), and Release-to-Release duration (RR duration, C1 release to C2 release).
\Cref{f:segmentation} shows an example of the segmentation of /pata/ (a) and /pada/ (b) from an Italian speaker.
Syllable rate (syllables per second) was used as a proxy to speech rate [@plug2018] for duration normalisation, and was calculated as the number of syllables divided by the duration of the sentence (8 syllables in Italian, 6 in Polish).

\begin{figure*}
  \figline{
    \fig{img/annotation-1.pdf}{0.5\textwidth}{(a)}
    \fig{img/annotation-2.pdf}{0.5\textwidth}{(b)}
  }
  \caption{Segmentation example.}
  \label{f:segmentation}
\end{figure*}

## Statistical analysis

<!-- add power analysis? -->
<!-- say what reference levels were used, so that when reading results people know it's reference level -->
<!-- reference to other packages I use :D -->

Given the exploratory nature of the study, all statistical analyses reported here are to be considered data-driven or hypothesis-generating rather than hypothesis-driven [@kerr1998; @gelman2013].
The durational measurements were analysed with linear mixed-effects models using `lme4` v1.1-17 in R v3.5.0 [@r-core-team2018; @bates2015].
All factors were coded with treatment contrasts.
*P*-values for the individual terms were obtained with `lmerTest` v3.0-1, which uses the Satterthwaite's approximation to degrees of freedom [@kuznetsova2017; @luke2017].
*P*-values below the alpha level 0.05 were considered significant.
<!-- explain why lmertest becuase likelihod ratio has high type I errors luke2017 and other reference 2000 in kuznetsova  -->

Bayes factors were used to specifically test the null hypotheses that word and RR duration are not affected by C2 voicing (i.e., the effect of C2 voicing on duration is `0`).
For each set of null/alternative hypotheses, a full model (with the predictor of interest) and a null model (excluding it) were fitted separately using Maximum Likelihood estimation [@bates2015, p. 34].
The BIC approximation was then used to obtain Bayes factors [@raftery1995; @raftery1999; @wagenmakers2007; @jarosz2014].
The approximation is calculated according to the equation in \ref{eq:bayes} [@wagenmakers2007, p. 796].

\begin{equation}
\label{eq:bayes}
BF_{01} \approx exp(\Delta{}BIC_{10}/2)
\end{equation}

where $\Delta{}BIC_{10} = BIC_1 - BIC_0$, $BIC_1$ is the BIC of the full model, and $BIC_0$ is the BIC of the null model.
Values of $BF_{01} > 1$ indicate a preference of H\textsubscript{0} over H\textsubscript{1}.
The interpretation of the Bayes factors follows the recommendations in @raftery1995 [p. 139].

The extracted measurements were filtered before statistical analysis.
Measures of vowel duration, closure duration, word duration, and RR duration that are 3 standard deviations lower or higher than the respective means were excluded from the final dataset.
This operation yields a total of 920 tokens of vowel and closure durations, 1176 tokens of word duration, and 848 tokens of RR duration.

# Results

The following sections report the results of the study in relation to the durations of vowels, consonant closure, word, and the Release to Release interval.
When discussing the output of statistical modelling, only the relevant predictors and interactions will be presented.
For the full output of the models and *p*-values, see \Cref{a:stats}.

<!-- INCLUDE STANDARD ERROR! -->
<!-- include CI? -->
<!-- include tables of lmer summary in appendix -->
<!-- change results to full model, no sense in using a new model with significant predictors -->
<!-- include raw sample means? -->

## Vowel duration

```{r vow-lm}
vow_lm <- lmer(
  vowel_duration ~
    c2_phonation *
    vowel *
    language +
    c2_place +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered
)
```

```{r vowels-plot, include=TRUE, fig.cap = "Vowel duration in Italian and Polish.", fig.lp="f:", out.extra="", cache=TRUE, dependson="read-data"}
durations_filtered %>%
  ggplot(aes(vowel, vowel_duration, fill = c2_phonation)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(), alpha = 0.4) +
  facet_grid(~ language) +
  labs(x = "Vowel", y = "Vowel duration (ms)") +
  scale_fill_discrete(name = "C2 voicing")
```

\Cref{f:vowels-plot} shows boxplots and the raw data of vowel duration in Italian (on the left) and Polish (on the right) for the three vowels /a, o, u/.
Vowel tend to be longer when followed by a voiced stop both in Italian and Polish.
The effect appears to be greater in Italian than in Polish, especially for the vowels /a/ and /o/
There is no clear effect of C2 voicing in /u/ in Italian, but the effect is discernible in Polish /u/.

A linear mixed-effects model with vowel duration as the outcome variable was fitted with the following predictors: fixed effects for C2 voicing (voiceless, voiced), C2 place of articulation (coronal, velar), vowel (a, o, u), language (Italian, Polish), and speech rate (as syllables per second); by-speaker and by-word random intercept with by-speaker random slopes for C2 voicing.
All possible interactions between C2 voicing, vowel, and language were included.
The following terms are significant according to *t*-tests with Satterthwaite's approximation to degrees of freedom: C2 voicing, vowel, language, and speech rate.
Only the interaction between C2 voicing and vowel is significant.
Vowels are 19 ms longer (se = 4.4) when followed by a voiced stop (C2 voicing).
The effect of C2 voicing is smaller with /u/ (around 5 ms, $\hat{\beta}$ = -14.4 ms, se = 6).
Polish has on average shorter vowels than Italian ($\hat{\beta}$ = -28 ms, se = 8), and the effect of voicing is estimated to be about 11 ms (although recall that the interaction between language and C2 voicing is deemed not significant).
Speech rate has unsurprisingly a negative effect on vowel duration, such that faster rates correlate with shorter vowel durations ($\hat{\beta}$ = -15 ms, se = 1).

## Consonant closure duration

```{r clo-lm}
clo_lm <- lmer(
  closure_duration ~
    c2_phonation *
    vowel *
    language +
    c2_place +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered
)
```

```{r closure-plot, echo=FALSE, include=TRUE, fig.cap = "Stop closure duration in Italian and Polish.", fig.lp="f:", out.extra="", cache=TRUE, dependson="read-data"}
durations_filtered %>%
  ggplot(aes(vowel, closure_duration, fill = c2_phonation)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(), alpha = 0.4) +
  facet_grid(~ language) +
  labs(x = "Vowel", y = "Closure duration (ms)") +
  scale_fill_discrete(name = "C2 voicing")
```

\Cref{f:closure-plot} illustrates stop closure durations with boxplots and individual raw data points.
A pattern opposite to that with vowel duration can be noticed: Closure duration is shorter for voiced than for voiceless stops.
The same model specification as with vowel duration has been fitted with consonant closure durations as the outcome variable.
C2 voicing, C2 place, and speech rate are significant.
Stop closure is 16.5 ms shorter (se = 3) if the stop is voiced and 3.5 ms longer (se = 1.5) if velar.
Finally, faster speech rates correlate with shorter closure durations ($\hat{\beta}$ = -8.5 ms, se = 1 ms).

## Vowel and closure duration

```{r vow-clo-lm, include=FALSE}
vow_clo_lm <- lmer(
  vowel_duration ~
    closure_duration *
    vowel +
    syl_rate +
    (1+closure_duration|speaker) +
    (1|item),
  data = durations_filtered
)
summary(vow_clo_lm)
```

```{r vow-clo-plot, include=TRUE, fig.cap = "Linear regression of closure and vowel duration per vowel.", fig.lp="f:", out.extra="", cache=TRUE, dependson="read-data"}
as_tibble(effect("closure_duration:vowel", vow_clo_lm)) %>%
  ggplot(aes(closure_duration, fit)) +
  geom_point(data = durations_filtered, aes(y = vowel_duration), alpha = 0.1) +
  geom_line() +
  geom_ribbon(aes(ymax = upper, ymin = lower), alpha = 0.2) +
  geom_rug(data = durations_filtered, aes(y = vowel_duration), alpha = 0.1) +
  facet_grid(~ vowel) +
  labs(x = "Closure duration (ms)", y = "Vowel duration (ms)")
```

A model addressing the relationship between vowel and stop closure duration was fitted with the following terms and interactions: vowel duration as the outcome variable; as fixed effects, closure duration, vowel, speech rate; an interaction between closure duration and vowel; by-speaker and by-word random intercepts, and by-speaker random slopes for C2 voicing.
Closure duration has a significant effect on vowel duration ($\hat{\beta}$ = -0.15 ms, se = 0.06 ms).
The effect with /u/ is greater than with /a/ and /o/ ($\hat{\beta}$ = -0.35 ms, se = 0.06 ms).
In general, closure duration is inversely correlated with vowel duration.
However such correlation is quite weak.
A 1 ms increase in closure duration corresponds to a 0.2--0.5 ms decrease in vowel duration.
\Cref{f:vow-clo-plot} shows for each of /a, o, u/ the individual data points and the regression lines with confidence intervals extracted from the linear model.

## Word duration

```{r word-bf}
word_lm <- lme4::lmer(
  word_duration ~
    c2_phonation +
    vowel +
    c2_place +
    language +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered,
  REML = FALSE
)

word_lm_null <- lme4::lmer(
  word_duration ~
    # c2_phonation +
    vowel +
    c2_place +
    language +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered,
  REML = FALSE
)

word_bf <- round(exp((BIC(word_lm) - BIC(word_lm_null)) / 2))
```

The following full and null models were fitted to test for the effect of C2 voicing on word duration.
The full model has the following fixed effects: C2 voicing, C2 place, vowel, speech rate, and language.
The model also includes by-speaker and by-word random intercepts, and a by-speaker random slope for C2 voicing.
The null model excludes the fixed effect of C2 voicing.
The Bayes factor of the null model against the full model is `r word_bf`.
Thus, the null model (in which the effect of C2 voicing is 0) is `r word_bf` times more likely under the observed data than the full model.
This indicates that there is strong evidence for word duration not being affected by C2 voicing.

## Release to Release interval duration

```{r rr-plot, include=TRUE, fig.cap="Release to Release interval duration.", fig.extra="", cache=TRUE, dependson="read-data", warning=FALSE}
rel_filtered %>%
  ggplot(aes(c2_phonation, rel_rel)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1, width = 0.3) +
  facet_grid(~ language) +
  labs(x = "C2 voicing", y = "Release to Release duration (ms)")
```

```{r rr-bf}
rr_lm <- lme4::lmer(
  rel_rel ~
    c2_phonation +
    vowel +
    c2_place +
    language +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered,
  REML = FALSE
)

rr_lm_null <- lme4::lmer(
  rel_rel ~
    # c2_phonation +
    vowel +
    c2_place +
    language +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  data = durations_filtered,
  REML = FALSE
)

rr_bf <- round(exp((BIC(rr_lm) - BIC(rr_lm_null)) / 2))
```

In \Cref{f:rr-plot}, boxplots show the durations of the Release to Release interval in words with a voiceless vs. a voiced C2 stop, in Italian (left side) and Polish (right side).
It can be seen, also from the single data points, that the distributions and main statistics of the durations in the two conditions do not differ much within both languages.
The models specifications for the Release to Release duration are the same as for word duration.
The Bayes factor of the null model against the full model is `r rr_bf`, which means that the null model (without C2 voicing) is `r rr_bf` times more likely than the full model.
The data suggests there is positive evidence that duration of the RR interval is not affected by C2 voicing.

# Discussion

The data and statistical analysis presented in the previous section indicate that the duration of interval between the releases of two consecutive consonants in CV́CV words (the Release to Release interval) is insensitive to the phonological voicing of the second consonant (C2).
In accordance with a compensatory temporal adjustment account [@slis1969; @lehiste1970], the difference in vowel duration before voiceless vs. voiced stops is the outcome of differences in stop closure duration.
More specifically, the timing of the closure onset of C2 within the invariant Release to Release interval determines the duration of the preceding vowel.
An earlier closure onset (like in the case of voiceless stops), relative to the onset of the preceding vowel, causes the vowel to be shorter.
On the other hand, a later closure onset (like with voiced stops) produces a longer vowel.
Hence, the voicing effect can be seen as a product of gestural timing, rather then a consequence of characteristics of voicing *per se*.

A limitation of the compensatory temporal adjustment account, in the variant of @slis1969 and @lehiste1970, is the lack of a precise identification of the internal structure of the word as the unit within which the compensation happens.
It is not clear, for example, why compensation should target the preceding stressed vowel, rather then the following unstressed vowel or any other segment in the word.
The invariance of Release to Release interval allows us to formulate a better defined internal organisation of the word.
Since the Release to Release interval contains (in a simplified view) just the vowel (broadly defined as a vocoid gesture) and the consonant closure, it follows by logic that differences in closure duration must affect the preceding vowel.

The invariance of the Release to Release interval can be seen as the explanans of the voicing effect, and it gives the compensatory account a more logical consistency by narrowing the scope of the temporal adjustment action.
On the other hand, the Release to Release invariance becomes in turn an explanandum and it thus requires to be motivated.
In the following section, I offer a gestural organisation account that allows the invariance or such interval to follow from the relative timing of the articulatory gestures in a CV́C sequence.

## Gestural alignment
\label{s:gestural}

<!-- need to introduce C-centre theory and say that although this is only acoustics, people have shown to be links between articulation and acoustics in this sense (somebody cited somewhere) -->

\begin{figure}
  \centering
  \includegraphics{img/gorganisation.pdf}
  \caption{Gestural organisation patterns for onsets (a), codas (b), heterosyllabic onsets (c). See \Cref{s:gestural} for details. Based on \citet{marin2010}.}
  \label{f:gorganisation}
\end{figure}

According to the coupled oscillator model of syllabic structure [@browman1988; @browman2000; @goldstein2006; @goldstein2014], articulatory gestures can be timed according to two coupling modes: in-phase (synchronous) mode, by which two gestures start in synchrony, or anti-phase (sequential) mode, in which one gesture starts when the preceding one has reached its target.
@marin2010 showed that onset consonants in American English are in-phase with the vowel nucleus and anti-phase with each other.
Such phasing pattern establishes a stable relationship between the centre of the consonant or consonant cluster and the following vowel.
Independent of the number of onset consonants, the midpoint of the onset, the so-called 'C-centre', is maintained at a fixed distance from the vowel, such that increasing number of consonants in the onset does not change the C-centre/vowel distance (\Cref{f:gorganisation}(a)).
On the other hand, coda consonants are timed anti-phase with the preceding vowel and between themselves.
Stability in codas is seen in the lag between the vowel and the left-most edge of the coda, which is not affected by the number of coda consonants (\Cref{f:gorganisation}(b)).
Other studies found further evidence for the synchronous and sequential coupling modes (see extensive review in @marin2010 and @marin2014), although the use of one mode over the other depends on the language and the consonants under study.

Consonants can thus be said to follow either a C-centre organisation pattern or a left-edge organisation pattern.
In both cases, of course, the pattern is relative to the tautosyllabic vowel (the following vowel for onsets, the preceding vowel for codas).
To the best of my knowledge, no study has reported the timing of onset consonants relative to the *preceding* (heterosyllabic) vowel.
The results from this acoustic study on Italian and Polish are compatible with a right-edge organisation pattern for onset consonants and preceding stressed vowels \Cref{f:gorganisation}(c).
The release of C2 (which is the onset of the second syllable in CV́CV words)---which can be thought as the acoustic parallel of the articulatory right edge of C2---is invariantly timed relative to V1 (which is the nucleus of the first syllable).

A consequence of a right-edge organisation pattern of C2 relative to V1 in CV́CV words is that differences in C2 closure duration do not affect the lag between V1 and the release of C2, as shown by the results of this study.
The invariance of the lag between the release of C1 and that of C2 then can be seen to follow from the invariance in timing between, on the one hand, C1 (which is always /p/ in this study) and V1, and, on the other, between V1 and the right edge of C2.

<!-- maybe footnote on circularity of argument of syllabic structure and gestural timing -->
<!-- also, all data so far is stressed vowels? -->
<!-- indeed we don't know much about how syllables are concatenated? -->

A right-edge organisation account is compatible with findings from electromyographic, x-ray microbeam, and ultrasonic data by, respectively, @raphael1975, @de-jong1991, and @celata2018.
@celata2018 show that vowels before tautosyllabic clusters have the same duration as before heterosyllabic clusters.
However, vowels followed by geminates are shorter than when followed by singletons, although from a syllabic structure point of view geminates correspond to heterosyllabic clusters and singletons to tautosyllabic clusters (i.e., V-final syllables followed by singletons and tautosyllabic clusters are open, while those followed by geminates and heterosyllabic clusters are closed).
@celata2018 argue that these results corroborate a rhythmic account in which the relevant unit is the rhythmic syllable, i.e. the VC(C) sequence (independent of the traditional syllabic structure), which is kept constant.
Such view reflects a gestural timing view in which the timing of the right edge of the consonant is held constant relative to the vowel.

@de-jong1991 reports that the closing gesture of voiceless stops (following stressed vowels) is faster than that of voiced stops, and that also it is timed earlier with respect to the opening gesture of the stressed vowel.
According to @de-jong1991, the differences in vowel duration are driven by the timing of the consonantal closing gesture relative to the vocalic opening gesture (also see \citealt{hertrich1997}).
<!-- The view proposed above reflects these findings. -->
Moreover, the data in @de-jong1991 show that the final portion of the vocalic opening gesture is prolonged before voiced stops.
This finding corresponds to what @raphael1975 reported based on electromyographic data.
The electromyographic signal corresponding to the vocalic gesture reaches its plateaux at the same time in the voiceless and voiced context, but the plateaux is held for longer in the case of vowels followed by voiced stops, indicating that muscular activation is kept for longer.

These studies taken together, plus the results from this study, bring evidence to the view that two factors contribute to the difference in vowel duration observed before consonants varying in their voicing specification.
These two factors are: (1) the right-edge alignment of coda consonants following stressed vowels relative to the latter, and (2) the differential timing of the closing gesture onset for voiceless vs. voiced stops.
These two factors together can be synthesised into a compensatory temporal adjustment account, in which the fixed interval is generated by factor (1) and the temporal adjustment is brought about by factor (2).

## Limitations and future work

The generalisations reported in this paper strictly apply to disyllabic words with a stressed vowel in the first syllable.
It is possible that the organisation pattern found in this context does not occur in sequences including an unstressed vowel.
For example, it is known that the difference in closure duration between voiceless and voiced stops is not stable when the stops precede a stressed vowel, although the vowels preceding the pre-stress stops have different durations [@davis1989].
According to the gestural interpretation given here, the absence in differences of closure duration should correspond to no difference in vowel duration.
Data from different contexts and different languages is thus needed to assess the generality of the claims put forward in this paper.

The constraints on experimental material enforced by the use of ultrasound tongue imaging have been previously mentioned in \Cref{s:materials}.
Given these constraints, temporal information from other vowels (like front vowels) and places of articulation is a desideratum.
\Cref{s:gestural} discusses the interpretation of the Release to Release invariance in CV́CV words as a consequence of the timing of C2 rather than of a holistic CV́C motor plan in which the RR interval is held constant.
Although beyond the scope of this paper, disambiguating between these two interpretations on articulatory grounds is fundamental for a general understanding of a theory of gestural organisation.

The compensatory temporal adjustment account presented here extends to other durational effects discussed in the literature.
In particular, the account bears predictions on the direction of the durational difference led by phonation types different from voicing, like aspiration and ejection.
For example, the mix of results with regard to the effect of aspiration [@durvasula2012] suggests that the conditions for a temporal adjustment might differ across the contexts and languages studied.
In light of the results in @begus2017, future studies will have to investigate the durational invariance of speech intervals in relation to a variety of phonation contrasts.

# Conclusion



\begin{acknowledgments}
Thanks to...
\end{acknowledgments}

\appendix

# Output of statistical models
\label{a:stats}

## Vowel duration

\scriptsize
```{r vow-table}
kable(tidy(vow_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect), format = "latex", booktabs = TRUE, digits = 4)
```
\normalsize

## Closure duration

\scriptsize
```{r clo-table}
kable(tidy(clo_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect), format = "latex", booktabs = TRUE, digits = 4)
```
\normalsize

## Vowel and closure duration

\scriptsize
```{r vow-clo-table}
kable(tidy(vow_clo_lm, effect = "fixed", conf.int = TRUE) %>% select(-effect), format = "latex", booktabs = TRUE, digits = 4)
```
\normalsize

# Socio-linguistic information of participants
\label{a:socioling}

<!-- say why there are missing IDs -->

\ctable[caption = Participants' sociolinguistic information.,
label = t:socio,
width=\textwidth,
star,
doinside = \footnotesize
]{llll>{\raggedright}p{5cm}lll}{}{
\FL
ID & Age & Sex    & Native L & Other Ls& City of birth & Spent most time in & > 6 mo \ML
it01 & 29 & Male   & Italian & English, Spanish                          & Verbania            & Verbania        & Yes \NN
it02 & 26 & Male   & Italian & Friulian, English, Ladin-Venetan          & Udine               & Tricesimo & Yes \NN
it03 & 28 & Female & Italian & English, German                           & Verbania            & Verbania        & No  \NN
it04 & 54 & Female & Italian & Calabrese                                 & Verbania            & Verbania        & No  \NN
it05 & 28 & Female & Italian & English                                   & Verbania            & Verbania        & No  \NN
it09 & 35 & Female & Italian & English                                   & Vignola & Vignola         & Yes \NN
it11 & 24 & Male   & Italian & English                                   & Monza               & Monza           & Yes \NN
it13 & 20 & Female & Italian & English, French, Arabic, Farsi            & Ancona              & Chiaravalle     & Yes \NN
it14 & 32 & Male & Italian & English, Spanish & Frosinone & Frosinone & Yes \NN
pl02 & 32 & Female & Polish  & English, Norwegian, French, German, Dutch & Koło                & Poznań          & Yes \NN
pl03 & 26 & Male   & Polish  & Russian, English, French, German          & Nowa Sol            & Poznań          & Yes \NN
pl04 & 34 & Female & Polish  & Spanish, English, French                  & Warsaw              & Warsaw          & No  \NN
pl05 & 42 & Male   & Polish  & English, French                           & Przasnysz           & Warsaw        & No  \NN
pl06 & 33 & Male   & Polish  & English                                   & Zgierz              & Zgierz          & Yes \NN
pl07 & 32 & Female & Polish  & English, Russian                          & Bielsk Podlaski     & Bielsk Podlaski & Yes \LL
}

# Target words
\label{a:targets}

\ctable[caption = Target words.,
label = t:targets,
star,
doinside = \footnotesize
]{lllllll}{}{
\FL
Italian   &        &      & \hspace{0.5cm} & Polish &      &      \NN
\cmidrule{1-3}\cmidrule{5-7}
pata      & poto*  & putu & & pata   & poto & putu \NN
pada      & podo   & pudu & & pada*  & podo & pudu \NN
paca*     & poco*  & pucu & & paka*  & poko & puku \NN
paga*     & pogo   & pugu & & paga   & pogo & pugu \LL
}
