---
title: This is a title and this is too
shorttitle: A subtitle goes on another line
author:
- name: Stefano Coretta
  affiliation: The University of Manchester
  email: stefano.coretta\@manchester.ac.uk
  thanks: other info
output:
  jasaarticle::jasa_article:
    citation_package: natbib
abstract: |
  Put your abstract here.
bibliography: linguistics.bib
header-includes:
- \usepackage{cleveref}
- \usepackage{ctable}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(effects)
```

```{r read-data, include=FALSE, message=FALSE}
speakers <- read_csv("./datasets/speakers.csv")
stimuli <- read_csv("./datasets/stimuli.csv")

durations <- list.files(path = "./datasets",
                   pattern = "*-durations.csv",
                   full.names = TRUE) %>%
    map_df(~read_csv(., na = "--undefined--")) %>%
    left_join(y = speakers) %>%
    left_join(y = stimuli) %>%
    mutate_if(is.character, as.factor) %>%
    mutate(
        c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced"))
    )
```


# Method

## Speakers

## Equipment

The acquisition of the audio signal was achieved with the software Articulate Assistant Advanced™ (AAA, v2.17.2) running on a Hawlett-Packard ProBook 6750b laptop with Microsoft Windows 7, with a sample rate of 22050 MHz (16-bit) in a proprietary format.
A FocusRight Scarlett Solo pre-amplifier and a Movo LV4-O2 Lavalier microphone were used for audio recording.

## Materials

The target stimuli were disyllabic words with C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{2} structure, where C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/, and V\textsubscript{2} = V\textsubscript{1} (e.g. /pata/, /pada/, /poto/, etc.).
The lexical stress of the target words was placed by speakers of both Italian and Polish on V\textsubscript{1}, as intended.
The make-up of the target words was constrained by the design of the experiment, which included ultrasound tongue imaging (UTI).
Front vowels are difficult to image with UTI, since their articulation involves tongue positions which are particularly far from the ultrasonic probe, hence reducing the visibility of the tongue contour.
For this reason, only central and back vowels were included.
Since one of the variables of interest in the exploratory study was the closing gesture of C\textsubscript{2}, only lingual consonants were used.
A labial stop was chosen as the first consonant to reduce possible coarticulation with the following vowel (although see \citealt{vazquez-alvarez2007}).
The target words were embedded in a frame sentence, *Dico X lentamente* 'I say X slowly' in Italian, and *Mówię X teraz* 'I say X now' in Polish.
These sentences were chosen in order to keep the placement of stress and emphasis similar across languages, so to ensure comparability of results.

## Procedure

The participant was asked to read the sentences with the target words which were sequentially presented on the computer screen.
The order of the sentence stimuli was randomised for each participant.
Each participant read the list of randomised sentence stimuli six times.
Due to software constraints, the order of the list was kept the same across the six repetitions within each participant.
Each speaker read a total of 72 sentences, with a grand total of 576 tokens (288 per language).
The reading task lasted between 15 and 20 minutes, with optional short breaks between one repetition and the other.

## Data processing, measurements, and statistical analysis

\ctable[caption = List of measurements as extracted from acoustics.,
label = t:dur-measures,
width=\textwidth,
star
]{ll>{\raggedright}p{9cm}}{}{
\FL
\textbf{landmark}               &                  & \textbf{criteria}                                                                                    \ML
vowel onset           & (V1 onset)         & appearance of higher formants in the spectrogram following the burst of /p/ (C1)            \NN
vowel offset          & (V1 offset)        & disappearance of the higher formants in the spectrogram preceding the target consonant (C2) \NN
consonant onset       & (C2 onset)         & corresponds to V1 offset                                                                    \NN
closure onset         & (C2 closure onset) & corresponds to V1 offset                                                                    \NN
consonant offset      & (C2 offset)        & appearance of higher formants of the vowel following C2 (V2); corresponds to V2 onset                                \NN
consonant burst onset & (C2 burst)         & automatic detection \citep{ananthapadmanabha2014}                                           \LL
}

The audio recordings were exported from AAA in `.wav` format for further processing.
A forced aligned transcription was accomplished through the SPeech Phonetisation Alignment and Syllabification software (SPPAS) [@bigi2015].
The outcome of the automatic annotation was manually corrected when necessary, according to the criteria in \Cref{t:dur-measures}.
The releases of C1 and C2 were detected automatically by means of a Praat scripting implementation of the algorithm described in @ananthapadmanabha2014.
The durations in milliseconds of the following intervals were extracted from the annotated acoustic landmarks with Praat scripting: vowel duration (V1 onset to V1 offset), consonant closure duration (V1 offset to C2 burst), and C1 release to C2 release duration (Rel-Rel duration).

The durational measurements were analysed with linear mixed-effects models using `lme4` in R [@r-core-team2017; @bates2015].
*P*-values were obtained with likelihood ratio tests comparing the full model with a nested model without the relevant predictor.

\begin{acknowledgments}
Thanks to...
\end{acknowledgments}

\appendix
\section{Optional appendix}
