---
title: "Tongue root advancement and vowel duration: a gradient effect?"
author: "Stefano Coretta"
date: "05/03/2018"
output:
  beamer_presentation:
    citation_package: natbib
    highlight: zenburn
    latex_engine: xelatex
fontsize: 12pt
bibliography: linguistics.bib
biblio-style: unified.bst
header-includes:
- \frenchspacing
- \usepackage{cleveref}
- \usetheme{metropolis}
- \usecolortheme[snowy]{owl}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(tidyverse)
theme_set(theme_minimal(base_size = 20))
library(lme4)
library(knitr)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r read-data, include=FALSE}
speakers <- read_csv("./presentations/2018-baap/datasets/speakers.csv")
stimuli <- read_csv("./presentations/2018-baap/datasets/stimuli.csv")

vowels <- list.files(
  path = "./presentations/2018-baap/datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced"))) %>%
  mutate_if(is.character, as.factor)
```

# Background

* **Correlates of voicing**
    * shorter VOT [@westbury1983;@davidson2016;@abramson2017]
    * tongue root advancement TRA [@westbury1983;@ohala2011]
    * correlation VOT ~ TRA [@ahn2015]
    * longer vowel duration [@house1953;@peterson1960;@chen1970;@klatt1973;@lisker1974;@fowler1992;@lampp2004]

* *Relation between vowel duration and TRA*

# Background

* **Voicing effect** (VE): vowels are longer when followed by voiced stops
    * *Italian*: voicing effect of 35 msec [@farnetani1986]
    * *Polish*: mixed results
        * @keating1984: no effect
        * @nowak2006 PhD dissertation: 4.5 msec effect
* Timing of laryngeal and tongue activity
    * simultaneous UTI + EGG + audio

# Methods (a summary)

* **Participants**: 4 Italians (2 F, 2 M), 4 Polish (2 F, 2 M)
* **Procedure**:
    * simultaneous ultrasound tongue imaging and audio recording
    * stabilisation headset [@articulate2008]
* **Materials**:
    * C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{1}
        * C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/
        * *pata*, *pada*, *paka*, ..., *poto*, *podo*, ...
        * stress on first syllable
    * frame sentence
        - *Dico X lentamente*, 'I say X slowly'
        - *Mówię X teraz*, 'I say X now'
        - no pauses between words

# Results: Vowel duration

```{r vowel-duration-plot, echo=FALSE}
vowels %>%
  group_by(speaker) %>%
  mutate(vowel_duration = scale(vowel_duration)) %>%
  ggplot(aes(vowel, vowel_duration, fill = c2_phonation)) +
  geom_boxplot() +
  facet_grid(~ language, labeller = labeller(language = c(italian = "Italian", polish = "Polish"))) +
  xlab("Vowel") +
  ylab("Vowel duration (scaled)") +
  scale_fill_manual(name = "C2 voicing", values = cbPalette[2:3])
```

# Results: Vowel duration

```{r vow-it-lm, include=FALSE}
vowels_it <- filter(vowels, language == "italian")

vowels_it_lm <- lmer(
  vowel_duration ~
    c2_phonation +
    c2_place +
    vowel +
    c2_phonation:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_it,
  REML = FALSE
)

vowels_it_lm_null <- lmer(
  vowel_duration ~
#    c2_phonation +
    c2_place +
    vowel +
#    c2_phonation:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_it,
  REML = FALSE
)

summary(vowels_it_lm)$coeff

vowels_it_mc <- anova(vowels_it_lm_null, vowels_it_lm)
```

```{r vow-pl-lm, include=FALSE}
vowels_pl <- filter(vowels, language == "polish")

vowels_pl_lm <- lmer(
  vowel_duration ~
    c2_phonation +
    c2_place +
    vowel +
    c2_phonation:vowel +
    c2_place:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_pl,
  REML = FALSE
)

vowels_pl_lm_null <- lmer(
  vowel_duration ~
#    c2_phonation +
    c2_place +
    vowel +
#    c2_phonation:vowel +
    c2_place:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_pl,
  REML = FALSE
)

summary(vowels_pl_lm)$coeff

vowels_pl_mc <- anova(vowels_pl_lm_null, vowels_pl_lm)
```

* Linear mixed-effects models [@bates2015;@kuznetsova2016]

* **Italian**
    - `voicing + place + vowel + sentence duration + voicing:vowel`
    - `(1+voicing|speaker) + (1|word)`
    - $\beta$ = `r round(summary(vowels_it_lm)$coeff[2, 1])` msec, $\chi^2$(`r vowels_it_mc[2,7]`) = `r round(vowels_it_mc[2,6], 2)`, p = `r vowels_it_mc[2,8]`

* **Polish**
    - `voicing + place + vowel + sentence duration + voicing:vowel + place:vowel`
    - `(1+voicing|speaker) + (1|word)`
    - $\beta$ = `r round(summary(vowels_pl_lm)$coeff[2, 1])` msec, $\chi^2$(`r vowels_pl_mc[2,7]`) = `r round(vowels_pl_mc[2,6], 2)`, p = `r vowels_pl_mc[2,8]`

# Results: Vowel duration

```{r vowels-lm-coef}
as_tibble(rbind(coef(vowels_it_lm)$speaker[2], coef(vowels_pl_lm)$speaker[2]), rownames = "speaker") %>%
  mutate(language = c(rep("Italian", 4), rep("Polish", 4))) %>%
  ggplot(aes(speaker, c2_phonationvoiced, fill = language)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(name = "Language", values = cbPalette[2:3]) +
  labs(title = "By-speaker random coefficients", y = "msec")
```

# Results: Tongue contours

* **Midsagittal tongue contours**
    * polar coordinates [@heyne2015a;@heyne2015;@mielke2015]
    * two timepoints:
        * at acoustic closure onset
        * at maximum displacement [within closure, @strycharczuk2015]

* **Generalised additive mixed effects models** [@wood2006;@soskuthy2017;@van-rij2017]
    * `rticulate` R package [@coretta2018;@coretta2018a]

* **General trends**
    * idiosyncratic use of TRA
    * 2 speakers with relatively greater TRA

# Results: Tongue contours


# Discussion

* Results summary
    * **effect of voicing on vowel duration**
        * Italian: +22 msec
        * Polish: +12 msec
    * **tongue contours**
        * 4 of 8 speakers (IT01, IT02, IT03, PL05) show TRA at maximum displacement
        * 2 of 8 (IT01, IT02) also at closure onset
    * 2 speakers (IT02, PL05) with stronger VE and greater TRA

# Discussion

* **New proposal**: *Longer vowel duration allows for greater tongue root advancement*.

* Similar reasoning to that of @halle1967
    * longer vowels allow for laryngeal adjustments from spontaneous voicing of vowels to obstruent voicing of voiced consonants.

* If the new proposal is correct, we might see a positive correlation between vowel duration and degree of TRA.

# Discussion: Vowel Duration ~ TRA

\begin{figure}
  \includegraphics{./graphics/tra-vow.png}
\end{figure}

