---
title: "Tongue root advancement and vowel duration: a gradient effect?"
author: "Stefano Coretta"
date: "05/03/2018"
output:
  beamer_presentation:
    citation_package: natbib
    highlight: zenburn
    latex_engine: xelatex
bibliography: linguistics.bib
biblio-style: unified.bst
header-includes:
- \frenchspacing
- \usepackage{cleveref}
- \usetheme{metropolis}
- \usecolortheme[snowy]{owl}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(tidyverse)
theme_set(theme_minimal())
library(lme4)
library(knitr)
```

```{r read-data, include=FALSE}
speakers <- read_csv("./presentations/2018-baap/datasets/speakers.csv")
stimuli <- read_csv("./presentations/2018-baap/datasets/stimuli.csv")

vowels <- list.files(
  path = "./presentations/2018-baap/datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced"))) %>%
  mutate_if(is.character, as.factor)
```

# Background

\begin{figure}
\includegraphics[width = \textwidth]{graphics/voicing.pdf}
\end{figure}

# Research hypotheses

* \textbf{H$_1$}: Vowels are longer if followed by voiced stops in Italian, but not in Polish.
* \textbf{H$_2$}: Italian speakers, but not Polish speakers, produce voiced consonants with tongue root advancement.

# Methods

* **Participants**: 4 Italians (2 F, 2 M), 4 Polish (2 F, 2 M)
* **Procedure**:
  * simultaneous ultrasound tongue imaging and audio recording
* **Materials**:
  * C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{1}
    * C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o/, C\textsubscript{2} = /t, d, k, g/
    * *pata*, *pada*, *paka*, ..., *poto*, *podo*, ...
  * frame sentence
    - *Dico X lentamente*, 'I say X slowly'
    - *Mówię X teraz*, 'I say X now'

# Results: vowel duration

```{r vow-it-lm, include=FALSE}
vowels_it <- filter(vowels, language == "italian")

vowels_it_lm <- lmer(
  vowel_duration ~
    c2_phonation +
    c2_place +
    vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_it,
  REML = FALSE
)

vowels_it_lm_null <- lmer(
  vowel_duration ~
#    c2_phonation +
    c2_place +
    vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_it,
  REML = FALSE
)

anova(vowels_it_lm_null, vowels_it_lm)
```

```{r vow-it-lm-coeff}
kable(summary(vowels_it_lm)$coeff)
```
