---
title: "Tongue root advancement and vowel duration: a gradient effect?"
author: "Stefano Coretta"
date: "The University of Manchester"
institute: "13 April, BAAP 2018 (Canterbury)"
output:
  beamer_presentation:
    citation_package: natbib
    highlight: zenburn
    latex_engine: xelatex
fontsize: 12pt
bibliography: linguistics.bib
biblio-style: unified.bst
header-includes:
- \frenchspacing
- \usepackage{cleveref}
- \usetheme{metropolis}
- \usecolortheme[snowy]{owl}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(tidyverse)
theme_set(theme_minimal(base_size = 20))
library(rticulate)
library(tidymv)
library(lme4)
library(effects)
library(knitr)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r read-data, include=FALSE}
speakers <- read_csv("./presentations/2018-baap/datasets/speakers.csv")
stimuli <- read_csv("./presentations/2018-baap/datasets/stimuli.csv")

vowels <- list.files(
  path = "./presentations/2018-baap/datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced"))) %>%
  mutate_if(is.character, as.factor)

duration <- list.files(
  path = "./presentations/2018-baap/datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  select(-speaker)

columns <- c(
  "speaker",
  "seconds",
  "rec_date",
  "prompt",
  "label",
  "TT_displacement_sm",
  "TT_velocity",
  "TT_velocity_abs",
  "TD_displacement_sm",
  "TD_velocity",
  "TD_velocity_abs",
  "TR_displacement_sm",
  "TR_velocity",
  "TR_velocity_abs"
)

tra_series <- list.files(
  path = "./presentations/2018-baap/datasets/ultrasound/",
  pattern = "*-vowel-series.tsv",
  full.names = TRUE
) %>%
  read_aaa(., columns, format = "wide") %>%
  select(-(X_1:Y_42)) %>%
  left_join(duration) %>%
  mutate(word = word(prompt, 2)) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(
    c2_phonation = ordered(c2_phonation, levels = c("voiceless", "voiced")),
    vowel = ordered(vowel, levels = c("a", "o", "u")),
    c2_place = ordered(c2_place, levels = c("coronal", "velar")),
    proportion = (seconds - v_onset) / (v_offset - v_onset),
    tra = -TR_displacement_sm
  ) %>%
  group_by(speaker) %>%
  mutate(tra_z = scale(tra)) %>%
  ungroup() %>%
  mutate_if(is.character, as.factor) %>%
  mutate(
    speaker_phon = interaction(speaker, c2_phonation),
    speaker_contexts = interaction(speaker, c2_phonation, c2_place, vowel),
    place_vowel = interaction(c2_place, vowel),
    place_vowel_phon = interaction(c2_place, vowel, c2_phonation)
  )

contrasts(tra_series$c2_phonation) <- "contr.treatment"
contrasts(tra_series$vowel) <- "contr.treatment"
contrasts(tra_series$c2_place) <- "contr.treatment"
```

# Background

* **Correlates of voicing**
    * *shorter VOT* [@westbury1983;@davidson2016;@abramson2017]
    * *tongue root advancement TRA* [@westbury1983;@ohala2011]
    * *correlation VOT ~ TRA* [@ahn2015]
    * *longer vowel duration* [@house1953;@peterson1960;@chen1970;@klatt1973;@lisker1974;@fowler1992;@lampp2004]

* **Relation between vowel duration and TRA**

# Background

* **Voicing effect** (VE): vowels are longer when followed by voiced stops
    * *Italian*: voicing effect of 35 msec [@farnetani1986]
    * *Polish*: mixed results
        * @keating1984: no effect
        * @nowak2006 PhD dissertation: 4.5 msec effect
* **Larger study**: relative timing of laryngeal and lingual activity
    * simultaneous UTI + EGG + audio
* **This study**: exploratory, data driven

# Methods (a summary)

* **Participants**: 4 Italians (2 F, 2 M), 4 Polish (2 F, 2 M)
* **Targets**
    * C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{1}
    * C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/
    * *pata*, *pada*, *paka*, ..., *poto*, *podo*, ...
* **Frame sentence**
    - *Dico X lentamente*, 'I say X slowly'
    - *Mówię X teraz*, 'I say X now'
* **Data**
    * durational data from acoustics
    * tongue contours from ultrasound tongue imaging

# Results: Vowel duration

```{r vowel-duration-plot, echo=FALSE}
vowels %>%
  group_by(speaker) %>%
  mutate(vowel_duration = scale(vowel_duration)) %>%
  ggplot(aes(vowel, vowel_duration, fill = c2_phonation)) +
  geom_boxplot() +
  facet_grid(~ language, labeller = labeller(language = c(italian = "Italian", polish = "Polish"))) +
  xlab("Vowel") +
  ylab("Vowel duration (scaled)") +
  scale_fill_manual(name = "C2 voicing", values = cbPalette[2:3])
```

# Results: Vowel duration

```{r vow-it-lm, include=FALSE}
vowels_it <- filter(vowels, language == "italian")

vowels_it_lm <- lmer(
  vowel_duration ~
    c2_phonation +
    c2_place +
    vowel +
    c2_phonation:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_it,
  REML = FALSE
)

vowels_it_lm_null <- lmer(
  vowel_duration ~
#    c2_phonation +
    c2_place +
    vowel +
#    c2_phonation:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_it,
  REML = FALSE
)

summary(vowels_it_lm)$coeff

vowels_it_mc <- anova(vowels_it_lm_null, vowels_it_lm)
```

```{r vow-pl-lm, include=FALSE}
vowels_pl <- filter(vowels, language == "polish")

vowels_pl_lm <- lmer(
  vowel_duration ~
    c2_phonation +
    c2_place +
    vowel +
    c2_phonation:vowel +
    c2_place:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_pl,
  REML = FALSE
)

vowels_pl_lm_null <- lmer(
  vowel_duration ~
#    c2_phonation +
    c2_place +
    vowel +
#    c2_phonation:vowel +
    c2_place:vowel +
    sentence_duration +
    (1+c2_phonation|speaker) +
    (1|word),
  data = vowels_pl,
  REML = FALSE
)

summary(vowels_pl_lm)$coeff

vowels_pl_mc <- anova(vowels_pl_lm_null, vowels_pl_lm)
```

```{r vow-lm-effects, include=FALSE}
vow_it_effects <- as_tibble(
  effect("c2_phonation", vowels_it_lm)
) %>%
  mutate(language = "Italian")

vow_pl_effects <- as_tibble(
  effect("c2_phonation", vowels_pl_lm)
) %>%
  mutate(language = "Polish")

vow_effects <- bind_rows(vow_it_effects, vow_pl_effects)
```

* Linear mixed-effects models [@bates2015;@kuznetsova2016]

* **Italian**: $\beta$ = `r round(summary(vowels_it_lm)$coeff[2, 1])` msec, $\chi^2$(`r vowels_it_mc[2,7]`) = `r round(vowels_it_mc[2,6], 2)`, p = `r vowels_it_mc[2,8]`

* **Polish**: $\beta$ = `r round(summary(vowels_pl_lm)$coeff[2, 1])` msec, $\chi^2$(`r vowels_pl_mc[2,7]`) = `r round(vowels_pl_mc[2,6], 2)`, p = `r vowels_pl_mc[2,8]`

```{r vow-lm-effects-plot, include=FALSE}
ggplot(vow_effects, aes(c2_phonation, fit)) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se), width = 0.2, colour = cbPalette[2]) +
  geom_point() +
  facet_grid(~ language, scales = "free")
```

# Results: Vowel duration

```{r vowels-lm-coef}
as_tibble(rbind(coef(vowels_it_lm)$speaker[2], coef(vowels_pl_lm)$speaker[2]), rownames = "speaker") %>%
  mutate(language = c(rep("Italian", 4), rep("Polish", 4))) %>%
  ggplot(aes(speaker, c2_phonationvoiced, fill = language)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(name = "Language", values = cbPalette[2:3]) +
  labs(title = "By-speaker random coefficients", y = "msec")
```

# Results: Tongue contours

* **Midsagittal tongue contours**
    * from *within consonant closure* [at maximum tongue displacement, @strycharczuk2015], polar coordinates [@heyne2015a;@heyne2015;@mielke2015]

* **Generalised additive mixed effects models (GAMMs)** [@wood2006;@soskuthy2017;@van-rij2017]

* **Polar GAMMs** with the `rticulate` R package [@coretta2018;@coretta2018a]

* **General trends**
    * idiosyncratic use of TRA
    * 2 speakers with relatively greater TRA

# Results: Tongue contours

\begin{figure}
  \includegraphics[width = \textwidth]{./graphics/it01-uti.pdf}
\end{figure}

# Results: Tongue contours

\begin{figure}
  \includegraphics[width = \textwidth]{./graphics/it02-uti.pdf}
\end{figure}

# Results: Tongue contours

\begin{figure}
  \includegraphics[width = \textwidth]{./graphics/pl04-uti.pdf}
\end{figure}

# Results: Tongue contours

\begin{figure}
  \includegraphics[width = 0.8\textwidth]{./graphics/pl05-uti.pdf}
\end{figure}

# Discussion: Results summary

* **Effect of voicing on vowel duration**
    * Italian: +22 msec
    * Polish: +12 msec
* **Tongue contours**
    * 4 of 8 speakers (IT01, IT02, IT03, PL05) show TRA within closure
* **2 speakers** (IT02, PL05) with stronger VE and greater TRA

# Discussion

* **TRA hypothesis**: *Longer vowel duration allows for greater tongue root advancement*.
    * cf. with @halle1967: laryngeal adjustments

* If TRA hypothesis is correct:
    - TRA during the vowel
    - greater TRA in IT02 and PL05

# Discussion: TRA during the vowel

```{r tra-series-gam, include=FALSE, eval=FALSE}
tra_series_gam <- bam(
  tra_z ~
    c2_phonation +
    vowel +
    c2_place +
    s(proportion) +
    s(proportion, by = c2_phonation) +
    s(proportion, by = vowel) +
    s(proportion, by = c2_place) +
#    s(proportion, rec_date, bs = "fs") +
    s(proportion, speaker, bs = "fs") +
    s(proportion, speaker_contexts, bs = "fs"),
  data = tra_series
)

summary(tra_series_gam)
```

```{r tra-series-gam-smooths, eval=FALSE, include=FALSE}
plot_smooths(
  tra_series_gam,
  proportion,
  c2_phonation,
  exclude_random = FALSE,
  conditions = quos(c2_place == "coronal", vowel == "a")
) + facet_wrap(~ speaker, ncol = 4) +
  labs(
    x = "Proportion of vowel duration (ratio)",
    y = "Tongue root advancement (scaled)"
  )
ggsave("./presentations/2018-baap/graphics/tra-series.pdf", width = 14, height = 8)
```

\begin{figure}
  \includegraphics[width = \textwidth]{./graphics/tra-series.pdf}
\end{figure}

# Discussion: Electromiography [EMG, @raphael1975]

<!--
    * electromiography (EMG)
    * sustained muscular activity after EMG peak in vowels followed by voiced stops
    * onset of consonant activity relative to vowel activity offset
-->

\begin{figure}
  \includegraphics[width = \textwidth]{./graphics/raphael.pdf}
\end{figure}

# Discussion

* @raphael1975: sustained muscular activity in voiced consonants
    * *extra time allows more tongue root advancement?*

* Is this a gradual (linear) relationship?
    * we might see a **positive correlation between vowel duration and degree of TRA** (but caveat!)

# Discussion: Vowel Duration ~ TRA

\begin{figure}
  \includegraphics{./graphics/tra-vow.png}
\end{figure}

