---
title: "Tongue contours"
author: "Stefano Coretta"
date: "09/03/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../../"))
library(tidyverse)
theme_set(theme_minimal())
library(tidymv)
library(rticulate)
library(itsadug)
```

# Data import

```{r read-data}
columns <- c(
  "speaker",
  "seconds",
  "rec_date",
  "prompt",
  "label",
  "TT_displacement_sm",
  "TT_velocity",
  "TT_velocity_abs",
  "TD_displacement_sm",
  "TD_velocity",
  "TD_velocity_abs",
  "TR_displacement_sm",
  "TR_velocity",
  "TR_velocity_abs"
)

speakers <- read_csv("./voicing-effect/data/datasets/speakers.csv")
stimuli <- read_csv("./voicing-effect/data/datasets/stimuli.csv")

tongue <- list.files(
  path = "./voicing-effect/data/datasets/ultrasound",
  pattern = "*-tongue-cart.tsv",
  full.names = TRUE
) %>%
  read_aaa(., columns) %>%
  na.omit() %>%
  mutate(word = word(prompt, 2)) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  filter(c1_phonation == "voiceless") %>%
  mutate(
    c2_phonation = ordered(c2_phonation, levels = c("voiceless", "voiced")),
    vowel = ordered(vowel, levels = c("a", "o", "u")),
    c2_place = ordered(c2_place, levels = c("coronal", "velar")),
    speaker_phon = interaction(speaker, c2_phonation)
  ) %>%
  mutate_if(is.character, as.factor)

contrasts(tongue$c2_phonation) <- "contr.treatment"
contrasts(tongue$vowel) <- "contr.treatment"
contrasts(tongue$c2_place) <- "contr.treatment"
```

# IT12

## Maximum displacement

```{r it12-max}
it12_max <- tongue %>%
  filter(
    speaker == "it12",
    label %in% c("max_TT", "max_TD")
  )
```

```{r it12-max-plot}
it12_max %>%
  ggplot(aes(X, Y, colour = c2_phonation)) +
  geom_point(alpha = 0.5) +
  facet_grid(c2_place ~ vowel)
```

```{r it12-max-plot-filt}
it12_max %>%
  filter(X < 35, Y > -40) %>%
  ggplot(aes(X, Y, colour = c2_phonation)) +
  geom_point(alpha = 0.5) +
  facet_grid(c2_place ~ vowel)
```

```{r it12-max-filt}
it12_max_filt <- tongue %>%
  filter(
    speaker == "it12",
    label %in% c("max_TT", "max_TD"),
    X < 35
  )
```

```{r it12-max-gam}
it12_max_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it12_max_filt,
  method = "ML",
  fan_lines = c(15, 22)
)
summary(it12_max_gam)

it12_max_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it12_max_filt,
  method = "ML",
  fan_lines = c(15, 22)
)

compareML(it12_max_gam_null, it12_max_gam)
```

```{r it12-max-smooths}
plot_polar_smooths(
  it12_max_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
)
``` 

IT12 tongue contours do not differ.

## Closure

```{r it12-clos}
it12_clo <- tongue %>%
  filter(
    speaker == "it12",
    label %in% c("closure_"),
    X < 35, Y > -40
  )
```

IT13 closure data are fileterd as the maximum displacement data.

```{r it12-clo-plot}
it12_clo %>%
  ggplot(aes(X, Y, colour = c2_phonation)) +
  geom_point(alpha = 0.5) +
  facet_grid(c2_place ~ vowel)
```

```{r it12-clo-gam}
it12_clo_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it12_clo,
  method = "ML",
  fan_lines = c(15, 22)
)
summary(it12_clo_gam)

it12_clo_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it12_clo,
  method = "ML",
  fan_lines = c(15, 22)
)

compareML(it12_clo_gam_null, it12_clo_gam)
```

```{r it12-clo-smooths}
plot_polar_smooths(
  it12_clo_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
)
```

Nor at closure there are differences.

# IT13

## Maximum displacement

```{r it13-max}
it13_max <- tongue %>%
  filter(
    speaker == "it13",
    label %in% c("max_TT", "max_TD")
  )
```

```{r it13-max-plot}
it13_max %>%
  ggplot(aes(X, Y, colour = c2_phonation)) +
  geom_point(alpha = 0.5) +
  facet_grid(c2_place ~ vowel)
```

```{r it13-max-plot-filt}
it13_max %>%
  filter(Y > -30) %>%
  ggplot(aes(X, Y, colour = c2_phonation)) +
  geom_point(alpha = 0.5) +
  facet_grid(c2_place ~ vowel)
```

```{r it13-max-filt}
it13_max_filt <- tongue %>%
  filter(
    speaker == "it13",
    label %in% c("max_TT", "max_TD"),
    Y > -30
  )
```

```{r it13-max-gam}
it13_max_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it13_max_filt,
  method = "ML"
)
summary(it13_max_gam)

it13_max_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it13_max_filt,
  method = "ML"
)

compareML(it13_max_gam_null, it13_max_gam)
```

```{r it13-max-smooths}
plot_polar_smooths(
  it13_max_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
)
```

c2_phonation improves the fit of the model. There is TRA and tongue dorsum lowering.

## Closure

```{r it13-clos}
it13_clo <- tongue %>%
  filter(
    speaker == "it13",
    label %in% c("closure_"),
    Y > -30
  )
```

IT13 closure data are fileterd as the maximum displacement data (Y > -30).

```{r it13-clo-plot}
it13_clo %>%
  ggplot(aes(X, Y, colour = c2_phonation)) +
  geom_point(alpha = 0.5) +
  facet_grid(c2_place ~ vowel)
```

```{r it13-clo-gam}
it13_clo_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it13_clo,
  method = "ML"
)
summary(it13_clo_gam)

it13_clo_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, by = c2_place, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it13_clo,
  method = "ML"
)

compareML(it13_clo_gam_null, it13_clo_gam)
```

```{r it13-clo-smooths}
plot_polar_smooths(
  it13_clo_gam,
  X,
  c2_phonation,
  facet_terms = c2_place + vowel
)
```

c2_phonation improves the fit of the model. There is TRA and tongue dorsum lowering at closure in IT13.

# IT14

IT14 does not have data for velars and you don't see much of the tongue tip.

## Maximum displacement

```{r it14-max}
it14_max <- tongue %>%
  filter(
    speaker == "it14",
    label %in% c("max_TT", "max_TD")
  )
```

```{r it14-max-plot}
it14_max %>%
  ggplot(aes(X, Y, colour = c2_phonation)) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ vowel)
```

I think the kinematics are so off that it is probably safer not to use this data.

## Closure

```{r it14-clo}
it14_clo <- tongue %>%
  filter(
    speaker == "it14",
    label %in% c("closure_")
  )
```

```{r it14-clo-plot}
it14_clo %>%
  ggplot(aes(X, Y, colour = c2_phonation)) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ vowel)
```

Data at closure are somewhat better, although still not great.

```{r it14-clo-gam}
it14_clo_gam <- polar_gam(
  Y ~
    c2_phonation +
    vowel +
    s(X, bs = "cr") +
    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it14_clo,
  method = "ML"
)
summary(it14_clo_gam)

it14_clo_gam_null <- polar_gam(
  Y ~
#    c2_phonation +
    vowel +
    s(X, bs = "cr") +
#    s(X, by = c2_phonation, bs = "cr") +
    s(X, by = vowel, bs = "cr") +
    s(X, rec_date, bs = "fs", xt = "cr", m = 1),
  data = it14_clo,
  method = "ML"
)

compareML(it14_clo_gam_null, it14_clo_gam)
```

```{r it14-clo-smooths}
plot_polar_smooths(
  it14_clo_gam,
  X,
  c2_phonation,
  facet_terms = vowel
)
```

