---
title: "Tongue root advancement"
author: "Stefano Coretta"
date: "11/03/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../../"))
library(tidyverse)
theme_set(theme_bw())
library(mgcv)
library(tidymv)
library(rticulate)
```

# Read data

```{r read-data}
columns <- c(
  "speaker",
  "seconds",
  "rec_date",
  "prompt",
  "label",
  "TT_displacement_sm",
  "TT_velocity",
  "TT_velocity_abs",
  "TD_displacement_sm",
  "TD_velocity",
  "TD_velocity_abs",
  "TR_displacement_sm",
  "TR_velocity",
  "TR_velocity_abs"
)

speakers <- read_csv("./voicing-effect/data/datasets/speakers.csv")
stimuli <- read_csv("./voicing-effect/data/datasets/stimuli.csv")

duration <- list.files(
  path = "voicing-effect/data/datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  select(-speaker)

tra_series <- list.files(
  path = "./voicing-effect/data/datasets/ultrasound",
  pattern = "*-vowel-series.tsv",
  full.names = TRUE
) %>%
  read_aaa(., columns, format = "wide") %>%
  select(-(X_1:Y_42)) %>%
  left_join(duration) %>%
  mutate(word = word(prompt, 2)) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(
    c2phonation = factor(c2phonation, levels = c("voiceless", "voiced")),
    proportion = (seconds - v_onset) / (v_offset - v_onset),
    tra = -TR_displacement_sm
  ) %>%
  group_by(speaker) %>%
  mutate(tra_z = scale(tra)) %>%
  ungroup() %>%
  mutate_if(is.character, as.factor)

tongue_root <- list.files(
  path = "./voicing-effect/data/datasets/ultrasound/",
  pattern = "*-tongue-cart.tsv",
  full.names = TRUE
) %>%
  map_df(~read_aaa(., column_names = columns, format = "wide")) %>%
  select(-(X_1:Y_42)) %>%
  filter(label == "closure_") %>%
  mutate(word = word(prompt, 2)) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  mutate(
    c2phonation = factor(c2phonation, levels = c("voiceless", "voiced")),
    tra = -TR_displacement_sm
  ) %>%
  mutate_if(is.character, as.factor) %>%
  # filter clear outlier
  filter(TR_displacement_sm > 40) %>%
  group_by(speaker) %>%
  mutate(tra_z = scale(tra)) %>%
  ungroup()
```

# Tongue root advancement at maximum displacement

Tongue root advancement is calculated as the inverse of tongue root displacement along a selected fan line.

```{r tr-density-phonation}
tongue_root %>%
  ggplot(aes(tra_z, fill = c2phonation, linetype = c2phonation, colour = c2phonation)) +
  geom_density(alpha = 0.2)
```

As a general trend, the root is more advanced in the voiced condition.

```{r tr-strip-voicing}
tongue_root %>%
  ggplot(aes(tra_z, c2phonation)) +
  geom_jitter(alpha = 0.2)
```

There are two clear outliers with `tra_z > 4`.

```{r tr-density-language}
tongue_root %>%
    ggplot(aes(tra_z, fill = c2phonation, linetype = c2phonation, colour = c2phonation)) +
    geom_density(alpha = 0.2) +
    facet_grid(. ~ language)
```

The pattern is stronger in Italian than in Polish.
This makes sense, given that most Polish speakers do not show differences in root position.

```{r tra-hist}
tongue_root %>%
  ggplot(aes(tra_z)) +
  geom_histogram(bins = 15)
```

```{r tra-density}
tongue_root %>%
  ggplot(aes(tra_z)) +
  geom_density() +
  geom_rug()
```

```{r tra-density-vowel}
tongue_root %>%
  ggplot(aes(tra_z, fill = vowel, linetype = vowel)) +
  geom_density(alpha = 0.5)
```

/u/ is bimodally distributed.
There is no clear pattern in terms of TRA and vowel height.

```{r tra-density-vowel-language}
tongue_root %>%
  ggplot(aes(tra_z, fill = vowel, linetype = vowel)) +
  geom_density(alpha = 0.5) +
  facet_grid(~ language)
```

Italian /u/ has a big peak after the peak of /a/ but there another peak at values that are inferior of those of /o/.

```{r tra-density-a}
tongue_root %>%
  filter(vowel == "a") %>%
  ggplot(aes(tra_z, fill = c2phonation, linetype = c2phonation)) +
  geom_density(alpha = 0.5) +
  facet_grid(language ~ c2place)
```

```{r tra-density-o}
tongue_root %>%
  filter(vowel == "o") %>%
  ggplot(aes(tra_z, fill = c2phonation, linetype = c2phonation)) +
  geom_density(alpha = 0.5) +
  facet_grid(language ~ c2place)
```

The two outliers are tokens of /o/.

```{r}
tongue_root %>%
  filter(vowel == "u") %>%
  ggplot(aes(tra_z, fill = c2phonation, linetype = c2phonation)) +
  geom_density(alpha = 0.5) +
  facet_grid(language ~ c2place)
```

/u/ shows bimodally distributions but I don't know why.

```{r tra-density-speakers}
tongue_root %>%
  filter(vowel == "u") %>%
  ggplot(aes(tra_z, fill = c2phonation)) +
  geom_density(alpha = 0.5) +
  facet_grid(speaker ~ c2place)
```

```{r tra-boxplot}
tongue_root %>%
  ggplot(aes(c2phonation, tra_z)) +
  geom_violin() +
  geom_boxplot(width = 0.2) +
  facet_grid(. ~ language)
```

```{r tra-boxplot-filtered}
tongue_root %>%
  filter(tra_z < 3) %>%
  ggplot(aes(c2phonation, tra_z)) +
  geom_violin() +
  geom_boxplot(width = 0.2) +
  facet_grid(. ~ language)
```

```{r tra-boxplot-speakers}
tongue_root %>%
  filter(tra < 3) %>%
  ggplot(aes(c2phonation, tra_z)) +
  geom_boxplot() +
  facet_wrap(~ speaker)
```

```{r tra-italian}
tongue_root %>%
  filter(language == "italian") %>%
    ggplot(aes(c2phonation, tra_z)) +
    geom_boxplot() +
    facet_grid(c2place ~ vowel)
```

```{r tra-density-vowels-filtered}
tongue_root %>%
  filter(tra_z < 4) %>%
  ggplot(aes(tra_z, fill = vowel, colour = vowel, linetype = vowel)) +
  geom_density(alpha = 0.5, size = 0.7) +
  scale_linetype_manual(values = c("solid", "dotted", "longdash")) +
  labs(x = "tongue root advancement (z-scores)")
```

```{r tra-density-vowels-place-language}
tongue_root %>%
  filter(tra_z < 4) %>%
  ggplot(aes(tra_z, fill = vowel, colour = vowel, linetype = vowel)) +
  geom_density(alpha = 0.5, size = 0.7) +
  scale_linetype_manual(values = c("solid", "dotted", "longdash")) +
  labs(x = "tongue root advancement (z-scores)") +
  facet_grid(language ~ c2place)
```

The bimodal distribution of /u/ seems to be related to the velar condition, especially in Italian.

# TRA trajectories

```{r}
tra_series %>%
  ggplot(aes(proportion, tra_z, colour = c2phonation, group = rec_date)) +
  geom_line(alpha = 0.2)
```

```{r}
tra_series %>%
  ggplot(aes(proportion, tra_z, colour = c2phonation)) +
  geom_smooth()
```

```{r}
tra_series %>%
  ggplot(aes(proportion, tra_z, colour = c2phonation, linetype = c2phonation)) +
  geom_smooth() +
  facet_grid(c2place ~ vowel)
```

```{r}
tra_series %>%
  filter(speaker == "it01") %>%
  ggplot(aes(proportion, tra_z, colour = c2phonation, linetype = c2phonation)) +
  geom_smooth() +
  facet_grid(c2place ~ vowel)
```

```{r}
tra_series %>%
  filter(speaker == "it01", c2place == "coronal", vowel != "u") %>%
  ggplot(aes(proportion, tra_z, colour = c2phonation, linetype = c2phonation)) +
  geom_smooth() +
  facet_grid(c2place ~ vowel)
```

```{r}
tra_series %>%
  filter(speaker == "it02") %>%
  ggplot(aes(proportion, tra_z, colour = c2phonation, linetype = c2phonation)) +
  geom_smooth() +
  facet_grid(c2place ~ vowel)
```

```{r}
tra_series %>%
  filter(speaker == "pl05") %>%
  ggplot(aes(proportion, tra_z, colour = c2phonation, linetype = c2phonation)) +
  geom_smooth() +
  facet_grid(c2place ~ vowel)
```
