---
title: 'Longer vowel duration correlates with tongue root advancement in Italian and Polish: An ultrasound study'
author: "Stefano Coretta"
date: "The University of Manchester"
institute: "LabPhon"
output:
  beamer_presentation:
    citation_package: natbib
    highlight: zenburn
    latex_engine: xelatex
fontsize: 12pt
bibliography: linguistics.bib
biblio-style: unified.bst
header-includes:
- \frenchspacing
- \usepackage{cleveref}
- \usetheme{metropolis}
- \titlegraphic{\includegraphics[height=1cm]{./graphics/logo.png}}
- \usecolortheme[snowy]{owl}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(tidyverse)
library(lme4)
```

```{r read-data}
speakers <- read_csv("./presentations/2018-labphon/datasets/speakers.csv")
stimuli <- read_csv("./presentations/2018-labphon/datasets/stimuli.csv")

duration <- list.files(
  path = "./presentations/2018-labphon/datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--")) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  filter(c1_phonation == "voiceless") %>%
  mutate(
    c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced"))
  ) %>%
  mutate_if(is.character, as.factor)
```

# Background

The **voicing effect**

* shorter vowels before voiceless stops, longer vowels before voiced stops

\tiny

@heffner1937; @house1953; @belasco1953; @peterson1960; @halle1967; @chen1970; @klatt1973; @lisker1974; @raphael1975; @javkin1976; @maddieson1976; @farnetani1986; @kluender1988; @lauefer1992; @fowler1992; @hussein1994; @esposito2002; @lampp2004; @warren2005; @durvasula2012

\normalsize

Still no consensus on source!

# Background

Proposed accounts:

* **production**
    * constant articulatory force [@belasco1953;@delattre1962]
    * durational trade-off [@slis1969;@lehiste1970]
    * laryngeal adjustment [@halle1967]
    * closing gesture duration [@chen1970]

* **perception**
    * misperception [@javkin1976]
    * enhancement [@kluender1988]

* but problems [@maddieson1976;@fowler1992]

# Background

**This talk**:

* support for *durational trade-off hypothesis*

* positive correlation between *vowel duration and tongue root position*

# Methods

* **Participants**: 11 Italians (5 F, 6 M), 6 Polish (3 F, 3 M)
* **Targets**
    * \textbf{C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{1}}
    * C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/
    * *pata*, *pada*, *paka*, ..., *poto*, *podo*, ...
* **Frame sentence**
    - *Dico X lentamente*, 'I say X slowly'
    - *Mówię X teraz*, 'I say X now'
* **Data**
    * Durational data from acoustics
    * Tongue root from ultrasound tongue imaging
* **Reproducibility**
    - <https://github.com/stefanocoretta/2018-labphon>

# Results

