---
title: 'Longer vowel duration correlates with tongue root advancement in Italian and Polish: An ultrasound study'
author: "Stefano Coretta"
date: "The University of Manchester"
institute: "LabPhon"
output:
  beamer_presentation:
    citation_package: natbib
    highlight: zenburn
    latex_engine: xelatex
fontsize: 12pt
bibliography: linguistics.bib
biblio-style: unified.bst
header-includes:
- \frenchspacing
- \usepackage{cleveref}
- \usetheme{metropolis}
- \titlegraphic{\includegraphics[height=1cm]{./graphics/logo.png}}
- \usecolortheme[snowy]{owl}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(tidyverse)
theme_set(theme_light())
library(rticulate)
library(lme4)
library(effects)
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
```

```{r read-data, include=FALSE}
speakers <- read_csv("./presentations/2018-labphon/datasets/speakers.csv")
stimuli <- read_csv("./presentations/2018-labphon/datasets/stimuli.csv")

duration_raw <- list.files(
  path = "./presentations/2018-labphon/datasets/acoustics/",
  pattern = "*-durations.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., na = "--undefined--"))

columns <- c(
  "speaker",
  "seconds",
  "rec_date",
  "prompt",
  "label",
  "TT_displacement_sm",
  "TT_velocity",
  "TT_velocity_abs",
  "TD_displacement_sm",
  "TD_velocity",
  "TD_velocity_abs",
  "TR_displacement_sm",
  "TR_velocity",
  "TR_velocity_abs"
)

duration <- list.files(
  path = "./presentations/2018-labphon/datasets/ultrasound/",
  pattern = "*-tongue-cart.tsv",
  full.names = TRUE
) %>%
  map_df(~read_aaa(., column_names = columns, format = "wide")) %>%
  select(-(X_1:Y_42)) %>%
  filter(label == "closure_") %>%
  mutate(word = word(prompt, 2)) %>%
  left_join(y = speakers) %>%
  left_join(y = stimuli) %>%
  left_join(y = duration_raw) %>%
  filter(c1_phonation == "voiceless") %>%
  droplevels() %>%
  mutate(
    c2_phonation = factor(c2_phonation, levels = c("voiceless", "voiced")),
    language = recode(language, italian = "Italian", polish = "Polish"),
    syl_rate = ifelse(
      language == "Italian",
      8 / sentence_duration,
      6 / sentence_duration
    )
  ) %>%
  mutate_if(is.character, as.factor) %>%
  # filter clear outlier
  filter(TR_displacement_sm > 40) %>%
  group_by(speaker) %>%
  mutate(
    vowel_duration_z = scale(vowel_duration),
    tongue_root = -TR_displacement_sm,
    tongue_root_z = scale(tongue_root)
  ) %>%
  ungroup() %>%
  filter(tongue_root_z < 3)
```

# Background

The **voicing effect**

* shorter vowels before voiceless stops, longer vowels before voiced stops

\tiny

@heffner1937; @house1953; @belasco1953; @peterson1960; @halle1967; @chen1970; @klatt1973; @lisker1974; @raphael1975; @javkin1976; @maddieson1976; @farnetani1986; @kluender1988; @laeufer1992; @fowler1992; @hussein1994; @esposito2002; @lampp2004; @warren2005; @durvasula2012

\normalsize

Still no consensus on source!

# Background

Proposed accounts:

* **production**
    * constant articulatory force [@belasco1953;@delattre1962]
    * durational trade-off [@slis1969;@lehiste1970]
    * laryngeal adjustment [@halle1967]
    * closing gesture duration [@chen1970]

* **perception**
    * misperception [@javkin1976]
    * enhancement [@kluender1988]

* but problems [@maddieson1976;@fowler1992]

# Background

**This talk**:

* support for *durational trade-off hypothesis*

* link between *vowel duration* and *tongue root position*

# Methods

* **Participants**: 11 Italians (5 F, 6 M), 6 Polish (3 F, 3 M)
* **Targets**
    * \textbf{C\textsubscript{1}V\textsubscript{1}C\textsubscript{2}V\textsubscript{1}}
    * C\textsubscript{1} = /p/, V\textsubscript{1} = /a, o, u/, C\textsubscript{2} = /t, d, k, g/
    * *pata*, *pada*, *paka*, ..., *poto*, *podo*, ...
* **Frame sentence**
    - *Dico X lentamente*, 'I say X slowly'
    - *Mówię X teraz*, 'I say X now'
* **Measurements**
    * Durational data from acoustics
    * Tongue root position from ultrasound tongue imaging
* **Reproducibility**
    - <https://github.com/stefanocoretta/2018-labphon>

# Methods

[image of acoustic annotation]

# Methods

[image of tongue root position]

# Results: Vowel duration

\centering

```{r vowel-duration-plot, echo=FALSE}
set.seed(7877)
duration %>%
  ggplot(aes(vowel, vowel_duration, fill = c2_phonation)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1) +
  facet_grid(~ language) +
  xlab("Vowel") +
  ylab("Vowel duration (ms)") +
  scale_fill_manual(name = "C2 voicing", values = cbPalette)
```

```{r vow-dur-lm, echo=FALSE, include=FALSE}
vow_dur_lm <- lmer(
  vowel_duration ~
    c2_phonation +
    c2_place +
    vowel +
    c2_phonation:vowel +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  REML = FALSE,
  data = duration
)

vow_dur_lm_null <- lmer(
  vowel_duration ~
#    c2_phonation +
    c2_place +
    vowel +
#    c2_phonation:vowel +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  REML = FALSE,
  data = duration
)

anova(vow_dur_lm_null, vow_dur_lm)
summary(vow_dur_lm)
```

```{r clos-dur-lm, echo=FALSE, include=FALSE}
clos_dur_lm <- lmer(
  closure_duration ~
    c2_phonation +
    c2_place +
    vowel +
    c2_phonation:vowel +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  REML = FALSE,
  data = duration
)

clos_dur_lm_null <- lmer(
  closure_duration ~
#    c2_phonation +
    c2_place +
    vowel +
#    c2_phonation:vowel +
    syl_rate +
    (1+c2_phonation|speaker) +
    (1|item),
  REML = FALSE,
  data = duration
)

anova(clos_dur_lm_null, clos_dur_lm)
summary(clos_dur_lm)
```

# Results: Vowel and closure duration

```{r vow-clos-lm, echo=FALSE, include=FALSE}
vow_clos_lm <- lmer(
  vowel_duration ~
    closure_duration +
    syl_rate +
    vowel +
    c2_place +
    (1|speaker) +
    (1|item),
  data = duration,
  REML = FALSE
)
summary(vow_clos_lm)
```

\centering

```{r vow-clos-plot, echo=FALSE, warning=FALSE}
as_tibble(effect("closure_duration", vow_clos_lm)) %>%
  ggplot(aes(closure_duration, fit)) +
  geom_point(
    data = filter(duration, vowel_duration < 151),
    aes(closure_duration, vowel_duration, colour = language),
    alpha = 0.2
  ) +
  geom_ribbon(aes(ymax = upper, ymin = lower), fill = cbPalette[3], alpha = 0.2) +
  geom_line(colour = cbPalette[3]) +
  coord_fixed(ratio = 1) +
  labs(
    x = "Closure duration (ms)",
    y = "Vowel duration (ms)"
  ) +
  scale_color_manual(name = "Language", values = cbPalette[1:2])
```

# Results: Vowel and closure duration

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/trade-off.pdf}
\end{figure}

# Results: Release to Release duration

```{r rr-box-lang, echo=FALSE, warning=FALSE}
set.seed(7877)
duration %>%
  group_by(speaker) %>%
  mutate(rel_rel = scale(rel_rel)) %>%
  ggplot(aes(c2_phonation, rel_rel, fill = c2_phonation)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1) +
  facet_grid(~ language) +
  scale_fill_manual(name = "C2 voicing", values = cbPalette[1:2]) +
  labs(
    x = "C2 voicing",
    y = "Release to Release duratio (z-scores)"
  ) +
  theme(legend.position = "none")
```

```{r rr-lm, echo=FALSE, include=FALSE}
rr_lm <- lmer(
  rel_rel ~
    c2_phonation +
    vowel +
    c2_place +
    syl_rate +
    language +
    (1+c2_phonation|speaker) +
    (1|item),
  data = duration,
  REML = FALSE
)

rr_lm_null <- lmer(
  rel_rel ~
    vowel +
    c2_place +
    syl_rate +
    language +
    (1+c2_phonation|speaker) +
    (1|item),
  data = duration,
  REML = FALSE
)

anova(rr_lm_null, rr_lm)
summary(rr_lm)
```

# Discussion: Durational trade-off

\begin{figure}
  \includegraphics[width=0.8\textwidth]{./graphics/trade-off-2.pdf}
\end{figure}

# Discussion: Vowel duration and tongue root position

```{r tr-lm, echo=FALSE, include=FALSE}
vow_tr_lm <- lmer(
  vowel_duration ~
    tongue_root +
    c2_phonation +
    vowel +
    syl_rate +
    (1+tongue_root|speaker) +
    (1|item),
  data = duration,
  REML = FALSE
)

vow_tr_lm_null <- lmer(
  vowel_duration ~
    c2_phonation +
    vowel +
    syl_rate +
    (1+tongue_root|speaker) +
    (1|item),
  data = duration,
  REML = FALSE
)

anova(vow_tr_lm_null, vow_tr_lm)
summary(vow_tr_lm)
```

```{r vow-tr-plot, echo=FALSE, warning=FALSE}
as_tibble(effect("tongue_root", vow_tr_lm)) %>%
  ggplot(aes(tongue_root, fit)) +
  geom_rug(
    inherit.aes = FALSE,
    data = duration,
    mapping = aes(x = tongue_root),
    alpha = 0.2
  ) +
  geom_ribbon(aes(ymax = upper, ymin = lower), fill = cbPalette[3], alpha = 0.2) +
  geom_line(colour = cbPalette[3]) +
  labs(
    x = "Tongue root position (mm)",
    y = "Vowel duration (ms)"
  ) +
  scale_color_manual(name = "Language", values = cbPalette[1:2])
```
