---
title: "Spline plotting and SSANOVA"
author: "Stefano Coretta"
date: "27/12/2016"
output: 
  pdf_document: 
    latex_engine: xelatex
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=normalizePath("../../"))
library(readr)
library(purrr)
library(tidyr)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(gss)
library(stringr)
```

# Data import

Before importing the data, we need to specify the column names for the data set header, since the raw data does not have a header. The number of splines is always 42 (saved as `num.splines`). `first.columns` is a factor containing the names of columns that are not the splines coordinates columns. Finally, we can concatenate `first.columns` with the names of the splines coordinates which is generated by the `paste0` function in the format `X_1, Y_1, X_2, Y_2, X_n, ...` (the underscore `_` will be useful for separating the axis from the fan number).

```{r col-names}
num.splines <- 42
first.columns <- c(
    "speaker",
    "seconds",
    "rec.date",
    "prompt",
    "label",
    "TT.displacement.sm",
    "TT.velocity",
    "TT.velocity.abs",
    "TD.displacement.sm",
    "TD.velocity",
    "TD.velocity.abs"
    )
columns <- c(first.columns,
             paste0(rep(c("X", "Y"), num.splines),
                    "_",
                    rep(1:num.splines, each = 2)
                    )
             )
```

We can now read in the file.

```{r read}
splines.raw <- list.files(path = "./pilot/results",
                   pattern = "*-aaa.txt",
                   full.names = TRUE) %>%
    map_df(~read_tsv(., col_names = columns, na = "*"))

rm(num.splines, first.columns, columns)

languages <- read_csv("./pilot/stimuli/languages.csv")
words <- read_csv("./pilot/stimuli/nonce.csv")
```

The following code applies tidy formatting to the data frame. It uses functions from the `tidyr` library.

```{r tidy}
splines <- splines.raw %>%
    gather(spline, coordinate, matches("[XY]_")) %>%
    separate(spline, c("axis", "fan"), convert = TRUE) %>%
    spread(axis, coordinate) %>%
    mutate(word = word(prompt, 2)) %>%
    left_join(y = languages) %>%
    left_join(y = words) %>%
    mutate_if(is.character, as.factor)

splines.it <- filter(splines, language == "italian")
splines.pl <- filter(splines, language == "polish")
```

# Some plotting

We can finally plot splines.

```{r velar, eval=FALSE}
filter(splines, label == "max_TD") %>%
ggplot(aes(x = X, y = Y, colour = c2phonation)) +
    geom_smooth(method = "loess") +
    coord_fixed(ratio = 1)
```

```{r coronal, eval=FALSE}
filter(splines, label == "max_TT") %>%
ggplot(aes(x = X, y = Y, colour = c2phonation)) +
    geom_smooth(method = "loess") +
    coord_fixed(ratio = 1)
```

# SSANOVA

```{r ssanovaTongue}
ssanovaTongue <- function(splines) {
    model <- ssanova(Y ~ c2phonation + X + c2phonation:X, data = splines)
    
    predicted <- expand.grid(X = seq(min(splines$X, na.rm = TRUE),
                                     max(splines$X, na.rm = TRUE),
                                     length = 100),
                             Y = seq(min(splines$Y, na.rm = TRUE),
                                     max(splines$Y, na.rm = TRUE),
                                     length = 100),
                             c2phonation = c("voiced", "voiceless")
    )
    
    predicted$splines.fit <- predict(model, predicted, se = T)$fit
    predicted$splines.SE <- predict(model, predicted, se = T)$se.fit
    
    return(predicted)
}
```

```{r ssanova-it}
splines.cor.a <- filter(splines.it, vowel == "a", label == "max_TT")
ssanova.cor.a <- ssanovaTongue(splines.cor.a)

splines.cor.o <- filter(splines.it, vowel == "o", label == "max_TT")
ssanova.cor.o <- ssanovaTongue(splines.cor.a)

splines.cor.u <- filter(splines.it, vowel == "u", label == "max_TT")
ssanova.cor.u <- ssanovaTongue(splines.cor.u)


splines.vel.a <- filter(splines.it, vowel == "a", label == "max_TD")
ssanova.vel.a <- ssanovaTongue(splines.vel.a)

splines.vel.o <- filter(splines.it, vowel == "o", label == "max_TD")
ssanova.vel.o <- ssanovaTongue(splines.vel.a)

splines.vel.u <- filter(splines.it, vowel == "u", label == "max_TD")
ssanova.vel.u <- ssanovaTongue(splines.vel.u)
```


```{r ssanova-cor-plot}
ggplot(ssanova.cor.a, aes(x = X, colour = c2phonation)) +
geom_line(aes(y = splines.fit)) +
geom_ribbon(aes(ymin = splines.fit - (1.96 * splines.SE),
                ymax = splines.fit + (1.96 * splines.SE),
                fill = c2phonation
                ),
            alpha = 0.5,
            color = "NA"
            )

ggplot(ssanova.cor.o, aes(x = X, colour = c2phonation)) +
geom_line(aes(y = splines.fit)) +
geom_ribbon(aes(ymin = splines.fit - (1.96 * splines.SE),
                ymax = splines.fit + (1.96 * splines.SE),
                fill = c2phonation
                ),
            alpha = 0.5,
            color = "NA"
            )

ggplot(ssanova.cor.u, aes(x = X, colour = c2phonation)) +
geom_line(aes(y = splines.fit)) +
geom_ribbon(aes(ymin = splines.fit - (1.96 * splines.SE),
                ymax = splines.fit + (1.96 * splines.SE),
                fill = c2phonation
                ),
            alpha = 0.5,
            color = "NA"
            )
```

```{r ssanova-vel-plot}
ggplot(ssanova.vel.a, aes(x = X, colour = c2phonation)) +
geom_line(aes(y = splines.fit)) +
geom_ribbon(aes(ymin = splines.fit - (1.96 * splines.SE),
                ymax = splines.fit + (1.96 * splines.SE),
                fill = c2phonation
                ),
            alpha = 0.5,
            color = "NA"
            )

ggplot(ssanova.vel.o, aes(x = X, colour = c2phonation)) +
geom_line(aes(y = splines.fit)) +
geom_ribbon(aes(ymin = splines.fit - (1.96 * splines.SE),
                ymax = splines.fit + (1.96 * splines.SE),
                fill = c2phonation
                ),
            alpha = 0.5,
            color = "NA"
            )

ggplot(ssanova.vel.u, aes(x = X, colour = c2phonation)) +
geom_line(aes(y = splines.fit)) +
geom_ribbon(aes(ymin = splines.fit - (1.96 * splines.SE),
                ymax = splines.fit + (1.96 * splines.SE),
                fill = c2phonation
                ),
            alpha = 0.5,
            color = "NA"
            )
```

# Consonantal gestures: target, maximum, release

Now we can create a separate data frame where the observasional unit is each word and the variables are the consonantal getures (target, maximum closure, release).

```{r gestures}
cons.gestures <- select(splines.raw, speaker:label) %>%
    mutate(word = word(prompt, 2)) %>%
    left_join(y = languages) %>%
    left_join(y = words) %>%
    separate(label, c("gesture", "tongue.area")) %>%
    spread(gesture, seconds) %>%
    mutate(nucleus = NOFF - NONS, deceleration = max - NONS, 
           acceleration = NOFF - max, skewness = deceleration / nucleus) %>%
    mutate_if(is.character, as.factor)
```

Let's plot the nucleus duration (NOFF - NONS) as a function of place of articulation, voicing of C2 (our target consonant in C1VC2V words), and language.

```{r nucleus-duration, eval = FALSE}
ggplot(cons.gestures, aes(c2phonation, nucleus)) +
    facet_grid(.~language + c2place) +
    geom_violin() +
    geom_boxplot(width=0.1) +
    theme(strip.background = element_rect(fill = "black"),
          strip.text = element_text(color = "white", face = "bold")
          )
```


```{r deceleration-duration, eval = FALSE}
ggplot(cons.gestures, aes(c2phonation, deceleration)) +
    facet_grid(.~language + c2place) +
    geom_violin() +
    geom_boxplot(width=0.1) +
    theme(strip.background = element_rect(fill = "black"),
          strip.text = element_text(color = "white", face = "bold")
          )
```

```{r acceleration-duration, eval = FALSE}
ggplot(cons.gestures, aes(c2phonation, acceleration)) +
    facet_grid(.~language + c2place) +
    geom_violin() +
    geom_boxplot(width=0.1) +
    theme(strip.background = element_rect(fill = "black"),
          strip.text = element_text(color = "white", face = "bold")
          )
```

The follwing plots the skweness (`deceleration / nucleus`) of the nucleus gesture.

```{r skewness-duration, eval = FALSE}
ggplot(cons.gestures, aes(c2phonation, skewness)) +
    facet_grid(.~language + c2place) +
    geom_violin() +
    geom_boxplot(width=0.1) +
    theme(strip.background = element_rect(fill = "black"),
          strip.text = element_text(color = "white", face = "bold")
          )
```

I want to check how good the gesture function in AAA is performing. The following gives the number of missing values per speaker.

```{r performance-na}
cons.gestures %>%
    select(speaker, GOFF:NONS) %>%
    group_by(speaker) %>%
    summarise_each(funs(sum(is.na(.)))) %>%
    select(speaker, GONS, NONS, max, NOFF, GOFF)
```

This chunk instead reports the total number of values per speaker.

```{r performance-total}
cons.gestures %>%
    select(speaker, GOFF:NONS) %>%
    group_by(speaker) %>%
    summarise_each(funs(n())) %>%
    select(speaker, GONS, NONS, max, NOFF, GOFF)
```







