---
title: "Voicing Effect project procedure"
author: "Stefano Coretta"
date: "19/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(tidyverse)
library(speakr)
library(beepr)
```

## Tangle code

```{r tangle}
lmt("./voicing-effect/code/process.praat.md")
```


## Ultrasound data

1. **Export** `.wav` and `.txt` from `AAA` to `/data/derived/ultrasound/[ID]/audio` folder ([ID] = participant ID) as [ID]-00n.wav
1. **Force alignment and search area**
   1. Run `alignment-input.praat`: the output is a concatenated `.wav` file and a `.txt` file listing the stimuli, written in `/data/derived/ultrasound/[ID]/alignment`
   1. Make sure the `.txt` contains "Mowie" (unicode safe because of AAA)
   1. **WARNING OVERWRITE** Run SPPAS on concatenated `.wav` and `.txt`
   1. Check alignment
   1. Run `search-area.praat`: the output is TextGrid files for each `AAA` `.wav` file, saved in `/data/derived/ultrasound/[ID]/audio`, and a merged TextGrid, saved in `/data/derived/ultrasound/[ID]/alignment`
        1. **NOTE**: Needed for EGG processing!
   1. Import TextGrids back in `AAA`
1. **Tongue tracking**
   1. Create a spline template for the speaker and save it (`.fst`)
   1. Batch process splines with template
   1. Check splines
   1. (Batch process snap-to-fit)
1. **Calculate kinematics**
   1. Check relevant splines for Tongue Dorsum and Tongue Tip
   1. Load `AnaVal-[ID].avl` from `Edit values` and change splines to TD and TT splines (Save it for the future)
   1. Calculate Maths for entire session
1. **Find consonantal gestures**
   1. Select TD stimuli
   1. `Find...` > `Load` and `execute` `td_function.srh`
   1. Select TT stimuli
   1. `Find...` > `Load` and `execute` `tt_function.srh`
1. **Export data** in `results` folder with `Export.xsu` to the file `[ID]-aaa.txt`
1. **Export kinematics annotations** (from `Export > Files...`) in `data/derived/ultrasound/[ID]/kinematics` folder as `[ID]-00n.TextGrid`
1. **Burst detection**
   1. Run `burst-detection.praat`: the output is a `.TextGrid` file, written in `/data/derived/ultrasound/[ID]/alignment`
   1. Check bursts
1. **Get durations**
   1. Run `get-durations.praat`: the output is written in the `results` folder

## EGG data

1. **Synchronise** EGG with `sync-egg.praat`: the output is written in `/data/derived/egg/[ID]`
1. **Extract vuv** (Voiced/UnVoiced) with `extract-vuv.praat`: the output is written in the same folder as above
1. **Calculate dEGG tracegrams** with `degg-tracing.praat`: the results are written in `results`

```{r sync-vuv-degg}
# speakers <- "it01"
speakers <- c("it01", "it02", "it03", "it07", "pl02", "pl03", "pl04", "pl05")

for (i in 1:length(speakers)) {
    praat_run(
        "./voicing-effect/code/sync-egg.praat",
        "voicing-effect",
        speakers[i],
        0
    )
}

for (i in 1:length(speakers)) {
    praat_run(
        "./voicing-effect/code/extract-vuv.praat",
        "voicing-effect",
        speakers[i],
        40,
        10000,
        11,
        0
    )
}

for (i in 1:length(speakers)) {
    praat_run(
        "./voicing-effect/code/degg-tracing.praat",
        "voicing-effect",
        speakers[i],
        40,
        10000,
        11
    )
}

beep()
```

1. **Calculate measurements** from ultrasound and EGG annotations with `get-measurements.praat`: the resulting TextGrids are written in `/data/derived/merged/[ID]` and the measurements in `results` to the file `[ID]-measurements.csv`

```{r get-measurements}
praat_run(
    "./voicing-effect/code/get-measurements.praat",
    speaker
)
```


## `AAA` files extensions
* `.esu`: tabs in the main window
* `.fst`: fans template
* `.avl`: analysis values chart
* `.srh`: search analysis value
* `.xsu`: export settings
